### 2013-07-23 paper/main.tex
Add beginning of paper

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\WarningsOff

\begin{document}
\newcommand{\idris}{\textsc{Idris}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{Draft for submission to IFL 2013}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

\title{Correct and Secure Web Programming using 
Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

\maketitle

\begin{abstract}

\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

%\keywords
%keyword1, keyword2

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits. The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development but at the same time do not give the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.

An alternative approach is to ensure that a program is semantically sound before it is run. Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled before a program successfully type-checks. A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer. This severely limits the uptake of dependently-typed languages.

This complexity can be ameliorated through the use of \textit{embedded domain-specific languages} (EDSLs), which aim to abstract away from the complexity of the underlying dependent type mechanisms, allowing users to write verfied domain-specific code without having to consider the underlying verification mechanisms.   %TODO: rephrase this

% Probably could do with rewriting a lot of this
Web applications are also intrinsically side-effecting, requiring extensive network communication, database queries and file I/O. Purely functional programming languages such as Haskell \cite{haskell} handle this monadically, but the composition of different effects can be difficult. Developers either make use of \textit{monad transformers} to compose different effects, or handle everything in a much coarser-grained monad.

\idris{} \cite{} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros. Through the use of \idris{}, and the \texttt{Effects} library \cite{}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.

\subsection{Contributions}
The primary contribution of this paper is the application of previous work to show that dependent types may be used to provide stronger static guarantees about the correctness and security of web applications, whilst minimising additional overhead. In particular:

\begin{itemize}
\item Representations of CGI, SQLite and a simple session handler as \textit{algebraic effects}, allowing programs to be accepted only when they adhere to specific resource usage protocols.
\item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
\item Sample applications, making use of the framework and demonstrating that it can be practically used.
\end{itemize}

All code used in this paper is available online at \\ \verb+http://www.github.com/SimonJF/IdrisWeb+.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} library}
\texttt{Effects} \cite{effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity. This section is intended to serve as a brief introduction to the library; for more information, see \cite{effects}.

An effect is represented as an instance of the \texttt{EFFECT} type. Such instances may be associated with a \textit{resource}, 
Describing a program in the \texttt{Effects} library is achieved using the following data type:
\begin{verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{verbatim}

\texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}). The list of input types can be used to ensure that an operation may only be called when the program is in a particular state, and since the list of input effects may differ from the list of output effects, we may modify the resource associated with the state after each operation. This allows us to simply enforce a resource usage protocol, and deviation from this protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.

%TODO: Execution
\begin{verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{verbatim}

Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

% =================================================
% =================================================

\section{Modelling resource usage protocols}
In this section, we show how three such effects; CGI, SQLite and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

% =================================================

\subsection{CGI}
%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {Initialised};
  \node[state]         (C) [below of=B] {TaskRunning};
  \node[state]         (D) [below of=C] {TaskCompleted};
  \node[state]         (E) [below of=D] {HeadersWritten};
  \node[state]         (F) [below of=E] {ContentWritten};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
\end{tikzpicture}
}
\caption{CGI States}
\label{fig:cgistates}
\end{figure}

CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client. Importantly, HTTP headers must be correctly written to the user prior to any other output; failure to do so will result in an internal server error.

A previous implementation of CGI in \idris{} implemented CGI as an extension of monadic IO. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are firstly written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI effect as the unit type, ().

\begin{verbatim}
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{verbatim}

Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses. Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the resource.

Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved. The only operation that can be performed in this state is initialisation: by calling \texttt{initalise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the \texttt{Cgi} algebraic data type, and involves transitioning from the uninitialised state to the initialised state.

{\small
\begin{verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
    ...

\end{verbatim}
}
User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in turn ensuring that the headers are first written.

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request. Finally, after the headers have been written, the content is then written, which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

In order to make use of the CGI framework, users specify a function of the following type:
\begin{verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{verbatim}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

% =================================================

\subsection{SQLite}

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExecuting};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=3em] {\texttt{closeDB}} (A)
        (B) edge[stateEdge, bend left=80]   node[edgeLabel, xshift=3em] {\texttt{connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{SQLite Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

SQLite is a lightweight SQL database engine \cite{sqlite} (TODO: More introduction-y stuff). We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems. Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{ffi}, which is beyond the scope of this paper.

The creation, preparation and execution of SQL queries has a very specific usage protocol, with several possible points of failure. Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user. Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.

Figure ~\ref{fig:sqlitestates} shows the resource usage protocol for the SQLite library. There are three main phases involved in the usage of the SQLite effect: connection to the database, preparation of a query, and execution of the query. This is reflected in the resource associated with the SQLite effect, which again is parameterised by the current state.

\begin{verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

data DBPointer = ValidConn Ptr
               | InvalidConn

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{verbatim}
If a failure happens at any point during the computation, the resource is updated to reflect the failure. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.

\begin{verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{verbatim}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, hiding implementation details from the developer and encouraging cleaner and less error-prone code.

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we present a simple example wherein a database is queried in order to enumerate the contents of a simple table. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

Firstly, we define a function of type:
\begin{verbatim}
Eff IO [SQLITE ()] (Either String (List (String, Int)))
\end{verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, Int)} pairs, representing names and ages in the database, or an error will be returned.

\begin{verbatim}
testSelect : Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT * FROM `test`;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{verbatim}

The program firstly attempts to open a connection to the ``people.db'' database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources.

A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
{\small
\begin{verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] $
\end{verbatim}
}

% =================================================

\subsection{A Simple Session Handler}
All but the most trivial web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects. By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {SessionUninitialised};
  \node[state]         (B) [below of=A] {SessionInitialised};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler. In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key. The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.

Much akin to the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect.

\begin{verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

data SessionRes : SessionStep -> Type where
  InvalidSession : SessionRes s  
  ValidSession   : SessionID -> 
                   SessionData -> 
                   SessionRes s
\end{verbatim}
The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.

% =================================================

\section{Type-aware form handling}
Programming web applications often involves processing user data, which may then be used in further effectuful computations. Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.

This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker. One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data. This may involve sending an integer detailing the current page, which could be used in a query such as:

\begin{verbatim}
SELECT `name`, `age` FROM `staff` LIMIT ' + page + ', 5
\end{verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information. Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.

\subsection{Form Construction}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an abstract data type, \texttt{FormTy}.
\begin{verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{verbatim}
In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between our abstract representation of permitted form types and concrete types:
\begin{verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{verbatim}
%
In order to specify a form, we once again make use of the \texttt{Effects} library in order to create a resource-aware EDSL. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form. If a handler function with with different types is supplied, then this will result in a compile-time type error.

\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation. This allows us to parameterise the resource associated with the form construction EDSL over the types accepted by each individual form element.

{\small
\begin{verbatim}
using (G : Vect FormTy n)
 data FormRes : Vect FormTy n -> Type where
   FR : Nat -> Vect FormTy n -> String -> FormRes G

 data Form : Effect where
   AddTextBox : (label : String) -> 
                (fty : FormTy) -> 
                (Maybe (interpFormTy fty)) -> 
                Form (FormRes G) (FormRes (fty :: G)) () 
   ...
   Submit : interpCheckedFnTy G effs t -> 
            String -> 
            (effs : List WebEffect) -> 
            (t : FormTy) -> 
            Form (FormRes G) (FormRes []) String
\end{verbatim}
}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but these are omitted here for brevity.

The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a vector of the types associated with each element. The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the vector of element types, and the currently generated HTML for the form. It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.

By adding elements to the form, the vector of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}. Additionally, HTML for the form element is generated, and stored in the resource, which is returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.

To specify a form instance, we define a function of type \texttt{UserForm}:
\begin{verbatim}
UserForm : Type
UserForm = EffM m [FORM (FormRes []) 
                        (FormRes [])] String
\end{verbatim}
All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation. As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.

The \texttt{interpCheckedFnTy} type of a valid handler function is calculated by the \texttt{interpCheckedFnTy} function, given the vector of types expected by form elements, the return type of the handler function, and an abstract representation of effects permitted to be used by a handler function. 
At this point, it is unfortunately not possible to specify arbitrary effects due to difficulties with serialising initial resource environments. This is, however, not an inherent limitation as the \texttt{Effects} EDSL allows for the introduction of effects within an effectful computation.
{\small
\begin{verbatim}
interpCheckedFnTy : Vect FormTy n -> 
                    List EFFECT -> 
                    Type -> 
                    Type
interpCheckedFnTy tys effs t = interpCheckedFnTy' (reverse tys)
 where interpCheckedFnTy' : Vect FormTy n -> Type
       interpCheckedFnTy' [] = FormHandler effs t
       interpCheckedFnTy' (x :: xs) = Maybe (interpFormTy x) ->
                                      interpCheckedFnTy' xs
\end{verbatim}
}
Since the DSL records types by prepending them to the type vector, we must reverse the list before calculating the type. Additionally, we pass all arguments encased in the \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.

Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.

Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this. Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.

Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page. The generated metadata describing the handler function is serialised as a hidden HTML field.

\subsection{Data Handling}
Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation. This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a three-tuple of the argument types, required effects, and return type.

\subsection{Example Code}

\section{Extended Example: Message Board}
\section{Related Work}
\section{Conclusions}
\subsection{Further Work}
% Cryptography bindings would be really good.
% Integration with a web server instead of doing everything over CGI
% Improvements to form handling system

% JS backend

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

\acks
% SICSA / 
%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-07-24 paper/main.tex
More work on paper, removal of unused / outdated types

@@
 \title{Correct and Secure Web Programming using 
-Dependent Types and Embedded Domain-Specific Languages}
+Embedded Domain-Specific Languages}
 %\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
@@
          (t : Type) -> Type
+\end{verbatim}
+
+Operations are made available to users by writing functions with the correct \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
+
+\begin{verbatim}
+queryPostVar : 
+ String -> 
+ Eff m [CGI (InitialisedCGI TaskRunning)] 
+   (Maybe String)
+queryPostVar query = (QueryPostVar query)
 \end{verbatim}
@@
 Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an abstract data type, \texttt{FormTy}.
+In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
 \begin{verbatim}
@@
 \end{verbatim}
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We use this to convert between our abstract representation of permitted form types and concrete types:
+In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between abstract and concrete representations of allowed form types:
 \begin{verbatim}
@@
 \idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation.
-This allows us to parameterise the resource associated with the form construction EDSL over the types accepted by each individual form element.
+We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.
 
@@
 
-The \texttt{interpCheckedFnTy} type of a valid handler function is calculated by the \texttt{interpCheckedFnTy} function, given the vector of types expected by form elements, the return type of the handler function, and an abstract representation of effects permitted to be used by a handler function.
+Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
+Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
+If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
+
+Before associating a handler function with the form, we must firstly specify the effects available to the handler.
+This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
+By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
+
+\begin{verbatim}
+useEffects : (effs : List WebEffect) ->
+             EffM m [FORM (FormRes G E)] 
+                    [FORM (FormRes G effs)] ()
+useEffects effs = (UseEffects effs)
+\end{verbatim}
 At this point, it is unfortunately not possible to specify arbitrary effects due to difficulties with serialising initial resource environments.
-This is, however, not an inherent limitation as the \texttt{Effects} EDSL allows for the introduction of effects within an effectful computation.
-{\small
+This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
+%
+We may specify a handler function of type \texttt{FormHandler}:
 \begin{verbatim}
-interpCheckedFnTy : Vect FormTy n -> 
-                    List EFFECT -> 
-                    Type -> 
-                    Type
-interpCheckedFnTy tys effs t = interpCheckedFnTy' (reverse tys)
- where interpCheckedFnTy' : Vect FormTy n -> Type
-       interpCheckedFnTy' [] = FormHandler effs t
-       interpCheckedFnTy' (x :: xs) = Maybe (interpFormTy x) ->
-                                      interpCheckedFnTy' xs
+FormHandler : List EFFECT -> Type
+FormHandler effs = Eff IO effs ()
 \end{verbatim}
-}
-Since the DSL records types by prepending them to the type vector, we must reverse the list before calculating the type.
-Additionally, we pass all arguments encased in the \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
+In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
+%
+\begin{verbatim}
+  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
+              (fns : HandlerList) ->
+              {default tactics { applyTactic 
+                findFn 100; solve; }
+               prf : FnElem f fns} ->
+              EffM m [FORM (FormRes G E)]
+                     [FORM (FormRes [] [])] 
+                     String
+  addSubmit f handlers {prf} = (Submit f name)
+    where name : String
+          name = getString' f handlers prf          
+\end{verbatim}
+%
+Let us look at each aspect of this function in turn.
+Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function.
+Note that since we prepend types to the list of \texttt{FormTy}s as opposed to prepending them, we must reverse the list.
+%
+\begin{verbatim}
+MkHandlerFnTy : Type
+MkHandlerFnTy = (List FormTy, List WebEffect)
+
+mkHandlerFn' : List FormTy -> List WebEffect -> Type
+mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
+mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
+                              mkHandlerFn' xs effs
+
+mkHandlerFn : MkHandlerFnTy -> Type 
+mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
+\end{verbatim}
+The \texttt{mkHandlerFn} function takes as its argument a tuple describing the arguments and web effects available to the function.
+When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
+We may then extract the information through pattern matching.
+%The type of a valid handler function is calculated by the \texttt{interpCheckedFnTy} function, given the vector of types expected by form elements, the return type of the handler function, and an abstract representation of effects permitted to be used by a handler function.
+%
+%{\small
+%\begin{verbatim}
+%interpCheckedFnTy : Vect FormTy n -> 
+%                    List EFFECT -> 
+%                    Type -> 
+%                    Type
+%interpCheckedFnTy tys effs t = interpCheckedFnTy' (reverse tys)
+% where interpCheckedFnTy' : Vect FormTy n -> Type
+%       interpCheckedFnTy' [] = FormHandler effs t
+%       interpCheckedFnTy' (x :: xs) = Maybe (interpFormTy x) ->
+%                                      interpCheckedFnTy' xs
+%\end{verbatim}
+%}
 

### 2013-07-25 paper/main.tex
Form construction / handling first draft completed

@@
 \preprintfooter{}   % 'preprint' option specified.
-
-\title{Correct and Secure Web Programming using 
+\title{Correct and Secure Web Programming using Dependent Types and
 Embedded Domain-Specific Languages}
@@
 Describing a program in the \texttt{Effects} library is achieved using the following data type:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 \texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}).
@@
 Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
-
+{\small
 %TODO: Execution
@@
 \end{verbatim}
-
+}
 Operations are made available to users by writing functions with the correct \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context.
@@
 We represent an uninitialised CGI effect as the unit type, ().
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
@@
 In order to make use of the CGI framework, users specify a function of the following type:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
@@
 This is reflected in the resource associated with the SQLite effect, which again is parameterised by the current state.
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 If a failure happens at any point during the computation, the resource is updated to reflect the failure.
@@
 We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
@@
 Firstly, we define a function of type:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 This means that the program will be run in the IO execution context, and must start and end with no active resources.
 The return type indicates that either a list of \text{(String, Int)} pairs, representing names and ages in the database, or an error will be returned.
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 The program firstly attempts to open a connection to the ``people.db'' database.
@@
 Much akin to the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
-
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
@@
 
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
@@
 In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between abstract and concrete representations of allowed form types:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 %
@@
 To specify a form instance, we define a function of type \texttt{UserForm}:
+{\small
 \begin{verbatim}
@@
                         (FormRes [])] String
-\end{verbatim}
+\end{verbatim}}
 All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
@@
 By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-
+{\small
 \begin{verbatim}
@@
 useEffects effs = (UseEffects effs)
-\end{verbatim}
-At this point, it is unfortunately not possible to specify arbitrary effects due to difficulties with serialising initial resource environments.
+\end{verbatim}}
+Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
 This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
@@
 Note that since we prepend types to the list of \texttt{FormTy}s as opposed to prepending them, we must reverse the list.
-%
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 The \texttt{mkHandlerFn} function takes as its argument a tuple describing the arguments and web effects available to the function.
@@
 We may then extract the information through pattern matching.
-%The type of a valid handler function is calculated by the \texttt{interpCheckedFnTy} function, given the vector of types expected by form elements, the return type of the handler function, and an abstract representation of effects permitted to be used by a handler function.
+
+To store a reference to a handler function, we use the \texttt{HandlerFn} type:
+\begin{verbatim}
+public
+HandlerFn : Type
+HandlerFn = (ft ** (mkHandlerFn ft, String))
+\end{verbatim}
 %
-%{\small
-%\begin{verbatim}
-%interpCheckedFnTy : Vect FormTy n -> 
-%                    List EFFECT -> 
-%                    Type -> 
-%                    Type
-%interpCheckedFnTy tys effs t = interpCheckedFnTy' (reverse tys)
-% where interpCheckedFnTy' : Vect FormTy n -> Type
-%       interpCheckedFnTy' [] = FormHandler effs t
-%       interpCheckedFnTy' (x :: xs) = Maybe (interpFormTy x) ->
-%                                      interpCheckedFnTy' xs
-%\end{verbatim}
-%}
+The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
+We also store a unique string identifier, which is used to serialise a reference to the handler function.
+We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
+
+\begin{verbatim}
+public
+HandlerList : Type
+HandlerList = List HandlerFn
+\end{verbatim}
+To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
+{\small
+\begin{verbatim}
+  data FnElem : mkHandlerFn ((reverse G), E) -> 
+                HandlerList -> Type where
+
+       FnHere : {xs : HandlerList, f : 
+                 mkHandlerFn ((reverse G), E)} ->
+         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
+       FnThere : {xs : HandlerList, f : 
+                 mkHandlerFn ((reverse G), E)} ->
+               FnElem f xs -> FnElem f (x :: xs)
+\end{verbatim}
+}
+\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
+\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
+%TODO: this should probably be rewritten
+We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
+The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
+If the automated proof search fails, compilation will fail with a compile error.
 
@@
 
-\subsection{Data Handling}
+\subsection{Form Handling}
 Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a three-tuple of the argument types, required effects, and return type.
+This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
 
-\subsection{Example Code}
+With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
+In order to apply the handler function to the data submitted in the form, we must firstly prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
+We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
 
+With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{viewfromleft}, to rewrite the arguments on the left hand side.
+This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
+If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
+If not, the computation is abandoned and an error is shown.
+% I gave it a shot, right?
+:P
+{\small
+\begin{verbatim}
+checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
+                 (frm_fn_ty : MkHandlerFnTy) -> 
+                 mkHandlerFn reg_fn_ty -> 
+                 Maybe (mkHandlerFn frm_fn_ty)
+checkFunctions reg_ty frm_ty reg_fn with 
+                             (decEq reg_ty frm_ty)
+  checkFunctions frm_ty frm_ty reg_fn | Yes refl = Just reg_fn
+  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
+\end{verbatim}}
+We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
+Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
 \section{Extended Example: Message Board}
@@
 
-% JS backend
-
 %\appendix
@@
 \acks
-% SICSA / 
-%Acknowledgments, if needed.
+% SICSA / EPSRC (grant number?)
+% #idris
+% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi
 

### 2013-07-26 paper/main.tex
More work on paper

@@
 \preprintfooter{}   % 'preprint' option specified.
-\title{Correct and Secure Web Programming using Dependent Types and
-Embedded Domain-Specific Languages}
+
+\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
 %\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
@@
 \begin{abstract}
+Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing for easier reasoning about the operation of such programs.
+Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker, and additional proofs may have to be specified by a user.
 
+Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over the underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without requiring complex proofs to be specified.
+
+In this paper, we apply this technique to web programming, showing that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
+Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks which exploit the loss of type information, such as SQL injection.
 \end{abstract}
@@
 
+In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
+In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
+{\small
+\begin{verbatim}
+data DBVal = DBInt Int
+           | DBText String
+           | DBFloat Float
+           | DBNull
+\end{verbatim}
+}
+We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
+\begin{verbatim}
+ResultSet : Type
+ResultSet = List (List DBVal)
+\end{verbatim}
+Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
+{\small
+\begin{verbatim}
+executeSelect : String ->
+                String -> 
+                List (Int, DBVal) -> 
+                (Eff m [SQLITE 
+                  (SQLiteRes PreparedStatementExecuting)
+                   ] 
+                 (List DBVal)
+                ) -> 
+                Eff m [SQLITE ()] (Either String ResultSet)
+\end{verbatim}
+}
 % =================================================
@@
 \section{Conclusions}
+Dependently-typed languages have, for a long time, remained relatively unused outside of academia.
+Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough tyeps) to be adopted without the additional proofs traditionally required by users of such languages.
+
+We believe we have demonstrated that embedded domain-specific languages can be used to make use of the power of dependently-typed languages without incurring the traditional overheads of 
 \subsection{Further Work}
+We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
+
+There are firstly many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}.
+Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence to their security.
+
+Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request request.
+Ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
+
+Dependently-typed languages provide great promise for the construction of secure and correct programs. Through the use of embedded domain-specific languages, we hope that
 % Cryptography bindings would be really good.
@@
 \acks
-% SICSA / EPSRC (grant number?)
+% SICSA / EPSRC (grant number?
 % #idris

### 2013-07-27 paper/main.tex
Changed executeInsert to be a self-contained operation.
Wrote up MessageBoard example

@@
 
-\idris{} \cite{} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros.
-Through the use of \idris{}, and the \texttt{Effects} library \cite{}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.
+\idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros.
+Through the use of \idris{}, and the \texttt{Effects} library \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.
 
@@
 \item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
-\item Sample applications, making use of the framework and demonstrating that it can be practically used.
+\item A message board application making use of the framework, and showing that the framework may be used for practical applications.
 \end{itemize}
@@
 \texttt{Effects} \cite{effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity.
-This section is intended to serve as a brief introduction to the library; for more information, see \cite{effects}.
+This section is intended to serve as a brief introduction to the library; for more information, see \cite{brady:effects}.
 
@@
 We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{ffi}, which is beyond the scope of this paper.
+Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{brady2011idris}, which is beyond the scope of this paper.
 
@@
 
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{viewfromleft}, to rewrite the arguments on the left hand side.
+With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
 This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
@@
 Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
+
+%-----------------------------
+%-----------------------------
+
 \section{Extended Example: Message Board}
+Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.
+
+Firstly, we create a database schema in which to record information stored by the message board.
+We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
+
+Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
+Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
+\subsection{Handling requests}
+Once the program is invoked via a CGI request, the \texttt{main} function will be called.
+We then run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
+
+{\small
+\begin{verbatim}
+main : IO ()
+main = do (runCGI [initCGIState, 
+                  InvalidSession, ()] 
+           handleRequest)
+          pure ()
+\end{verbatim}
+}
+The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function.
+If so, then the form handling routine is called, which parses the POST data in order to call the correct form handling function.
+If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
+{\small
+\begin{verbatim}
+handleRequest : CGIProg [SESSION 
+  (SessionRes SessionUninitialised), 
+  SQLITE ()] ()
+handleRequest = do 
+  handler_set <- isHandlerSet
+  if handler_set then do
+    lift' (handleForm handlers)
+    Effects.pure ()
+  else do
+    action <- lift' (queryGetVar "action")
+    thread_id <- lift' (queryGetVar "thread_id")
+    handleNonFormRequest action (map strToInt thread_id)
+\end{verbatim}
+}
+\subsection{Thread Creation}
+We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
+One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
+
+{\small
+\begin{verbatim}
+newThreadForm : UserForm
+newThreadForm = do
+  addTextBox "Title" FormString Nothing
+  addTextBox "Post Content" FormString Nothing 
+  useEffects [CgiEffect, SessionEffect, SqliteEffect]
+  addSubmit handleNewThread handlers
+\end{verbatim}
+}
+The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
+Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
+The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
+The \texttt{handlers} argument refers to the list of form handlers, and takes the following form:
+{\small
+\begin{verbatim}
+handlers : HandlerList
+handlers = [(([FormString, FormString], 
+              [CgiEffect, SessionEffect, SqliteEffect]) ** 
+               (handleRegisterForm, "handleRegisterForm")),
+
+            (([FormString, FormString], 
+              [CgiEffect, SessionEffect, SqliteEffect]) ** 
+               (handleNewThread, "handleNewThread")),
+             ...
+            ]
+\end{verbatim}
+}
+Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
+In order to do this, we make use of the session handler.
+We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
+If not, then a failure function is called instead.
+Should the form handler function be called with invalid arguments, an error is shown.
+{\small
+\begin{verbatim}
+handleNewThread : Maybe String -> 
+                  Maybe String -> 
+                  FormHandler [CGI (InitialisedCGI 
+                                 TaskRunning), 
+                               SESSION (SessionRes 
+                                 SessionUninitialised), 
+                               SQLITE ()]  
+handleNewThread (Just title) (Just content) = do 
+  withSession (addNewThread title content) notLoggedIn
+  pure ()
+handleNewThread _ _ = do 
+  outputWithPreamble "<h1>Error</h1><br />There was 
+       an error posting your thread."
+  pure ()
+\end{verbatim}
+}
+Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
+The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
+This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
+In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
+In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
+{\small
+\begin{verbatim}
+threadInsert : Int -> 
+               String -> 
+               String -> 
+               Eff IO [SQLITE ()] Bool
+threadInsert uid title content = do
+  let query = "INSERT INTO `Threads` 
+    (`UserID`, `Title`) VALUES (?, ?)"
+  insert_res <- (executeInsert DB_NAME query 
+    [(1, DBInt uid), (2, DBText title)]
+  case insert_res of
+    Left err => pure False
+    Right thread_id => postInsert uid thread_id content
+\end{verbatim}
+}
+\subsection{Listing Threads}
+Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
+{\small
+\begin{verbatim}
+getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
+getThreads = executeSelect DB_NAME query [] collectThreadResults
+  where query = "SELECT `ThreadID`, `Title`, `UserID`, 
+    `Username` FROM `Threads` NATURAL JOIN `Users`"
+\end{verbatim}
+}
+Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
+Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
+{\small
+\begin{verbatim}
+traverseThreads : ResultSet -> 
+  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
+traverseThreads [] = pure ()
+traverseThreads (x::xs) = do traverseRow x
+                             traverseThreads xs
+  where traverseRow : List DBVal -> 
+           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
+        traverseRow ((DBInt thread_id)::
+                     (DBText title)::
+                     (DBInt user_id)::
+                     (DBText username)::[]) =
+           (output \$ "<tr><td>
+            <a href=\"?action=showthread&thread_id=" ++ 
+            (show thread_id) ++ "\">" ++ 
+            title ++ "</a></td><td>" ++ 
+            username ++ "</td></tr>") 
+        traverseRow _ = pure ()
+\end{verbatim}
+}
+\subsection{Authentication}
+Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
+If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
+The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
+Any failures, for example with creating a new session or querying the database, are reported to the user.
+{\small
+\begin{verbatim}
+handleLoginForm (Just name) (Just pwd) = do
+  auth_res <- lift' (authUser name pwd)
+  case auth_res of
+    Right (Just uid) => do
+      set_sess_res <- setSession uid
+      if set_sess_res then do
+        lift' (output \$ "Welcome, " ++ name)
+        pure ()
+      else do
+        lift' (output "Could not set session")
+        pure ()
+    Right Nothing => do
+      lift' (output "Invalid username or password")
+      pure ()
+    Left err => do
+      lift' (output \$ "Error: " ++ err)
+      pure ()
+\end{verbatim}}
+Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
+
+Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
+Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
+Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
+%-----------------------------
+%-----------------------------
+
 \section{Related Work}
+
+%-----------------------------
+%-----------------------------
+
 \section{Conclusions}
@@
 % Improvements to form handling system
+% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
 
@@
 %This is the text of the appendix, if you need one.
+
+%-----------------------------
+%-----------------------------
 
@@
 % We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
+% #firstworldanarchists
+\bibliographystyle{plainnat}
 
@@
 
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+\bibliography{refs}
+%\begin{thebibliography}{}
+%\softraggedright
+%
+%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
+%P. Q. Smith, and X. Y. Jones. ...reference text...
+%
+%\end{thebibliography}
 

### 2013-07-28 paper/main.tex
More work on paper.
Rephrased things, updated code snippets to reflect latest code, added motivating example in the intro

@@
 \usepackage{tikz}
+\usepackage{url}
+\usepackage[hidelinks]{hyperref}
 \usetikzlibrary{arrows,automata}
@@
 \conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
+\copyrightyear{2013} 
 \copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
@@
 
-Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over the underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without requiring complex proofs to be specified.
+Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
 
@@
 
+%%------- EXAMPLE
+Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}.
+We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:
+
+{\small
+\begin{verbatim}
+$conn = new mysqli("localhost", "username", 
+                    "password", "db");
+\end{verbatim}
+}
+
+We must subsequently check to see if the connection was successful, and cease execution of the database routine if so.
+Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.
+
+{\small
+\begin{verbatim}
+  if (mysqli_connect_errno()) {
+      exit();
+  }
+\end{verbatim}
+}
+
+We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:
+
+{\small
+\begin{verbatim}
+  $stmt = $conn->prepare("SELECT `name`, `address` 
+     FROM `staff` WHERE `dept` = ?);
+  $stmt->bind_param($dept);
+\end{verbatim}
+}
+
+After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn.
+At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:
+
+{\small
+\begin{verbatim}
+  $stmt->execute();
+  $stmt->bind_result($name, $address);
+
+  while ($stmt->fetch()) {
+    printf("Name: %s, Age: %s", $name, $age);
+  }
+\end{verbatim}
+}
+
+Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:
+
+{\small
+\begin{verbatim}
+  $stmt->close();
+  $conn->close();
+\end{verbatim}
+}
+
+Even in this small example, there exists a very specific resource usage protocol.
+Firstly, a connection to the database must be opened.
+The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code.
+Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound.
+The statement is then executed, and each row is retrieved from the database.
+Finally, the resources are freed.
+
+There is, however, no way to statically guarantee that this protocol is adhered to.
+A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error.
+Similarly, a programmer may omit closing the statement or connection, and can lead to substantial problems such as handle leaks in longer-running server applications.
+
 An alternative approach is to ensure that a program is semantically sound before it is run.
-Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled before a program successfully type-checks.
+Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled prior to being accepted by a type checker.
 A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer.
-This severely limits the uptake of dependently-typed languages.
+This severely limits the uptake of dependently-typed languages in real-world settings.
 
@@
 
-% Probably could do with rewriting a lot of this
-Web applications are also intrinsically side-effecting, requiring extensive network communication, database queries and file I/O.
-Purely functional programming languages such as Haskell \cite{haskell} handle this monadically, but the composition of different effects can be difficult.
-Developers either make use of \textit{monad transformers} to compose different effects, or handle everything in a much coarser-grained monad.
-
 \idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros.
-Through the use of \idris{}, and the \texttt{Effects} library \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.
+Through the use of \idris{}, and the \texttt{Effects} framework \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.
 
@@
 
-All code used in this paper is available online at \\ \verb+http://www.github.com/SimonJF/IdrisWeb+.
+We structure the remainder of this paper as follows.
+We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocol for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
+
+The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
 % =================================================
@@
 
-\section{An overview of the \texttt{Effects} library}
-\texttt{Effects} \cite{effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity.
-This section is intended to serve as a brief introduction to the library; for more information, see \cite{brady:effects}.
+\section{An overview of the \texttt{Effects} framework}
+\label{effects}
+\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity.
+This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.
 
@@
 }
-Operations are made available to users by writing functions with the correct \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
+Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
 {\small
@@
 \section{Modelling resource usage protocols}
+\label{rup}
 In this section, we show how three such effects; CGI, SQLite and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
@@
 The only operation that can be performed in this state is initialisation: by calling \texttt{initalise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
-The \texttt{Init} operation is defined as part of the \texttt{Cgi} algebraic data type, and involves transitioning from the uninitialised state to the initialised state.
+The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
 
@@
 Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in turn ensuring that the headers are first written.
+It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in accordance with the resource usage protocol.
 
@@
 At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
-Finally, after the headers have been written, the content is then written, which completes the process.
+Finally, the headers and content are written which completes the process.
 Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
@@
   \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=3em] {\texttt{closeDB}} (A)
-        (B) edge[stateEdge, bend left=80]   node[edgeLabel, xshift=3em] {\texttt{connFail}} (A)
+        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
         (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{stmtFail}} (A)
+        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
         (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
@@
         (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{finalise}} (B)
+        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
         (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
@@
 
-SQLite is a lightweight SQL database engine \cite{sqlite} (TODO: More introduction-y stuff).
+SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
 We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
@@
 \subsection{A Simple Session Handler}
-All but the most trivial web applications require some persistent state across separate requests.
+More complex web applications require some persistent state across separate requests.
 This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
@@
 \section{Type-aware form handling}
+\label{form}
 Programming web applications often involves processing user data, which may then be used in further effectuful computations.
@@
 In order to specify a form, we once again make use of the \texttt{Effects} library in order to create a resource-aware EDSL.
-By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form.
-If a handler function with with different types is supplied, then this will result in a compile-time type error.
+By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
 
@@
 \begin{verbatim}
-using (G : Vect FormTy n)
- data FormRes : Vect FormTy n -> Type where
-   FR : Nat -> Vect FormTy n -> String -> FormRes G
+using (G : List FormTy, E : List WebEffect)
+  data FormRes : List FormTy -> 
+                 List WebEffect -> Type where
+    FR : Nat -> 
+         List FormTy -> 
+         List WebEffect -> 
+         String -> 
+         FormRes G E
 
@@
                  (Maybe (interpFormTy fty)) -> 
-                Form (FormRes G) (FormRes (fty :: G)) () 
+                 Form (FormRes G E) 
+                      (FormRes (fty :: G) E) () 
    ...
-   Submit : interpCheckedFnTy G effs t -> 
+
+    Submit : (mkHandlerFn ((reverse G), E)) ->
              String -> 
-            (effs : List WebEffect) -> 
-            (t : FormTy) -> 
-            Form (FormRes G) (FormRes []) String
+             Form (FormRes G E) (FormRes [] []) String
 \end{verbatim}
@@
 
-The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a vector of the types associated with each element.
-The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the vector of element types, and the currently generated HTML for the form.
+The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a list of the types associated with each element and the effects supported by the handler function.
+The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
 It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
 
-By adding elements to the form, the vector of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
+By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
 Additionally, HTML for the form element is generated, and stored in the resource, which is returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
@@
 We may specify a handler function of type \texttt{FormHandler}:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-%
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-%
+}
 Let us look at each aspect of this function in turn.
 Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to prepending them, we must reverse the list.
+Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
 {\small
@@
 The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail with a compile error.
+If the automated proof search fails, compilation will fail.
 
@@
 \subsection{Form Handling}
+\label{formhandling}
 Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
@@
 \section{Extended Example: Message Board}
+\label{messageboard}
 Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.
@@
 The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function.
-If so, then the form handling routine is called, which parses the POST data in order to call the correct form handling function.
+If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}.
 If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
@@
 The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and takes the following form:
+The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
 {\small
@@
 
-We believe we have demonstrated that embedded domain-specific languages can be used to make use of the power of dependently-typed languages without incurring the traditional overheads of 
 \subsection{Further Work}
@@
 % We recommend abbrvnat bibliography style.
-% #firstworldanarchists
+
 \bibliographystyle{plainnat}

### 2013-07-29 paper/main.tex
Touchups on paper

@@
 \begin{abstract}
-Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing for easier reasoning about the operation of such programs.
-Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker, and additional proofs may have to be specified by a user.
-
-Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, showing that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
-Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks which exploit the loss of type information, such as SQL injection.
+Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing easier reasoning about the operation of such programs.
+Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.
+
+Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
+
+In this paper, we apply this technique to web programming, and show that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
+Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection, which exploit the loss of type information.
 \end{abstract}
@@
 Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits.
-The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development but at the same time do not give the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.
+The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development.
+A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.
 
@@
 A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error.
-Similarly, a programmer may omit closing the statement or connection, and can lead to substantial problems such as handle leaks in longer-running server applications.
+Similarly, a programmer may omit closing the statement or connection, which can lead to substantial problems such as handle leaks in longer-running server applications.
 
@@
 \label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity.
+\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing greater code modularity.
 This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.
@@
 
-By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are firstly written correctly.
+By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
 In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
@@
                            (InitialisedCGI TaskRunning) a
-    ...
-
-\end{verbatim}
-}
+\end{verbatim}
+}
+Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
+
 User code executes in the \texttt{TaskRunning} state.
 Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in accordance with the resource usage protocol.
+It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
+This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
 
@@
 
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, hiding implementation details from the developer and encouraging cleaner and less error-prone code.
+By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
 
@@
 Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we present a simple example wherein a database is queried in order to enumerate the contents of a simple table.
+To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
 Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
 
-Firstly, we define a function of type:
-{\small
-\begin{verbatim}
-Eff IO [SQLITE ()] (Either String (List (String, Int)))
+We define a function of type:
+{\small
+\begin{verbatim}
+String ->
+Eff IO [SQLITE ()] (Either String (List (String, String)))
 \end{verbatim}
@@
 This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, Int)} pairs, representing names and ages in the database, or an error will be returned.
-{\small
-\begin{verbatim}
-testSelect : Eff IO [SQLITE ()] 
+The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
+{\small
+\begin{verbatim}
+testSelect : String -> Eff IO [SQLITE ()] 
              (Either String (List (String, Int)))
-testSelect = do
+testSelect dept = do
   open_db <- openDB "people.db"
   if open_db then do
-    let sql = "SELECT * FROM `test`;"
+    let sql = "SELECT name, address FROM `staff` 
+                    WHERE dept = ?;"
     sql_prep_res <- prepareStatement sql
@@
       startBind
+      bindText 1 dept
       finishBind
@@
 }
-The program firstly attempts to open a connection to the ``people.db'' database.
+The program initially attempts to open a connection to the \texttt{people.db} database.
 At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
 The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources.
+If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
 
@@
 We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
@@
 
-Much akin to the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
+Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
 {\small
@@
 \begin{verbatim}
-SELECT `name`, `age` FROM `staff` LIMIT ' + page + ', 5
+SELECT `name`, `address` FROM `staff` LIMIT ' + page + ', 5
 \end{verbatim}
@@
 }
-The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but these are omitted here for brevity.
+The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
 
@@
 
-Before associating a handler function with the form, we must firstly specify the effects available to the handler.
+Before associating a handler function with the form, we must specify the effects available to the handler.
 This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
@@
 To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{verbatim}
-public
+{\small
+\begin{verbatim}
 HandlerFn : Type
@@
 \begin{verbatim}
-public
 HandlerList : Type
@@
 Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
+This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
 
 With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must firstly prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
+In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
 We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
@@
 \label{messageboard}
-Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.
+Using these components, we may now consider a larger example of a message board which allowsusers to register, log in, view and create threads, and list and create new posts in threads.
 
@@
 }
-The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function.
+The \texttt{handleRequest} function checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function.
 If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}.
@@
 
-There are firstly many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}.
+There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}.
 Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence to their security.

### 2013-07-29 GCrypt/gcrypt_idr.c
GCrypt intermediate library for hashes

@@ -1,0 +1,135 @@
+#include <gcrypt.h>
+#include <gcrypt-module.h>
+#include <gpg-error.h>
+#include <stdbool.h>
+#include <stdio.h>
+#include "gcrypt_idr.h"
+
+// LibGCrypt bindings for Idris. Provides a layer between the
+// Idris code and the GCrypt code, to provide convenience functions
+// and datatype marshalling.
+
+
+
+// Initialises the libgcrypt library.
+// A required version may be specified: if this is
+// NULL, then GCRYPT_VERSION will be used instead.
+const char* idris_gcry_init(const char* version) {
+    const char* ret;
+    if (version != NULL) {
+        ret = gcry_check_version(version);
+    } else {
+        ret = gcry_check_version(GCRYPT_VERSION);
+    }
+
+    if (ret) {
+        // We're ignoring secure memory for the moment,
+        // although it may be good to add in future.
+        gcry_control (GCRYCTL_DISABLE_SECMEM, 0);
+        gcry_control (GCRYCTL_INITIALIZATION_FINISHED, 0);
+    }
+    return ret;
+}
+
+// Returns the gcry_error_t value from a call
+gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct) {
+    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
+    return struct_ptr->err;
+}
+
+// Returns the data associated with a call
+void* idris_gcry_get_struct_data(void* gcry_res_struct) {
+    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
+    return struct_ptr->res;
+}
+
+// Frees the result of an idris_grcy_res struct
+void idris_gcry_dispose_res(void* gcry_res_struct) {
+    free(gcry_res_struct);
+}
+
+// Initialises a result structure, setting the initial value of the data ptr to NULL
+idris_gcry_res* init_res_struct() {
+    idris_gcry_res* ret = malloc(sizeof(idris_gcry_res));
+    ret->res = NULL;
+    return ret;
+}
+
+// Opens a hashing context
+// TODO: flags should be more idris-friendly... Maybe an array?
+void* idris_gcry_init_md(int algorithm_key, unsigned int flags) {
+    idris_gcry_res* result_struct = init_res_struct();
+    // Allocate the context memory, perform the call
+    gcry_md_hd_t* context = malloc(sizeof(gcry_md_hd_t));
[…]

### 2013-07-29 GCrypt/gcrypt_idr.h

@@ -1,0 +1,23 @@
+#ifndef IDRIS_GCRYPT_H
+#define IDRIS_GCRYPT_H
+
+typedef struct idris_gcry_res {
+    gcry_error_t err;
+    void* res;
+} idris_gcry_res;
+
+const char* idris_gcry_init(const char* version);
+gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct);
+void* idris_gcry_get_struct_data(void* gcry_res_struct);
+void idris_gcry_dispose(void* gcry_res_struct);
+idris_gcry_res* init_res_struct();
+void* idris_gcry_init_md(int algorithm_key, unsigned int flags);
+gcry_error_t idris_gcry_enable_algorithm(void* context, int algorithm_key);
+void idris_gcry_dispose_md(void* context);
+void idris_gcry_reset_md(void* context);
+char* idris_hash_string(void* context, char* str, int algorithm_key);
+
+
+
+
+#endif

### 2013-07-30 paper/main.tex
Started on GCrypt handler bindings, finished minor revisions on paper
...although major revisions, incl conclusion, yet to do...

@@
 \begin{abstract}
-Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing easier reasoning about the operation of such programs.
+Dependently-typed languages allow for much expressive types to be used within programs in turn allowing easier reasoning about the operation of such programs.
 Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.
 
-Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and show that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
-Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection, which exploit the loss of type information.
+Embedded domain-specific languages (EDSLs) address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
+
+In this paper, we apply this technique to web programming, and introduce an EDSL to facilitate the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection.
+We also show how EDSLs may be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
 \end{abstract}
@@
 \section{Introduction}
-Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits.
-The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development.
+Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection or cross-site scripting.
+Many web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development.
 A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.
@@
 \subsection{Contributions}
-The primary contribution of this paper is the application of previous work to show that dependent types may be used to provide stronger static guarantees about the correctness and security of web applications, whilst minimising additional overhead.
+The primary contribution of this paper is the application of previous work to show that dependent types may be used to provide stronger static guarantees about the correctness and security of web applications, whilst minimising additional development complexity.
 In particular:
@@
 \begin{itemize}
+\item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
 \item Representations of CGI, SQLite and a simple session handler as \textit{algebraic effects}, allowing programs to be accepted only when they adhere to specific resource usage protocols.
-\item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
 \item A message board application making use of the framework, and showing that the framework may be used for practical applications.
@@
 We structure the remainder of this paper as follows.
-We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocol for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
+We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
 
@@
 \label{rup}
-In this section, we show how three such effects; CGI, SQLite and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
+In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
 
@@
 CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the user prior to any other output; failure to do so will result in an internal server error.
-
-A previous implementation of CGI in \idris{} implemented CGI as an extension of monadic IO.
+Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.
+
+A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
 Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
@@
 
-By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
+By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
 In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
 This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI effect as the unit type, ().
+We represent an uninitialised CGI process as the unit type, ().
 {\small
@@
 Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
-Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the resource.
+Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.
 
 Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved.
-The only operation that can be performed in this state is initialisation: by calling \texttt{initalise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
+The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
 The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
@@
 
-In order to make use of the CGI framework, users specify a function of the following type:
+In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
+We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
 {\small
@@
 
+A simple ``Hello, world!'' program would be defined as follows:
+
+{\small
+\begin{verbatim}
+sayHello : CGIProg [] ()
+sayHello = output "Hello, world!"
+
+main : IO ()
+main = do runAction [initCGIState] sayHello
+          pure ()
+\end{verbatim}
+}
 % =================================================
 
-\subsection{SQLite}
+\subsection{Database access with SQLite}
 
@@
 }
-\caption{SQLite Resource Usage Protocol}
+\caption{Database Resource Usage Protocol}
 \label{fig:sqlitestates}
@@
 We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{brady2011idris}, which is beyond the scope of this paper.
-
-The creation, preparation and execution of SQL queries has a very specific usage protocol, with several possible points of failure.
+
+The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
 Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
@@
 
-Figure ~\ref{fig:sqlitestates} shows the resource usage protocol for the SQLite library.
-There are three main phases involved in the usage of the SQLite effect: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the resource associated with the SQLite effect, which again is parameterised by the current state.
+Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
+There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
+This is reflected in the associated resource, which again is parameterised by the current protocol state.
 {\small
@@
 At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial state.
+The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
 
 %Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the library function \texttt{sqlite3\_step()}.
+SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
 In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
@@
 }
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
+Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
 Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
@@
 The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.
+This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.
 
@@
 \label{form}
-Programming web applications often involves processing user data, which may then be used in further effectuful computations.
+Programming web applications often involves processing user data, which may then be used in further effectful computations.
 Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.
@@
 }
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between abstract and concrete representations of allowed form types:
+Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
 {\small
@@
 %
-In order to specify a form, we once again make use of the \texttt{Effects} library in order to create a resource-aware EDSL.
+In order to specify a form, we once again use \texttt{Effects}.
 By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
@@
 Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function.
+Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
 Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
@@
 }
-The \texttt{mkHandlerFn} function takes as its argument a tuple describing the arguments and web effects available to the function.
+The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
 When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-We may then extract the information through pattern matching.
 
@@
 \label{messageboard}
-Using these components, we may now consider a larger example of a message board which allowsusers to register, log in, view and create threads, and list and create new posts in threads.
+Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
 
@@
 }
-The \texttt{handleRequest} function checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function.
-If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}.
+We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
+If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
 If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
@@
 \acks
+This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
+The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
 % SICSA / EPSRC (grant number?

### 2013-07-30 GCrypt/gcrypt_idr.c

@@ -13,10 +13,10 @@
 
 // Initialises the libgcrypt library.
 // A required version may be specified: if this is
-// NULL, then GCRYPT_VERSION will be used instead.
+// the empty string, then GCRYPT_VERSION will be used instead.
 const char* idris_gcry_init(const char* version) {
     const char* ret;
-    if (version != NULL) {
+    if (strlen(version) >= 0) {
         ret = gcry_check_version(version);
     } else {
         ret = gcry_check_version(GCRYPT_VERSION);
@@ -114,7 +114,7 @@ char* idris_hash_string(void* context, char* str, int algorithm_key) {
     return str_output;
 }
 
-
+/*
 int main(int argc, char** argv) {
     if (argc < 2) {
         printf("Usage: thing [arg]");
@@ -133,3 +133,4 @@ int main(int argc, char** argv) {
         fprintf(stderr, "Failed to initialise");
     }
 }
+*/

### 2013-07-30 GCrypt/gcrypt_idr.h

@@ -16,6 +16,7 @@ gcry_error_t idris_gcry_enable_algorithm(void* context, int algorithm_key);
 void idris_gcry_dispose_md(void* context);
 void idris_gcry_reset_md(void* context);
 char* idris_hash_string(void* context, char* str, int algorithm_key);
+void idris_gcry_dispose_res(void* gcry_res_struct);
 
 
 

### 2013-07-30 paper/main.tex
More work on paper

@@
 \usepackage[hidelinks]{hyperref}
+\usepackage{fancyvrb}
 \usetikzlibrary{arrows,automata}
@@
 \begin{document}
+\fvset{fontsize=\small}
 \newcommand{\idris}{\textsc{Idris}}
@@
 \section{Introduction}
-Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection or cross-site scripting.
-Many web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development.
+Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection \cite{owasp:sqli} or cross-site scripting \cite{owasp:xss}.
+Security breaches using such exploits are far-reaching, and high profile cases involve large corporations such as Sony, who suffered a well-publicised and extremely costly SQL Injection breach in 2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012 \cite{imperva:yahoo}.
+
+Many web applications are written in dynamically-checked scripting languages such as PHP, Ruby or Python, which facilitate rapid development \cite{w3techs:webpls}.
 A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.
@@
 We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:
-
-{\small
-\begin{verbatim}
+%
+\begin{Verbatim}
 $conn = new mysqli("localhost", "username", 
                     "password", "db");
-\end{verbatim}
-}
-
+\end{Verbatim}
+%
 We must subsequently check to see if the connection was successful, and cease execution of the database routine if so.
 Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.
-
-{\small
-\begin{verbatim}
+%
+\begin{Verbatim}
   if (mysqli_connect_errno()) {
@@
   }
-\end{verbatim}
-}
-
+\end{Verbatim}
+%
 We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:
-
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   $stmt = $conn->prepare("SELECT `name`, `address` 
      FROM `staff` WHERE `dept` = ?);
-  $stmt->bind_param($dept);
-\end{verbatim}
-}
-
+  $stmt->bind_param('s', $dept);
+\end{Verbatim}
+%
 After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn.
 At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:
-
-{\small
-\begin{verbatim}
+%
+\begin{Verbatim}
   $stmt->execute();
   $stmt->bind_result($name, $address);
-
   while ($stmt->fetch()) {
@@
   }
-\end{verbatim}
-}
-
+\end{Verbatim}
+%
 Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:
-
-{\small
-\begin{verbatim}
+%
+\begin{Verbatim}
   $stmt->close();
   $conn->close();
-\end{verbatim}
-}
-
-Even in this small example, there exists a very specific resource usage protocol.
+\end{Verbatim}
+%
+Even in this small example, there exists a specific resource usage protocol.
 Firstly, a connection to the database must be opened.
@@
 Describing a program in the \texttt{Effects} library is achieved using the following data type:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 EffM : (m : Type -> Type) ->
@@
        (a : Type) -> Type
-\end{verbatim}
-}
+\end{Verbatim}
 \texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}).
@@
 Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
-{\small
 %TODO: Execution
-\begin{verbatim}
+\begin{Verbatim}
 Effect : Type
@@
          (t : Type) -> Type
-\end{verbatim}
-}
+\end{Verbatim}
 Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 queryPostVar : 
@@
 queryPostVar query = (QueryPostVar query)
-\end{verbatim}
-}
+\end{Verbatim}
 Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context.
@@
 We represent an uninitialised CGI process as the unit type, ().
-{\small
-\begin{verbatim}
+\begin{Verbatim}[samepage]
 data CGIStep = Initialised 
@@
   ICgi : CGIInfo -> InitialisedCGI s
-\end{verbatim}
-}
+\end{Verbatim}
 Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
@@
 
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 data Cgi : Effect where
@@
                            (InitialisedCGI TaskRunning) a
-\end{verbatim}
-}
+\end{Verbatim}
 Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
@@
 We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 CGIProg : List EFFECT -> Type -> Type
@@
  Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{verbatim}
-}
+\end{Verbatim}
 This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
@@
 
-{\small
-\begin{verbatim}
+\begin{Verbatim}
+import Cgi
+import Effects
+
 sayHello : CGIProg [] ()
@@
           pure ()
-\end{verbatim}
-}
+\end{Verbatim}
 % =================================================
@@
 Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
+This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.
+
+\begin{Verbatim}[samepage]
+data Sqlite : Effect where
+  OpenDB : 
+    String -> 
+    Sqlite () 
+           (SQLiteRes ConnectionOpened) Bool
+  CloseDB : 
+    Sqlite (SQLiteRes ConnectionOpened) () Bool
+  PrepareStatement : 
+    String -> 
+    Sqlite (SQLiteRes ConnectionOpened) 
+           (SQLiteRes PreparedStatementOpen) Bool
+  StartBind : 
+    Sqlite (SQLiteRes PreparedStatementOpen) 
+           (SQLiteRes PreparedStatementBinding) ()
+  BindText : 
+    ArgPos -> String -> Int -> 
+    Sqlite (SQLiteRes PreparedStatementBinding) 
+           (SQLiteRes PreparedStatementBinding) 
+           Bool
+  FinishBind : 
+    Sqlite (SQLiteRes PreparedStatementBinding) 
+           (SQLiteRes PreparedStatementBound) 
+           Bool
+  ExecuteStatement : 
+    Sqlite (SQLiteRes PreparedStatementBound) 
+           (SQLiteRes PreparedStatementExecuting) ()
+  Finalise : 
+    Sqlite (SQLiteRes PreparedStatementExecuting) 
+           (SQLiteRes ConnectionOpened) 
+           Bool
+  GetColumnText : 
+    Int -> 
+    Sqlite (SQLiteRes PreparedStatementExecuting)
+           (SQLiteRes PreparedStatementExecuting)
+           String
+  RowStep : 
+    Sqlite (SQLiteRes PreparedStatementExecuting)
+           (SQLiteRes PreparedStatementExecuting)
+           StepResult
+\end{Verbatim}
 There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource, which again is parameterised by the current protocol state.
-{\small
-\begin{verbatim}
+This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
+\begin{Verbatim}
 data SQLiteRes : Step -> Type where
@@
                  | InvalidStmt 
-\end{verbatim}
-}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure.
+\end{Verbatim}
+If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
 At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
@@
 We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 data StepResult = Unstarted
@@
                 | NoMoreRows
-\end{verbatim}
-}
+\end{Verbatim}
 Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
@@
 We define a function of type:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 String ->
 Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{verbatim}
-}
+\end{Verbatim}
 This means that the program will be run in the IO execution context, and must start and end with no active resources.
 The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 testSelect : String -> Eff IO [SQLITE ()] 
@@
           return $ Left err $
-\end{verbatim}
-}
+\end{Verbatim}
 The program initially attempts to open a connection to the \texttt{people.db} database.
@@
 This function is then recursively called until there are no more rows to process.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 collectResults : 
@@
       StepFail => return [] $
-\end{verbatim}
-}
-
+\end{Verbatim}
 In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
 In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 data DBVal = DBInt Int
@@
            | DBNull
-\end{verbatim}
-}
+\end{Verbatim}
 We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   ResultSet : Type
   ResultSet = List (List DBVal)
-\end{verbatim}
-}
+\end{Verbatim}
 Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-{\small
-\begin{verbatim}
-executeSelect : String ->
-                String -> 
-                List (Int, DBVal) -> 
-                (Eff m [SQLITE 
-                  (SQLiteRes PreparedStatementExecuting)
-                   ] 
-                 (List DBVal)
-                ) -> 
+\begin{Verbatim}
+executeSelect : 
+  String -> String -> List (Int, DBVal) -> 
+  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
+                 (List DBVal)) -> 
   Eff m [SQLITE ()] (Either String ResultSet)
-\end{verbatim}
-}
+\end{Verbatim}
 % =================================================
-
 \subsection{A Simple Session Handler}
@@
 A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
-
 %TODO: Fix up the labels
+\begin{Verbatim}[samepage]
+data Session : Effect where
+  LoadSession : 
+  SessionID -> 
+  Session (SessionRes SessionUninitialised) 
+          (SessionRes SessionInitialised) 
+          (Maybe SessionData)
+  UpdateSession : 
+    SessionData -> 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionInitialised) ()
+  CreateSession : 
+    SessionData -> 
+    Session (SessionRes SessionUninitialised) 
+            (SessionRes SessionInitialised) 
+            (Maybe SessionID)
+  DeleteSession : 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionUninitialised) Bool 
+  WriteToDB : 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionUninitialised) Bool
+  DiscardSessionChanges : 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionUninitialised) ()
+  GetSessionID : 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionInitialised) 
+            (Maybe SessionID)
+  GetSessionData : 
+    Session (SessionRes SessionInitialised) 
+            (SessionRes SessionInitialised) 
+            (Maybe SessionData)
+\end{Verbatim}
 \begin{figure}[htpb!]
@@
 \end{figure}
-
 Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler.
@@
 Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 data SessionStep = SessionUninitialised
@@
                    SessionRes s
-\end{verbatim}
-}
+\end{Verbatim}
 The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
@@
 
-{\small
-\begin{verbatim}
-SELECT `name`, `address` FROM `staff` LIMIT ' + page + ', 5
-\end{verbatim}
-}
+\begin{Verbatim}
+'SELECT `name`, `address` FROM `staff` LIMIT ' + page + '
+                                                     , 5'
+\end{Verbatim}
 The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
@@
 In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 data FormTy = FormString
@@
             | FormList FormTy 
-\end{verbatim}
-}
+\end{Verbatim}
 Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 interpFormTy : FormTy -> Type
@@
 interpFormTy (FormList a) = List (interpFormTy a)
-\end{verbatim}
-}
+\end{Verbatim}
 %
@@
 
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 using (G : List FormTy, E : List WebEffect)
@@
              Form (FormRes G E) (FormRes [] []) String
-\end{verbatim}
-}
+\end{Verbatim}
 The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
 
-The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a list of the types associated with each element and the effects supported by the handler function.
-The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
+We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form.
+The resource allows us to record the types associated with each form element and the HTML required to display the form.
+he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
+
+By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
 It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
@@
 By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource, which is returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
+Additionally, HTML for the form element is generated, and stored in the resource.
+The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
 
 To specify a form instance, we define a function of type \texttt{UserForm}:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   UserForm : Type
@@
                           (FormRes [])] String
-\end{verbatim}}
+\end{Verbatim}
 All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
@@
 By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   useEffects : (effs : List WebEffect) ->
@@
   useEffects effs = (UseEffects effs)
-\end{verbatim}}
+\end{Verbatim}
 Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
+This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
 %
 We may specify a handler function of type \texttt{FormHandler}:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   FormHandler : List EFFECT -> Type
   FormHandler effs = Eff IO effs ()
-\end{verbatim}
-}
+\end{Verbatim}
 In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-{\small
-\begin{verbatim}
-  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
+\begin{Verbatim}
+  addSubmit : 
+    (f :  mkHandlerFn ((reverse G), E)) ->
     (fns : HandlerList) ->
-              {default tactics { applyTactic 
-                findFn 100; solve; }
+    {default tactics 
+      { applyTactic findFn 100; solve; }
       prf : FnElem f fns} ->
@@
           name = getString' f handlers prf          
-\end{verbatim}
-}
+\end{Verbatim}
+
 Let us look at each aspect of this function in turn.
@@
 Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 MkHandlerFnTy : Type
@@
 mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{verbatim}
-}
+\end{Verbatim}
 The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
@@
 To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 HandlerFn : Type
 HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{verbatim}
-}
+\end{Verbatim}
 %
@@
 
-\begin{verbatim}
+\begin{Verbatim}
 HandlerList : Type
 HandlerList = List HandlerFn
-\end{verbatim}
+\end{Verbatim}
 To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
   data FnElem : mkHandlerFn ((reverse G), E) -> 
@@
                FnElem f xs -> FnElem f (x :: xs)
-\end{verbatim}
-}
+\end{Verbatim}
 \texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
@@
 :P
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
@@
                              (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = Just reg_fn
+  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
+                                            Just reg_fn
   checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{verbatim}}
+\end{Verbatim}
 We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
@@
 \subsection{Handling requests}
-Once the program is invoked via a CGI request, the \texttt{main} function will be called.
-We then run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-
-{\small
-\begin{verbatim}
+The entry point to any CGI application is the \texttt{main} function.
+From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
+\begin{Verbatim}
 main : IO ()
@@
           pure ()
-\end{verbatim}
-}
+\end{Verbatim}
 We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
@@
 If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 handleRequest : CGIProg [SESSION 
@@
     handleNonFormRequest action (map strToInt thread_id)
-\end{verbatim}
-}
+\end{Verbatim}
 \subsection{Thread Creation}
@@
 One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 newThreadForm : UserForm
@@
   addSubmit handleNewThread handlers
-\end{verbatim}
-}
+\end{Verbatim}
 The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
@@
 The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 handlers : HandlerList
-handlers = [(([FormString, FormString], 
+handlers = 
+  [(([FormString, FormString], 
      [CgiEffect, SessionEffect, SqliteEffect]) ** 
@@
   ]
-\end{verbatim}
-}
+\end{Verbatim}
 Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
@@
 Should the form handler function be called with invalid arguments, an error is shown.
-{\small
-\begin{verbatim}
-handleNewThread : Maybe String -> 
-                  Maybe String -> 
-                  FormHandler [CGI (InitialisedCGI 
-                                 TaskRunning), 
-                               SESSION (SessionRes 
-                                 SessionUninitialised), 
+\begin{Verbatim}
+handleNewThread : 
+  Maybe String -> Maybe String -> 
+  FormHandler [CGI (InitialisedCGI TaskRunning), 
+               SESSION (SessionRes SessionUninitialised), 
                SQLITE ()]  
@@
   pure ()
-\end{verbatim}
-}
+\end{Verbatim}
 Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
@@
 In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 threadInsert : Int -> 
@@
     Right thread_id => postInsert uid thread_id content
-\end{verbatim}
-}
+\end{Verbatim}
 \subsection{Listing Threads}
 Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = executeSelect DB_NAME query [] collectThreadResults
+getThreads = 
+  executeSelect DB_NAME query [] collectThreadResults
  where query = "SELECT `ThreadID`, `Title`, `UserID`, 
    `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{verbatim}
-}
+\end{Verbatim}
 Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
 Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 traverseThreads : ResultSet -> 
@@
         traverseRow _ = pure ()
-\end{verbatim}
-}
+\end{Verbatim}
 \subsection{Authentication}
@@
 Any failures, for example with creating a new session or querying the database, are reported to the user.
-{\small
-\begin{verbatim}
+\begin{Verbatim}
 handleLoginForm (Just name) (Just pwd) = do
@@
       pure ()
-\end{verbatim}}
+\end{Verbatim}
 Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
@@
 Dependently-typed languages have, for a long time, remained relatively unused outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough tyeps) to be adopted without the additional proofs traditionally required by users of such languages.
-
+Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
+
+Our framework, through the use of dependent types, provides several static guarantees.
+Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
+This immediately eliminates the possibilities of SQL injection attacks on non-string types.
+Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
+Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
+
+Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
+Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
+By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
+Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
+
+In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
+With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
+Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
+
+Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
 \subsection{Further Work}
@@
 
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence to their security.
-
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request request.
-Ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
-
-Dependently-typed languages provide great promise for the construction of secure and correct programs. Through the use of embedded domain-specific languages, we hope that
+There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
+Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
+
+Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
+We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
+
+Dependently-typed languages provide great promise for the construction of secure and correct programs.
+Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
 % Cryptography bindings would be really good.

### 2013-07-31 paper/main.tex
Draft IFL paper complete

@@
 \usepackage{amsmath}
-\usepackage{silence}
+%\usepackage{silence}
 \usepackage{pgf}
@@
 \usepackage{fancyvrb}
+\usepackage[numbers]{natbib}
 \usetikzlibrary{arrows,automata}
-\WarningsOff
+%\WarningsOff
 
@@
 \newcommand{\idris}{\textsc{Idris}}
+\newcommand{\idata}{\textsf{iData}}
+\newcommand{\itasks}{\textsf{iTasks}}
 \special{papersize=8.5in,11in}
@@
 
-\titlebanner{Draft for submission to IFL 2013}        % These are ignored unless
+\titlebanner{}        % These are ignored unless
 \preprintfooter{}   % 'preprint' option specified.
@@
 \begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within programs in turn allowing easier reasoning about the operation of such programs.
+Dependently-typed languages allow much expressive types to be used within programs, in turn allowing easier reasoning about the operation of such programs.
 Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.
@@
 \begin{Verbatim}
+module Main
 import Cgi
-import Effects
 
@@
 main : IO ()
-main = do runAction [initCGIState] sayHello
-          pure ()
+main = runCGI [initCGIState] sayHello
 \end{Verbatim}
@@
   else do err <- connFail
-          return $ Left err $
+          return $ Left err 
 \end{Verbatim}
@@
       NoMoreRows => return []
-      StepFail => return [] $
+      StepFail => return [] 
 \end{Verbatim}
@@
 \begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + page + '
-                                                     , 5'
+'SELECT `name`, `address` FROM `staff` LIMIT ' + 
+       page + ', 5';
 \end{Verbatim}
@@
 
+We begin with a simple example of a form which requests a user's name, and echoes it back.
+Firstly, we define a form handler which echoes back a string provided by the form handler.
+It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.
+
+\begin{Verbatim}
+sayHello : Maybe String -> 
+           FormHandler [CGI (InitialisedCGI TaskRunning)]
+sayHello (Just name) = output ("Hello, " ++ name ++ "!")
+sayHello _ = output "Error!"
+\end{Verbatim}
+
+We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
+
+\begin{Verbatim}
+handlers : HandlerList
+handlers = [(handler args=[FormString], 
+                     effects=[CgiEffect], 
+                     fn=sayHello, 
+                     name="sayHello")
+           ]
+\end{Verbatim}
+
+We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
+
+\begin{Verbatim}
+showHelloForm : UserForm
+showHelloForm = do
+  addTextBox "Name" FormString Nothing
+  useEffects [CgiEffect]
+  addSubmit sayHello handlers
+\end{Verbatim}
+
+Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
+
+\begin{Verbatim}
+cgiHello : CGIProg [] ()
+cgiHello = do
+  handler_set <- isHandlerSet
+  if handler_set then do
+    handleForm handlers
+    return ()
+  else do
+    addForm "nameform" "helloform" showHelloForm
+    return ()
+
+main : IO ()
+main = runCGI [initCGIState] cgiHello
+\end{Verbatim}
+When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user.
+Upon submission of the form, the form handler will be invoked, and the name will be used in the output.
+
+In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
 \subsection{Form Construction}
+\label{formcons}
 Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
@@
 We also store a unique string identifier, which is used to serialise a reference to the handler function.
+
+In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
+\begin{Verbatim}
+syntax 
+  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
+  ", name=" [name] = ((args, effs) ** (fn, name))
+\end{Verbatim}
+We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
+This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
+\begin{Verbatim}
+handler args=[FormString], 
+        effects=[CgiEffect], 
+        fn=sayHello, 
+        name="sayHello"
+\end{Verbatim}
+
 We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-
 \begin{Verbatim}
@@
 handlers : HandlerList
-handlers = 
-  [(([FormString, FormString], 
-     [CgiEffect, SessionEffect, SqliteEffect]) ** 
-     (handleRegisterForm, "handleRegisterForm")),
-
-   (([FormString, FormString], 
-     [CgiEffect, SessionEffect, SqliteEffect]) ** 
-     (handleNewThread, "handleNewThread")),
+handlers = [
+ (handler args=[FormString, FormString], 
+          effects=[CgiEffect, SessionEffect, SqliteEffect], 
+          fn=handleRegisterForm, 
+          name="handleRegisterForm"),
+
+ (handler args=[FormString, FormString], 
+          effects=[CgiEffect, SessionEffect, SqliteEffect], 
+          fn=handleNewThread, 
+          name="handleNewThread"),
      ...
@@
                      (DBText username)::[]) =
-           (output \$ "<tr><td>
+           (output $ "<tr><td>
             <a href=\"?action=showthread&thread_id=" ++ 
@@
       if set_sess_res then do
-        lift' (output \$ "Welcome, " ++ name)
+        lift' (output $ "Welcome, " ++ name)
         pure ()
@@
     Left err => do
-      lift' (output \$ "Error: " ++ err)
+      lift' (output $ "Error: " ++ err)
       pure ()
@@
 \section{Related Work}
+\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser.
+The library also provides support for cookies and basic form handling.
+
+\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods.
+In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs.
+Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs.
+Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations.
+Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.
+
+\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction.
+Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven.
+This means that upon a component being edited, a computation occurs, given the state of the current form.
+This is then saved to be used in future computations.
+Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted.
+This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.
+
+The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another.
+This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion.
+These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application.
+Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.
+
+Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs.
+Through the use of the Ur\/Web DSL, developers may generate well-typed code from records.
+By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure.
+Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.
+
+WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres.
+WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes.
+The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables.
+These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy.
+Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing.
+WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation.
+We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.
 
@@
 \section{Conclusions}
-Dependently-typed languages have, for a long time, remained relatively unused outside of academia.
+Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
 Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
@@
 By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
+Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
 
@@
 We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
+
+Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
+Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
+Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
+
+Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
+Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
+If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
+This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
+Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
+Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
 

### 2013-08-14 paper/main.tex
Update introduction

--- 2013-07-30 paper/main.tex
@@
 \begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within programs in turn allowing easier reasoning about the operation of such programs.
-Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an EDSL to facilitate the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection.
-We also show how EDSLs may be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
+Dependently-typed languages allow for much expressive types to be used within
+programs in turn allowing easier reasoning about the operation of such
+programs. Stronger specifications do however bring with them the disadvantage
+that it becomes increasingly difficult to write programs that are accepted by
+the type checker and additional proofs may have to be specified by a user.
+
+Embedded domain-specific languages (EDSLs) address this problem by introducing
+a layer of abstraction over more specific underlying types, allowing
+domain-specific code to be written in high-level languages which use dependent
+types to enforce certain invariants without additional proof obligations.
+
+In this paper, we apply this technique to web programming, and introduce an
+EDSL to facilitate the creation and handling of web forms which retain type
+information, reducing the scope for programmer error and attacks such as SQL
+injection. We also show how EDSLs may be constructed to enforce resource usage
+protocols associated with common web operations such as CGI, database access
+and session handling.
+
 \end{abstract}
@@
 \section{Introduction}
-Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection \cite{owasp:sqli} or cross-site scripting \cite{owasp:xss}.
-Security breaches using such exploits are far-reaching, and high profile cases involve large corporations such as Sony, who suffered a well-publicised and extremely costly SQL Injection breach in 2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012 \cite{imperva:yahoo}.
-
-Many web applications are written in dynamically-checked scripting languages such as PHP, Ruby or Python, which facilitate rapid development \cite{w3techs:webpls}.
-A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.
+
+Web applications, whilst ubiquitous, are also prone to incorrect construction
+and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
+scripting \cite{owasp:xss}. Security breaches using such exploits are
+far-reaching, and high profile cases involve large corporations such as Sony,
+who suffered a well-publicised and extremely costly SQL Injection breach in
+2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
+\cite{imperva:yahoo}.
+
+Many web applications are written in dynamically-checked scripting languages
+such as PHP, Ruby or Python, which facilitate rapid development
+\cite{w3techs:webpls}. A significant drawback, however, is that such languages
+do not provide  the same static guarantees about runtime behaviour afforded by
+programs with more expressive, static type systems, instead relying on
+extensive unit testing to ensure correctness and security.
 
 %%------- EXAMPLE
-Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}.
-We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:
+Let us consider a simple database access routine, written in
+PHP, where we wish to obtain the name and address of every employee working in
+a given department, \texttt{\$dept}. We firstly construct an object
+representing a database connection, where the arguments are the database host,
+user, password and name respectively:
+
 %
@@
 %
-We must subsequently check to see if the connection was successful, and cease execution of the database routine if so.
-Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.
+We then check to see if the connection was successful, and exit
+if not.  This check is optional, so
+it would be possible to omit it: this would, however, cause
+problems with later steps.
+
 %
@@
 %
-We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:
+We then create a prepared statement detailing our query, and bind the
+\texttt{`dept'} value:
+
 \begin{Verbatim}
@@
 %
-After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn.
-At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:
+After the parameters have been bound, we execute the statement, assign
+variables into which results will be stored, and fetch each row in turn. At
+this point, a developer may forget to execute the statement before beginning to
+fetch results or may have forgotten to bind the parameters to the statement:
+
 %
@@
 %
-Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:
+Finally, once the statement and connection are no longer needed, they should be
+closed in order to discard the associated resources:
+
 %
@@
 Even in this small example, there exists a specific resource usage protocol.
-Firstly, a connection to the database must be opened.
-The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code.
-Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound.
-The statement is then executed, and each row is retrieved from the database.
+Firstly, a connection to the database must be opened. The object-oriented style
+used in the example encapsulates this to an extent, as the object must be
+created in order for operations to be performed, however it is less obvious in
+a procedural version of the code. Secondly, a prepared statement is created,
+using the raw SQL and placeholders to which variables are later bound. The
+statement is then executed, and each row is retrieved from the database.
 Finally, the resources are freed.
 
-There is, however, no way to statically guarantee that this protocol is adhered to.
-A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error.
-Similarly, a programmer may omit closing the statement or connection, which can lead to substantial problems such as handle leaks in longer-running server applications.
-
-An alternative approach is to ensure that a program is semantically sound before it is run.
-Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled prior to being accepted by a type checker.
-A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer.
-This severely limits the uptake of dependently-typed languages in real-world settings.
-
-This complexity can be ameliorated through the use of \textit{embedded domain-specific languages} (EDSLs), which aim to abstract away from the complexity of the underlying dependent type mechanisms, allowing users to write verfied domain-specific code without having to consider the underlying verification mechanisms.
-%TODO: rephrase this
-
-\idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros.
-Through the use of \idris{}, and the \texttt{Effects} framework \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.
+Problems may arise if the protocol is not followed correctly.
+A developer may, for example, accidentally close a statement whilst still
+retrieving rows, which would cause a runtime error. Similarly, a programmer may
+omit closing the statement or connection, which can lead to
+problems such as resource leaks in longer-running server applications.
+However, in conventional programming languages, there is no way to check
+automatically that a protocol is followed.
+
+In contrast, the use of \textit{dependent types} makes it possible
+to specify a program's behaviour precisely, and to check that a 
+specification is followed.
+%
+The difficulty is 
+that automatic verification by a compiler can be difficult or
+often impossible, requiring additional proofs to be given by the programmer.
+
+This complexity can be ameliorated through the use of \textit{embedded
+domain-specific languages} (EDSLs), which aim to abstract away the
+complexity of the underlying type system. EDSLs allow domain experts to
+focus on a problem domain, rather than the specific details of their
+solution.
+
+\idris{} \cite{brady2011idris} is a language with full dependent types, and
+extensive support for EDSLs through overloading and syntax macros. Through the
+use of \idris{}, and a framework for describing resource protocols using
+\emph{algebraic effects}~\cite{brady:effects}, we
+present a dependently-typed web framework, which allows the construction of
+programs with additional guarantees about correctness and security, whilst
+minimising the increase in development complexity.
 
 \subsection{Contributions}
-The primary contribution of this paper is the application of previous work to show that dependent types may be used to provide stronger static guarantees about the correctness and security of web applications, whilst minimising additional development complexity.
-In particular:
+The primary contribution of this paper is the application of 
+dependent types may to provide strong static guarantees
+about the correctness and security of web applications, whilst minimising
+additional development complexity. In particular, we present:
 
 \begin{itemize}
-\item A form-handling mechanism, which preserves type information and therefore increases applications' resilience to attacks such as SQL injection.
-\item Representations of CGI, SQLite and a simple session handler as \textit{algebraic effects}, allowing programs to be accepted only when they adhere to specific resource usage protocols.
-\item A message board application making use of the framework, and showing that the framework may be used for practical applications.
+\item A form-handling mechanism, which preserves type information and
+manages user input, therefore
+increasing applications' resilience to attacks such as SQL injection.
+
+\item Representations of CGI, Databases and sessions as
+\textit{algebraic effects}, allowing programs to be accepted only when they
+follow clearly defined resource usage protocols.
+
+\item A practical demonstration of the framework, a message board application.
+
 \end{itemize}
@@
 \label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing greater code modularity.
-This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.
-
-An effect is represented as an instance of the \texttt{EFFECT} type. Such instances may be associated with a \textit{resource}, 
-Describing a program in the \texttt{Effects} library is achieved using the following data type:
+\texttt{Effects} is an EDSL which handles side-effects such as database access
+as \textit{algebraic effects}, allowing greater code modularity. This section
+is intended to serve as a brief introduction to the framework; for more
+information, see \cite{brady:effects}.
+
+An effect is represented as an instance of the \texttt{EFFECT} type. Such
+instances may be associated with a \textit{resource}, Describing a program in
+the \texttt{Effects} library is achieved using the following data type:
+
 \begin{Verbatim}
@@
 \end{Verbatim}
-\texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}).
-The list of input types can be used to ensure that an operation may only be called when the program is in a particular state, and since the list of input effects may differ from the list of output effects, we may modify the resource associated with the state after each operation.
-This allows us to simply enforce a resource usage protocol, and deviation from this protocol will result in a compile-time error.
-
-Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect.
-Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
+\texttt{EffM} is a type that is parameterised over an execution context
+(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
+of effects at the end of the program's execution (\texttt{es'}), and a return
+type (\texttt{a}). The list of input types can be used to ensure that an
+operation may only be called when the program is in a particular state, and
+since the list of input effects may differ from the list of output effects, we
+may modify the resource associated with the state after each operation. This
+allows us to simply enforce a resource usage protocol, and deviation from this
+protocol will result in a compile-time error.
+
+Implementing a new effect involves creating a new algebraic data type, of kind
+\texttt{Effect}, which describes the available operations and how they modify
+the resource associated with the effect. Each operation consists of an initial
+resource \texttt{res}, the resource at the end of the operation \texttt{res'},
+and the output type of the operation \texttt{t}.
+
 %TODO: Execution
@@
 \end{Verbatim}
-Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
+
+Operations are made available to users by writing functions with the
+corresponding \texttt{EffM} type, which may then be used to invoke the
+operation from within an effectful program.
+
 \begin{Verbatim}
@@
 \end{Verbatim}
-Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context.
-Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.
+
+Additionally, we define \textit{handlers}, which describe how each operation is
+interpreted in a given execution context. Finally, \texttt{Effect} instances
+are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
+constructor, which is provided with the initial resource value.
 

### 2013-08-14 paper/main.tex

--- 2013-07-31 paper/main.tex
@@
 \begin{abstract}
-Dependently-typed languages allow much expressive types to be used within programs, in turn allowing easier reasoning about the operation of such programs.
-Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an EDSL to facilitate the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection.
-We also show how EDSLs may be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
+Dependently-typed languages allow for much expressive types to be used within
+programs in turn allowing easier reasoning about the operation of such
+programs. Stronger specifications do however bring with them the disadvantage
+that it becomes increasingly difficult to write programs that are accepted by
+the type checker and additional proofs may have to be specified by a user.
+
+Embedded domain-specific languages (EDSLs) address this problem by introducing
+a layer of abstraction over more specific underlying types, allowing
+domain-specific code to be written in high-level languages which use dependent
+types to enforce certain invariants without additional proof obligations.
+
+In this paper, we apply this technique to web programming, and introduce an
+EDSL to facilitate the creation and handling of web forms which retain type
+information, reducing the scope for programmer error and attacks such as SQL
+injection. We also show how EDSLs may be constructed to enforce resource usage
+protocols associated with common web operations such as CGI, database access
+and session handling.
+
 \end{abstract}

### 2013-08-14 paper/main.tex

@@
 \authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
+           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
            {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
@@
 In this paper, we apply this technique to web programming, and introduce an
-EDSL to facilitate the creation and handling of web forms which retain type
+EDSL, to facilitate the creation and handling of web forms which retain type
 information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how EDSLs may be constructed to enforce resource usage
+injection. We also show how to enforce resource usage
 protocols associated with common web operations such as CGI, database access
@@
 if not.  This check is optional, so
-it would be possible to omit it: this would, however, cause
+it would be possible to omit it. However, this would cause
 problems with later steps.
@@
 
-We structure the remainder of this paper as follows.
-We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
+We structure the remainder of this paper as follows. We provide a brief
+overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
+how this may be used to ensure adherence to resource usage protocols for CGI,
+SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
+type-safe form handling in Section ~\ref{form}, implemented using
+\texttt{Effects}; and discuss the larger example
+of a message board system making use of these components in Section
+~\ref{messageboard}.
 
@@
 \label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access
-as \textit{algebraic effects}, allowing greater code modularity. This section
-is intended to serve as a brief introduction to the framework; for more
-information, see \cite{brady:effects}.
-
-An effect is represented as an instance of the \texttt{EFFECT} type. Such
-instances may be associated with a \textit{resource}, Describing a program in
-the \texttt{Effects} library is achieved using the following data type:
+
+\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
+side-effects such as state, exceptions, and I/O as \emph{algebraic
+effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
+by an input and output \emph{state}, which permits effectful programs to track
+the progress of a resource usage protocol. Effectful programs are written
+in a monadic style, with \texttt{do}-notation, with their type stating which
+specific effects are available.
+Effectful
+programs are described using the following data type,
+in the simplest case:
+
+\begin{Verbatim}
+Eff : (m : Type -> Type) ->
+      (es : List EFFECT) -> (a : Type) -> Type
+\end{Verbatim}
+
+\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
+which describes the context in which the effectful program will be run, a
+list of side effects \texttt{es} that the program is permitted to use, and the
+program’s return type \texttt{a}. The name \texttt{m} for the computation context is
+suggestive of a monad, but there is no requirement for it to be so.
+
+For example, the following type carries an integer state,
+throws an exception of type \texttt{String} if the state reaches 100, 
+and runs in a \texttt{Maybe} context:
+
+\begin{Verbatim}
+addState : Eff Maybe [STATE Int, EXCEPTION String] ()
+addState = do val <- get
+              when (val == 100) (raise "State too big")
+              put (val + 1)
+\end{Verbatim}
+
+\subsection{Implementing Effects}
+
+Effects such as state and exception are described as algebraic data types,
+and run by giving \emph{handlers} for specific computation contexts. 
+Effects have a corresponding \emph{resource} (in the case of state, the
+resource is simply the current state). Executing an effectful operation may
+change the resource and return a value:
+
+\begin{Verbatim}
+Effect : Type
+Effect = (in_res : Type) -> (out_res : Type) -> 
+         (val : Type) -> Type
+\end{Verbatim}
+
+For example, the state effect is described as follows:
+
+\begin{Verbatim}
+data State : Effect where
+     Get :      State a a a
+     Put : b -> State a b ()
+\end{Verbatim}
+
+That is, \texttt{Get} returns a value of type \texttt{a} without updating
+the resource type. \texttt{Put} returns nothing, but has the effect of updating
+the resource. In order to make an effect usable, we implement a handler
+for a computation context by making an instance of the following class:
+
+\begin{Verbatim}
+class Handler (e : Effect) (m : Type -> Type) where
+     handle : res -> 
+              (eff : e res res' t) -> 
+              (k: res' -> t -> m a) -> m a
+\end{Verbatim}
+
+The \texttt{handle} function takes the input resource, an effect which may
+update that resource and execute a side-effect, and a continuation \texttt{k}
+which takes the updated resource and the return value of the effect. We use
+a continuation here primarily because there is no restriction on the number of
+times a handler may invoke the continuation (raising an exception, for example,
+will not call the continuation). Reading and updating states in handled
+for all computation contexts \texttt{m} as follows:
+
+\begin{Verbatim}
+instance Handler State m where
+     handle st Get     k = k st st
+     handle st (Put n) k = k n ()
+\end{Verbatim}
+
+Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
+the \texttt{Get} and \texttt{Put} operations into functions in \texttt{Eff}, as
+follows:
+
+\begin{Verbatim}
+STATE : Type -> EFFECT
+STATE t = MkEff t State
+
+get : Eff m [STATE x] x
+get = Get
+
+put : x -> Eff m [STATE x] ()
+put val = Put val
+\end{Verbatim}
+
+A concrete effect is simply an algebraic effect type paired with its current
+resource type. This, and other
+technical details, are explained in full elsewhere~\cite{brady:effects}.
+For the purposes of this paper, it suffices to know how to describe and
+handle new algebraic effects.
+
+\subsection{Resource Protocols as Effects}
+
+More generally, a program might modify the set of effects available.
+This might be desirable for several reasons, such as adding a new
+effect, or to update an index of a dependently typed state. In this
+case, we describe programs using the \texttt{EffM} data type:
 
@@
 EffM : (m : Type -> Type) ->
-       (es : List EFFECT) ->
-       (es’ : List EFFECT) ->
+       (es : List EFFECT) -> (es' : List EFFECT) ->
        (a : Type) -> Type
 \end{Verbatim}
-\texttt{EffM} is a type that is parameterised over an execution context
-(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
-of effects at the end of the program's execution (\texttt{es'}), and a return
-type (\texttt{a}). The list of input types can be used to ensure that an
-operation may only be called when the program is in a particular state, and
-since the list of input effects may differ from the list of output effects, we
-may modify the resource associated with the state after each operation. This
-allows us to simply enforce a resource usage protocol, and deviation from this
-protocol will result in a compile-time error.
-
-Implementing a new effect involves creating a new algebraic data type, of kind
-\texttt{Effect}, which describes the available operations and how they modify
-the resource associated with the effect. Each operation consists of an initial
-resource \texttt{res}, the resource at the end of the operation \texttt{res'},
-and the output type of the operation \texttt{t}.
-
-%TODO: Execution
-\begin{Verbatim}
-Effect : Type
-Effect = (res : Type) -> (res’ : Type) ->
-         (t : Type) -> Type
-\end{Verbatim}
-
-Operations are made available to users by writing functions with the
-corresponding \texttt{EffM} type, which may then be used to invoke the
-operation from within an effectful program.
-
-\begin{Verbatim}
-queryPostVar : 
- String -> 
- Eff m [CGI (InitialisedCGI TaskRunning)] 
-   (Maybe String)
-queryPostVar query = (QueryPostVar query)
-\end{Verbatim}
-
-Additionally, we define \textit{handlers}, which describe how each operation is
-interpreted in a given execution context. Finally, \texttt{Effect} instances
-are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
-constructor, which is provided with the initial resource value.
+
+\texttt{EffM} is parameterised over the context and type as before, but
+separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
+In fact, \texttt{Eff}
+is defined in terms of \texttt{EffM}, with equal input/output effects.
+We can use this to describe how effects follow resource protocols. A simple
+example is a file access protocol, where a file must be opened before it
+is read or written, and a file must be closed on exit:
+
+\begin{Verbatim}
+data FileIO : Effect where
+     Open  : String -> (m : Mode) -> 
+             FileIO () (OpenFile m) ()
+     Close : FileIO (OpenFile m) () ()
+
+     ReadLine  : FileIO (OpenFile Read)  
+                        (OpenFile Read) String
+     WriteLine : String -> 
+                 FileIO (OpenFile Write) 
+                        (OpenFile Write) ()
+     EOF       : FileIO (OpenFile Read)  
+                        (OpenFile Read) Bool
+\end{Verbatim}
+
+Note that the types of the input and output resources describes how resource
+state changes in each operation: opening a file changes an empty resource to
+a resource containing an open file; reading a line is only possible if the
+resoource is a file open for reading, etc.
+The handler for this effect for an \texttt{IO} computation context will
+execute the required primitive I/O actions, as well as throwing an exception
+if any operation fails.
+
+The following program type checks, and therefore implicitly carries 
+a proof that the file resource protocol is followed correctly:
+
+\begin{Verbatim}
+testFile : Eff IO [FILE_IO (), STDIO] () 
+testFile = catch (do open "testFile" Read
+                     str <- readLine
+                     close
+                     putStrLn str)
+                 (\err => putStrLn ("Error: " ++ show err))
+\end{Verbatim}
+
+The type of \texttt{testFile} states
+that File I/O and console I/O are available effects, and in particular that
+the resource associated with the File I/O will be in the same state on entry
+and exit.
+Therefore, attempting to write to the file, or failing to open or close the
+file, would cause a \emph{compile-time} error.
+
+We will use this technique extensively
+throughout this paper: describe a resource usage protocol in terms of
+state transitions; implement an effect which captures that protocol; implement
+programs which, by using this effet, implicitly carry a proof that the resource
+protocol has been correctly followed.
 
@@
 \end{Verbatim}
-Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
-Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.
-
-Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved.
-The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
-The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
-
+Figure~\ref{fig:cgistates} shows the states through which the CGI program
+progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
+as a dependent algebraic effect. Each operation performed in an effectful
+program requires the resource to be of a certain type, and the completion of
+the operation may alter the type or value of the resource.
+
+Upon creation, the CGI application is uninitialised, meaning that the
+environment variables and request have not been retrieved. The only operation
+that can be performed in this state is initialisation: by calling
+\texttt{initialise}, a CGIInfo record is populated, and the state transitions
+to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
+\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
+the initialised state.
+
+\begin{figure}[h]
 \begin{Verbatim}
@@
 \end{Verbatim}
+\caption{CGI Effect}
+\label{fig:cgieffect}
+\end{figure}
+
 Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
@@
 
-Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
-This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.
-
+Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
+access, which we have implemented for the SQLite library. This is encapsulated
+by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
+bind and retrieve data types other than \texttt{String}, in the interest of
+brevity.
+
+\begin{figure}[h]
 \begin{Verbatim}[samepage]
@@
 \end{Verbatim}
+\caption{Database Effect}
+\label{fig:dbeffect}
+\end{figure}
+
 There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
@@
 %TODO: Fix up the labels
+
+\begin{figure}[h]
 \begin{Verbatim}[samepage]
@@
 \end{Verbatim}
+\caption{Session Effect}
+\label{fig:sessioneffect}
+\end{figure}
+
 \begin{figure}[htpb!]
@@
 \end{figure}
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler.
-In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key.
-The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.
-This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.
-
-Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
+
+Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
+with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
+algebraic effect. In this application, there exist two states:
+In \texttt{SessionUninitialised}, the user may load or create a
+session.
+In \texttt{SessionInitialised}, the user may update the
+representation of the session in memory, serialise the session and write it to
+the database, or delete the session and invalidate the user's session key. The
+introduction of these two states ensures that changes are explicitly either
+written or discarded, eliminating the possibility of a developer updating the
+session but neglecting to commit it to persistent storage. This, of course, is
+under the assumption that the process exits cleanly: we attempt to facilitate
+this by writing total functions where possible.
+
+Much like the SQLite effect, we encapsulate failure by reflecting it in the
+resource associated with the effect.
+
 \begin{Verbatim}
@@
 \label{form}
-Programming web applications often involves processing user data, which may then be used in further effectful computations.
-Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.
-
-This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker.
-One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be used in a query such as:
+Programming web applications often involves processing user data, which may
+then be used in further effectful computations. Data submitted using a form is
+transmitted over the internet as a string as part of an HTTP request, which
+traditionally involves losing associated type information.
+
+This can in turn lead to risks; developers may assume that data is
+of a certain type, and therefore discount the possibility that it may have been
+modified by an attacker. One example would be the traversal of paginated data,
+in which a form is used to make a request to retrieve the next page of data.
+This may involve sending an integer detailing the current page, which could be
+used in a query such as:
 
@@
 \end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
-% Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here
-
-In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information.
-Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and echoes it back.
-Firstly, we define a form handler which echoes back a string provided by the form handler.
-It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.
+The \texttt{page} variable is assumed to be an integer, but may instead be
+modified by an attacker to include a malicious string which would alter the
+semantics of the query, allowing an attacker to execute a blind SQL injection
+attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
+clauses here
+
+In this section, we present a mechanism by which we introduce a DSL
+for the creation of web forms which preserve type information, implemented
+as a dependent algebraic effect. Once the form has
+been submitted, retrieved information is passed directly to a
+developer-specified function for handling, without the need to manually check
+and deserialise data.
+
+We begin with a simple example of a form which requests a user's name, and
+echoes it back. Firstly, we define a form handler which echoes back a string
+provided by the form handler. It has one argument of type \texttt{Maybe
+String}, which accounts for the possibility that the user may have specified
+incorrect data within the form.
 
@@
 \end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user.
-Upon submission of the form, the form handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
+When this CGI application is invoked, it will begin by outputting a form to the
+page, requesting a name from the user. Upon submission of the form, the form
+handler will be invoked, and the name will be used in the output.
+
+In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
+of the form-handling system: namely, the effect which allows the creation of
+forms, and the handling code which deserialises the data and passes it to the
+user-specified handler function.
+
 \subsection{Form Construction}
@@
 %
-In order to specify a form, we once again use \texttt{Effects}.
-By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation.
-We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.
+In order to specify a form, we once again use \texttt{Effects}. By recording
+the type of each form element as it is added in the type of the form, we may
+statically ensure that the user-supplied handler function is of the correct
+type to handle the data supplied by the form: the specification of an
+incompatible handler will result in a compile-time type error.
+
+\idris{} allows for implicit arguments to be bound across a block of code
+through \texttt{using} notation. We may therefore parameterise the 
+form data types over
+the types associated with each form element, and the effects required by the
+handler function.
 
@@
 \end{Verbatim}
-The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
-
-We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form.
-The resource allows us to record the types associated with each form element and the HTML required to display the form.
-he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
+The implementation of the form effect also contains other constructs for
+additional form elements such as radio buttons and check boxes, but are omitted
+in the interest of brevity.
+
+We make use of the resource associated with the effect, \texttt{FormRes}, to
+construct the form. The resource allows us to record the types associated with
+each form element and the HTML required to display the form. he constructor,
+\texttt{FR}, requires the number of elements in the form in order to allow for
+the naming of new elements, the list of element types, the list of effects
+supported by the handler function, and the currently generated HTML for the
+form.
 
@@
 \end{Verbatim}
-All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
-As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.
+All forms are required to include a submit button, as mandated by the
+requirement that the input and output resource contains an empty list of types;
+this requirement is fulfilled as per the output resource type of the
+\texttt{AddSubmit} operation. As the creation of a form is a pure function
+which does not include side effects, we do not restrict the handler to IO as
+with previously-discussed effects, instead denoting the fact that it may be run
+in any handler with the implicit variable \texttt{m}.
 
@@
 
-Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a hidden HTML field.
+Running form construction is achieved as an operation of the CGI
+effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
+The generated metadata describing the handler function is serialised as a
+hidden HTML field.
 
@@
 \section{Related Work}
-\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser.
-The library also provides support for cookies and basic form handling.
-
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods.
-In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs.
-Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs.
-Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations.
-Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.
-
-\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction.
-Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven.
-This means that upon a component being edited, a computation occurs, given the state of the current form.
-This is then saved to be used in future computations.
-Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted.
-This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.
-
-The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another.
-This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion.
-These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application.
-Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.
-
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs.
-Through the use of the Ur\/Web DSL, developers may generate well-typed code from records.
-By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure.
-Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.
-
-WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres.
-WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes.
-The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables.
-These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy.
-Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing.
-WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation.
-We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.
+\citet{meijer:cgi} implemented a CGI library which was among the first
+libraries to handle web scripting monadically, and allows the user to implement
+application logic without having to consider the low-level details such as
+parsing in CGI data from the environment, or printing headers to the remote
+browser. The library also provides support for cookies and basic form handling.
+
+\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining
+state, and provides more sophisticated form-handling methods. In particular,
+callbacks may be associated with submit buttons, with nameless representations
+for form inputs. Due to the unavailability of full dependent types in Haskell,
+however, this implementation does not statically verify the suitability of the
+callback function for the form inputs. Both implementations of the CGI library,
+being built upon monads, mean that the use of additional effects such as
+database access is achieved either through monad transformers or the ability to
+perform arbitrary IO operations. Both of these approaches are limited---the
+former does not scale well to multiple effects, and the latter allows for the
+introduction of errors by allowing the violation of resource usage protocols.
+
+\citet{plasmeijer:idata} describe an alternative approach to type-safe form
+handling which does not require the use of dependent types through the
+\textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of
+processing being triggered by form submission, as in the approach described in
+this paper, applications created in the \idata{} toolkit are instead
+edit-driven. This means that upon a component being edited, a computation
+occurs, given the state of the current form. This is then saved to be used in
+future computations. Should a user enter invalid data, for example by entering
+text in a field designated for integers, the change will be reverted. This is
+demonstrated practically through the use of \idata{} to implement a conference
+management system \cite{plasmeijer:cms}.
+
+The concept of \idata{} is taken further by the introduction of \itasks{}
+\cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple
+\idata{} forms to interact with one another. This is achieved through the use
+of high-level combinators which allow the implementation of concepts such as
+recurseion, sequence and choice in a scalable fashion. These high-level
+abstractions are elegant, but the style and syntax differ substantially from a
+traditional web application. Our approach takes the concept of type-safe input
+handling and uses it in a more traditional fashion, whilst retaining the
+type-retention guarantees afforded by \idata{} elements.
+
+Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
+dependent types to enable the generation of provably-correct and secure queries
+and DOM code, without requiring overly complex types or proofs. Through the use
+of the Ur\/Web DSL, developers may generate well-typed code from records. By
+statically verifying the \textit{code generator}, it is therefore possible to
+assert that the generated code is secure. Such ideas are extremely interesting
+for the implementation of object-relational mapping, which we hope to implement
+in the near future.
+
+WebDSL \cite{webdsl} is a domain-specific language written primarily to
+introduce new abstractions which aim to reduce the amount of boilerplate code
+that must be written and maintained by developeres. WebDSL is built on top of
+Java, which often includes a large amount of redundant code such as accessor
+and mutator functions within entity classes. The Java Persistence API (JPA)
+\cite{jpa} provides an object-relational mapping through the use of Java 5
+annotations, which may then be used to construct database tables. These
+annotations soon become complex, however, and coupled with redundant
+boilerplate code, data model declarations may soon become unwieldy. Through the
+use of WebDSL, these data model declarations can be much more compactly
+declared, and elaborated into Java code by parsing the data-modelling DSL into
+an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL
+also applies similar concepts to implement a \textit{template system} for such
+objects, which allows the data to be used in code generation. We look to
+implement many of these ideas, but as effects within the IdrisWeb
+framework, as with the form construction effect.
 

### 2013-08-21 paper/main.tex
Paper revisions

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows,automata}
 \usepackage{url}
@@
 \usepackage[numbers]{natbib}
-\usetikzlibrary{arrows,automata}
 %\WarningsOff
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
+\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
 \copyrightyear{2013} 
@@
 
-\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
+\title{Practical Dependently-Typed Programming for the Web}
+\subtitle{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
 %\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
@@
 \begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within
-programs in turn allowing easier reasoning about the operation of such
-programs. Stronger specifications do however bring with them the disadvantage
+Dependently-typed languages allow very expressive types to be used during development, in turn facilitating easier reasoning about the operation of programs written in such languages.
+Stronger type specifications do however bring with them the disadvantage
 that it becomes increasingly difficult to write programs that are accepted by
@@
 In this paper, we apply this technique to web programming, and introduce an
-EDSL, to facilitate the creation and handling of web forms which retain type
+EDSL to facilitate the creation and handling of web forms which retain type
 information, reducing the scope for programmer error and attacks such as SQL
@@
 After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. At
-this point, a developer may forget to execute the statement before beginning to
-fetch results or may have forgotten to bind the parameters to the statement:
+variables into which results will be stored, and fetch each row in turn. 
+Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.
 
@@
 complexity of the underlying type system. EDSLs allow domain experts to
-focus on a problem domain, rather than the specific details of their
-solution.
+write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.
 
@@
 The primary contribution of this paper is the application of 
-dependent types may to provide strong static guarantees
+dependent types to provide strong static guarantees
 about the correctness and security of web applications, whilst minimising
@@
 manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection.
+increasing applications' resilience to attacks such as SQL injection and cross-site scripting.
 
 \item Representations of CGI, Databases and sessions as
-\textit{algebraic effects}, allowing programs to be accepted only when they
+\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
 follow clearly defined resource usage protocols.
 
-\item A practical demonstration of the framework, a message board application.
+\item A message board application, demonstrating the usage of the framework.
 
@@
 state transitions; implement an effect which captures that protocol; implement
-programs which, by using this effet, implicitly carry a proof that the resource
+programs which, by using this effect, implicitly carry a proof that the resource
 protocol has been correctly followed.
@@
 CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.
+Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error being shown.
 
@@
 progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
-as a dependent algebraic effect. Each operation performed in an effectful
+as a resource-dependent algebraic effect. Each operation performed in an effectful
 program requires the resource to be of a certain type, and the completion of
@@
 
-Upon creation, the CGI application is uninitialised, meaning that the
-environment variables and request have not been retrieved. The only operation
+Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
 that can be performed in this state is initialisation: by calling
@@
 
-After the user code has finished execution, control returns to the library code.
-At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
+After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, and the headers are written.
 Finally, the headers and content are written which completes the process.
@@
 bind and retrieve data types other than \texttt{String}, in the interest of
-brevity.
+brevity).
 
@@
 attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
-clauses here
+% clauses here
 
@@
 \label{formcons}
-Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
+Each form element is specified  to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function.
 In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
@@
 construct the form. The resource allows us to record the types associated with
-each form element and the HTML required to display the form. he constructor,
+each form element and the HTML required to display the form. The constructor,
 \texttt{FR}, requires the number of elements in the form in order to allow for
@@
 this requirement is fulfilled as per the output resource type of the
-\texttt{AddSubmit} operation. As the creation of a form is a pure function
+\texttt{AddSubmit} operation. As the creation of a form is a return ()function
 which does not include side effects, we do not restrict the handler to IO as
@@
 Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
+This is, however, not an inherent limitation as the \texttt{Effects} library permits introduction of additional effects within an effectful computation.
 %
@@
            handleRequest)
-          pure ()
+          return ()
 \end{Verbatim}
@@
     lift' (handleForm handlers)
-    Effects.pure ()
+    Effects.return ()
   else do
@@
   withSession (addNewThread title content) notLoggedIn
-  pure ()
+  return ()
 handleNewThread _ _ = do 
@@
        an error posting your thread."
-  pure ()
+  return ()
 \end{Verbatim}
@@
   case insert_res of
-    Left err => pure False
+    Left err => return False
     Right thread_id => postInsert uid thread_id content
@@
   Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = pure ()
+traverseThreads [] = return ()
 traverseThreads (x::xs) = do traverseRow x
@@
             username ++ "</td></tr>") 
-        traverseRow _ = pure ()
+        traverseRow _ = return ()
 \end{Verbatim}
@@
         lift' (output $ "Welcome, " ++ name)
-        pure ()
+        return ()
       else do
         lift' (output "Could not set session")
-        pure ()
+         return ()
     Right Nothing => do
       lift' (output "Invalid username or password")
-      pure ()
+      return ()
     Left err => do
       lift' (output $ "Error: " ++ err)
-      pure ()
+      return ()
 \end{Verbatim}
@@
 
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
-dependent types to enable the generation of provably-correct and secure queries
-and DOM code, without requiring overly complex types or proofs. Through the use
-of the Ur\/Web DSL, developers may generate well-typed code from records. By
-statically verifying the \textit{code generator}, it is therefore possible to
-assert that the generated code is secure. Such ideas are extremely interesting
-for the implementation of object-relational mapping, which we hope to implement
-in the near future.
+Ur\/Web \cite{urweb} is a library built for the Ur language, which does not use dependent types but does have a more expressive type system, making extensive use of record types and type-level computation.
+By using these concepts, Ur\/Web may generate provably correct and unexploitable DOM code and SQL queries from records, without requiring developers to supply proofs. 
+In contrast to using runtime code generation, which is prone to obscure code generation errors, Ur\/Web makes use of its static type system to guarantee that metaprograms---in this case, generated SQL and DOM code---must be correct and secure.
+Such ideas regarding the use of static checking of metaprogram generation will be extremely useful when considering an object-relational mapping system, which we hope to implement in the near future.
+It will also be interesting to see how such concepts may be applied with a yet more expressive type system involving full dependent types.
+%Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
+%dependent types to enable the generation of provably-correct and secure queries
+%and DOM code, without requiring overly complex types or proofs. Through the use
+%of the Ur\/Web DSL, developers may generate well-typed code from records. By
+%statically verifying the \textit{code generator}, it is therefore possible to
+%assert that the generated code is secure. Such ideas are extremely interesting
+%for the implementation of object-relational mapping, which we hope to implement
+%in the near future.
 
@@
 With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
+Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error or memory access violation.
 
@@
 
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
+Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used during compilation.
 Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
+If data structures within the program do not conform to a given database schema, for example, then the program will not type-check and as such will produce a compile-time error.
 This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.

### 2013-11-09 Common/Random/rand_c.c
Tried transitioning to the new build system, but having trouble with effects

@@ -1,22 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <time.h>
-#include "rand_c.h"
-
-
-/* Massive cop-out so I don't have to mess around doing this in Idris! */
-// Taken from a StackOverflow post by Abhay Budakoti
-int random_number(int min_num, int max_num)
-{
-    int result=0,low_num=0,hi_num=0;
-    if(min_num<max_num) {
-        low_num=min_num;
-        hi_num=max_num+1; // this is done to include max_num in output.
-    } else {
-        low_num=max_num+1;// this is done to include max_num in output.
-        hi_num=min_num;
-    }
-    srand(time(NULL));
-    result = (rand()%(hi_num-low_num))+low_num;
-    return result;
-}

### 2013-11-09 Common/Random/rand_c.h

@@ -1,4 +1,0 @@
-#ifndef RANDOM_C
-#define RANDOM_C
-int random_number(int min_num, int max_num);
-#endif

### 2013-11-09 GCrypt/gcrypt_idr.c

@@ -1,136 +1,0 @@
-#include <gcrypt.h>
-#include <gcrypt-module.h>
-#include <gpg-error.h>
-#include <stdbool.h>
-#include <stdio.h>
-#include "gcrypt_idr.h"
-
-// LibGCrypt bindings for Idris. Provides a layer between the
-// Idris code and the GCrypt code, to provide convenience functions
-// and datatype marshalling.
-
-
-
-// Initialises the libgcrypt library.
-// A required version may be specified: if this is
-// the empty string, then GCRYPT_VERSION will be used instead.
-const char* idris_gcry_init(const char* version) {
-    const char* ret;
-    if (strlen(version) >= 0) {
-        ret = gcry_check_version(version);
-    } else {
-        ret = gcry_check_version(GCRYPT_VERSION);
-    }
-
-    if (ret) {
-        // We're ignoring secure memory for the moment,
-        // although it may be good to add in future.
-        gcry_control (GCRYCTL_DISABLE_SECMEM, 0);
-        gcry_control (GCRYCTL_INITIALIZATION_FINISHED, 0);
-    }
-    return ret;
-}
-
-// Returns the gcry_error_t value from a call
-gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct) {
-    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
-    return struct_ptr->err;
-}
-
-// Returns the data associated with a call
-void* idris_gcry_get_struct_data(void* gcry_res_struct) {
-    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
-    return struct_ptr->res;
-}
-
-// Frees the result of an idris_grcy_res struct
-void idris_gcry_dispose_res(void* gcry_res_struct) {
-    free(gcry_res_struct);
-}
-
-// Initialises a result structure, setting the initial value of the data ptr to NULL
-idris_gcry_res* init_res_struct() {
-    idris_gcry_res* ret = malloc(sizeof(idris_gcry_res));
-    ret->res = NULL;
-    return ret;
-}
-
-// Opens a hashing context
-// TODO: flags should be more idris-friendly... Maybe an array?
-void* idris_gcry_init_md(int algorithm_key, unsigned int flags) {
-    idris_gcry_res* result_struct = init_res_struct();
-    // Allocate the context memory, perform the call
-    gcry_md_hd_t* context = malloc(sizeof(gcry_md_hd_t));
[…]

### 2013-11-09 GCrypt/gcrypt_idr.h

@@ -1,24 +1,0 @@
-#ifndef IDRIS_GCRYPT_H
-#define IDRIS_GCRYPT_H
-
-typedef struct idris_gcry_res {
-    gcry_error_t err;
-    void* res;
-} idris_gcry_res;
-
-const char* idris_gcry_init(const char* version);
-gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct);
-void* idris_gcry_get_struct_data(void* gcry_res_struct);
-void idris_gcry_dispose(void* gcry_res_struct);
-idris_gcry_res* init_res_struct();
-void* idris_gcry_init_md(int algorithm_key, unsigned int flags);
-gcry_error_t idris_gcry_enable_algorithm(void* context, int algorithm_key);
-void idris_gcry_dispose_md(void* context);
-void idris_gcry_reset_md(void* context);
-char* idris_hash_string(void* context, char* str, int algorithm_key);
-void idris_gcry_dispose_res(void* gcry_res_struct);
-
-
-
-
-#endif

### 2013-11-09 src/IdrisWeb/GCrypt/gcrypt_idr.c
Might help if I added the files

@@ -1,0 +1,136 @@
+#include <gcrypt.h>
+#include <gcrypt-module.h>
+#include <gpg-error.h>
+#include <stdbool.h>
+#include <stdio.h>
+#include "gcrypt_idr.h"
+
+// LibGCrypt bindings for Idris. Provides a layer between the
+// Idris code and the GCrypt code, to provide convenience functions
+// and datatype marshalling.
+
+
+
+// Initialises the libgcrypt library.
+// A required version may be specified: if this is
+// the empty string, then GCRYPT_VERSION will be used instead.
+const char* idris_gcry_init(const char* version) {
+    const char* ret;
+    if (strlen(version) >= 0) {
+        ret = gcry_check_version(version);
+    } else {
+        ret = gcry_check_version(GCRYPT_VERSION);
+    }
+
+    if (ret) {
+        // We're ignoring secure memory for the moment,
+        // although it may be good to add in future.
+        gcry_control (GCRYCTL_DISABLE_SECMEM, 0);
+        gcry_control (GCRYCTL_INITIALIZATION_FINISHED, 0);
+    }
+    return ret;
+}
+
+// Returns the gcry_error_t value from a call
+gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct) {
+    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
+    return struct_ptr->err;
+}
+
+// Returns the data associated with a call
+void* idris_gcry_get_struct_data(void* gcry_res_struct) {
+    idris_gcry_res* struct_ptr = (idris_gcry_res*) gcry_res_struct;
+    return struct_ptr->res;
+}
+
+// Frees the result of an idris_grcy_res struct
+void idris_gcry_dispose_res(void* gcry_res_struct) {
+    free(gcry_res_struct);
+}
+
+// Initialises a result structure, setting the initial value of the data ptr to NULL
+idris_gcry_res* init_res_struct() {
+    idris_gcry_res* ret = malloc(sizeof(idris_gcry_res));
+    ret->res = NULL;
+    return ret;
+}
+
+// Opens a hashing context
+// TODO: flags should be more idris-friendly... Maybe an array?
+void* idris_gcry_init_md(int algorithm_key, unsigned int flags) {
+    idris_gcry_res* result_struct = init_res_struct();
+    // Allocate the context memory, perform the call
+    gcry_md_hd_t* context = malloc(sizeof(gcry_md_hd_t));
[…]

### 2013-11-09 src/IdrisWeb/GCrypt/gcrypt_idr.h

@@ -1,0 +1,24 @@
+#ifndef IDRIS_GCRYPT_H
+#define IDRIS_GCRYPT_H
+
+typedef struct idris_gcry_res {
+    gcry_error_t err;
+    void* res;
+} idris_gcry_res;
+
+const char* idris_gcry_init(const char* version);
+gcry_error_t idris_gcry_get_struct_err(void* gcry_res_struct);
+void* idris_gcry_get_struct_data(void* gcry_res_struct);
+void idris_gcry_dispose(void* gcry_res_struct);
+idris_gcry_res* init_res_struct();
+void* idris_gcry_init_md(int algorithm_key, unsigned int flags);
+gcry_error_t idris_gcry_enable_algorithm(void* context, int algorithm_key);
+void idris_gcry_dispose_md(void* context);
+void idris_gcry_reset_md(void* context);
+char* idris_hash_string(void* context, char* str, int algorithm_key);
+void idris_gcry_dispose_res(void* gcry_res_struct);
+
+
+
+
+#endif

### 2013-11-09 src/rand_c.h

@@ -1,0 +1,4 @@
+#ifndef RANDOM_C
+#define RANDOM_C
+int random_number(int min_num, int max_num);
+#endif

### 2013-11-09 src/sqlite3api.c

@@ -1,0 +1,543 @@
+/*
+ *  sqlite3api.c
+ *  
+ *
+ *  Created by Melissa Farinaz MOZIFIAN on 22/06/2012.
+ *  Copyright 2012. All rights reserved.
+ *
+ */
+
+
+#include "sqlite3api.h"
+
+
+// buffer used for storing queries
+// Used in high-level functions. 
+static char sql_query_buffer[2000];
+// array used to store the string
+//value obtain by get_val_text
+unsigned char* array;
+/*
+ open an SQLite database file as 
+ specified by the filename argument. 
+ Returns pointer to DB struct on success
+ Null on failure.
+ */
+void* sqlite3_open_idr(const char *filename){
+	
+	sqlite3 *db;
+	int res =sqlite3_open(filename,&db);
+	
+	if(res != SQLITE_OK){
+		printf("Error occured while opening the databse.");
+		return NULL;
+	}
+	DBinfo *dbi = malloc(sizeof(DBinfo));
+	dbi->db_ptr = db;
+	return dbi;
+	
+}
+/*
+ Frees the resource and returns 0 on success
+ */
+int sqlite3_close_idr(void* db){
+	
+	DBinfo* dbi =(DBinfo*) db;
+	int res =sqlite3_close(dbi->db_ptr);
+	if (res == SQLITE_OK){
+		free(dbi);
+		return 0;
+	}
+	else {
+		return res;
+	}	
+}
+/*
+ SQLite wrapper around  sqlite3_prepare_v2(), 
+ sqlite3_step(), and sqlite3_finalize().
+ This version of exec cannot be used with prepare.
+ For executing queries, must use the exec_db function.
+ */
+int sqlite3_exec_idr(void* db, const char *sql)
+{
+	DBinfo* dbi =(DBinfo*) db;
+	char* err;
+	int rc;
+	rc = sqlite3_exec(dbi->db_ptr,sql,NULL, NULL, &err);
+	if (rc != SQLITE_OK && err != NULL) {
+		strncpy(dbi->buffer, err, sizeof(dbi->buffer));
+		sqlite3_free(err);
+	}
+	return rc;
+}
+/*
+ Gets the error store in the buffer
+ in the struct. Some certain functions 
+ have the feature to store errors in a buffer.
+ */
+char* sqlite3_get_error(void* db) {
+	DBinfo* dbi =(DBinfo*) db;
+	return dbi->buffer;
[…]

### 2013-11-11 paper/effects.tex
Added paper revisions so far Also started the rewrite of the DB section with the new DB stuff

--- 2013-08-14 paper/main.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-%\usepackage{silence}
-\usepackage{pgf}
-\usepackage{tikz}
-\usepackage{url}
-\usepackage[hidelinks]{hyperref}
-\usepackage{fancyvrb}
-\usepackage[numbers]{natbib}
-\usetikzlibrary{arrows,automata}
-%\WarningsOff
-
-\begin{document}
-\fvset{fontsize=\small}
-\newcommand{\idris}{\textsc{Idris}}
-\newcommand{\idata}{\textsf{iData}}
-\newcommand{\itasks}{\textsf{iTasks}}
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{2013} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{}        % These are ignored unless
-\preprintfooter{}   % 'preprint' option specified.
-
-\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
-%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
-
-\authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
-           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
-
-\maketitle
-
-\begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within
-programs in turn allowing easier reasoning about the operation of such
-programs. Stronger specifications do however bring with them the disadvantage
-that it becomes increasingly difficult to write programs that are accepted by
-the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in high-level languages which use dependent
-types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an
-EDSL, to facilitate the creation and handling of web forms which retain type
-information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how to enforce resource usage
-protocols associated with common web operations such as CGI, database access
-and session handling.
-
-\end{abstract}
-
-%\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-%\keywords
-%keyword1, keyword2
-
-\section{Introduction}
-
-Web applications, whilst ubiquitous, are also prone to incorrect construction
-and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
-scripting \cite{owasp:xss}. Security breaches using such exploits are
-far-reaching, and high profile cases involve large corporations such as Sony,
-who suffered a well-publicised and extremely costly SQL Injection breach in
-2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
-\cite{imperva:yahoo}.
-
-Many web applications are written in dynamically-checked scripting languages
-such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. A significant drawback, however, is that such languages
-do not provide  the same static guarantees about runtime behaviour afforded by
-programs with more expressive, static type systems, instead relying on
-extensive unit testing to ensure correctness and security.
-
-%%------- EXAMPLE
-Let us consider a simple database access routine, written in
-PHP, where we wish to obtain the name and address of every employee working in
-a given department, \texttt{\$dept}. We firstly construct an object
-representing a database connection, where the arguments are the database host,
-user, password and name respectively:
-
-%
-\begin{Verbatim}
-$conn = new mysqli("localhost", "username", 
-                    "password", "db");
-\end{Verbatim}
-%
-We then check to see if the connection was successful, and exit
-if not.  This check is optional, so
-it would be possible to omit it. However, this would cause
-problems with later steps.
-
-%
-\begin{Verbatim}
-  if (mysqli_connect_errno()) {
-      exit();
-  }
-\end{Verbatim}
-%
-We then create a prepared statement detailing our query, and bind the
-\texttt{`dept'} value:
-
-\begin{Verbatim}
-  $stmt = $conn->prepare("SELECT `name`, `address` 
-     FROM `staff` WHERE `dept` = ?);
-  $stmt->bind_param('s', $dept);
-\end{Verbatim}
-%
-After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. At
-this point, a developer may forget to execute the statement before beginning to
-fetch results or may have forgotten to bind the parameters to the statement:
-
-%
-\begin{Verbatim}
-  $stmt->execute();
-  $stmt->bind_result($name, $address);
-  while ($stmt->fetch()) {
-    printf("Name: %s, Age: %s", $name, $age);
-  }
-\end{Verbatim}
-%
-Finally, once the statement and connection are no longer needed, they should be
-closed in order to discard the associated resources:
-
-%
-\begin{Verbatim}
-  $stmt->close();
-  $conn->close();
-\end{Verbatim}
-%
-Even in this small example, there exists a specific resource usage protocol.
-Firstly, a connection to the database must be opened. The object-oriented style
-used in the example encapsulates this to an extent, as the object must be
-created in order for operations to be performed, however it is less obvious in
-a procedural version of the code. Secondly, a prepared statement is created,
-using the raw SQL and placeholders to which variables are later bound. The
-statement is then executed, and each row is retrieved from the database.
-Finally, the resources are freed.
-
-Problems may arise if the protocol is not followed correctly.
-A developer may, for example, accidentally close a statement whilst still
-retrieving rows, which would cause a runtime error. Similarly, a programmer may
-omit closing the statement or connection, which can lead to
-problems such as resource leaks in longer-running server applications.
-However, in conventional programming languages, there is no way to check
-automatically that a protocol is followed.
-
-In contrast, the use of \textit{dependent types} makes it possible
-to specify a program's behaviour precisely, and to check that a 
-specification is followed.
-%
-The difficulty is 
-that automatic verification by a compiler can be difficult or
-often impossible, requiring additional proofs to be given by the programmer.
-
-This complexity can be ameliorated through the use of \textit{embedded
-domain-specific languages} (EDSLs), which aim to abstract away the
-complexity of the underlying type system. EDSLs allow domain experts to
-focus on a problem domain, rather than the specific details of their
-solution.
-
-\idris{} \cite{brady2011idris} is a language with full dependent types, and
-extensive support for EDSLs through overloading and syntax macros. Through the
-use of \idris{}, and a framework for describing resource protocols using
-\emph{algebraic effects}~\cite{brady:effects}, we
-present a dependently-typed web framework, which allows the construction of
-programs with additional guarantees about correctness and security, whilst
-minimising the increase in development complexity.
-
-\subsection{Contributions}
-The primary contribution of this paper is the application of 
-dependent types may to provide strong static guarantees
-about the correctness and security of web applications, whilst minimising
-additional development complexity. In particular, we present:
-
-\begin{itemize}
-\item A form-handling mechanism, which preserves type information and
-manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection.
-
-\item Representations of CGI, Databases and sessions as
-\textit{algebraic effects}, allowing programs to be accepted only when they
-follow clearly defined resource usage protocols.
-
-\item A practical demonstration of the framework, a message board application.
-
-\end{itemize}
-
-We structure the remainder of this paper as follows. We provide a brief
-overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
-how this may be used to ensure adherence to resource usage protocols for CGI,
-SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
-type-safe form handling in Section ~\ref{form}, implemented using
-\texttt{Effects}; and discuss the larger example
-of a message board system making use of these components in Section
-~\ref{messageboard}.
-
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
-% =================================================
-% =================================================
-
 \section{An overview of the \texttt{Effects} framework}
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{efftype}
+
 Eff : (m : Type -> Type) ->
       (es : List EFFECT) -> (a : Type) -> Type
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{efftype}
+
+\noindent
 \texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
@@
 list of side effects \texttt{es} that the program is permitted to use, and the
-program’s return type \texttt{a}. The name \texttt{m} for the computation context is
+programs return type \texttt{a}. The name \texttt{m} for the computation context is
 suggestive of a monad, but there is no requirement for it to be so.
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{addState}
+
 addState : Eff Maybe [STATE Int, EXCEPTION String] ()
@@
               put (val + 1)
-\end{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{addState}
 
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{efffam}
+
 Effect : Type
@@
          (val : Type) -> Type
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{efffam}
+
+\noindent
 For example, the state effect is described as follows:
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{stateeff}
+
 data State : Effect where
@@
      Put : b -> State a b ()
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{stateeff}
+
+\noindent
 That is, \texttt{Get} returns a value of type \texttt{a} without updating
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{effhandler}
+
 class Handler (e : Effect) (m : Type -> Type) where
@@
               (k: res' -> t -> m a) -> m a
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{effhandler}
+
+\noindent
 The \texttt{handle} function takes the input resource, an effect which may
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{statehandler}
+
 instance Handler State m where
@@
      handle st (Put n) k = k n ()
-\end{Verbatim}
 
+\end{SaveVerbatim}
+
+\noindent
 Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{stateconc}
+
 STATE : Type -> EFFECT
@@
 put val = Put val
-\end{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{stateconc}
 
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{effmtype}
+
 EffM : (m : Type -> Type) ->
@@
        (a : Type) -> Type
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{effmtype}
+
+\noindent
 \texttt{EffM} is parameterised over the context and type as before, but
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{fileeff}
+
 data FileIO : Effect where
@@
                         (OpenFile Read) Bool
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{fileeff}
+
+\noindent
 Note that the types of the input and output resources describes how resource
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{testfile}
+
 testFile : Eff IO [FILE_IO (), STDIO] () 
@@
             (\err => putStrLn ("Error: " ++ show err))
-\end{Verbatim}
 
+\end{SaveVerbatim}
+\useverb{testfile}
+
+\noindent
 The type of \texttt{testFile} states
@@
 % =================================================
-
-\section{Modelling resource usage protocols}
-\label{rup}
-In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
-
-% =================================================
-
-\subsection{CGI}
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {Initialised};
-  \node[state]         (C) [below of=B] {TaskRunning};
-  \node[state]         (D) [below of=C] {TaskCompleted};
-  \node[state]         (E) [below of=D] {HeadersWritten};
-  \node[state]         (F) [below of=E] {ContentWritten};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
-\end{tikzpicture}
-}
-\caption{CGI States}
-\label{fig:cgistates}
-\end{figure}
-
-CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.
-
-A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
-Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
-If this were to happen, then the program would fail due to the fact that headers had not been written to the client.
-
-By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
-In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
-This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI process as the unit type, ().
-\begin{Verbatim}[samepage]
-data CGIStep = Initialised 
-             | TaskRunning 
-             | TaskCompleted 
-             | HeadersWritten 
-             | ContentWritten
-
-data InitialisedCGI : CGIStep -> Type where
-  ICgi : CGIInfo -> InitialisedCGI s
-\end{Verbatim}
-Figure~\ref{fig:cgistates} shows the states through which the CGI program
-progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
-as a dependent algebraic effect. Each operation performed in an effectful
-program requires the resource to be of a certain type, and the completion of
-the operation may alter the type or value of the resource.
-
-Upon creation, the CGI application is uninitialised, meaning that the
-environment variables and request have not been retrieved. The only operation
-that can be performed in this state is initialisation: by calling
-\texttt{initialise}, a CGIInfo record is populated, and the state transitions
-to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
-\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
-the initialised state.
-
-\begin{figure}[h]
-\begin{Verbatim}
-data Cgi : Effect where
-    Init : Cgi () (InitialisedCGI Initialised) ()
-    StartRun : Cgi (InitialisedCGI Initialised) 
-                   (InitialisedCGI TaskRunning) ()
-    FinishRun : Cgi (InitialisedCGI TaskRunning) 
-                    (InitialisedCGI TaskCompleted) ()
-    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
-                       (InitialisedCGI HeadersWritten) ()
-    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
-                       (InitialisedCGI ContentWritten) ()
-    OutputData : String -> 
-                 Cgi (InitialisedCGI TaskRunning) 
-                     (InitialisedCGI TaskRunning) ()
-    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
-                               :: effs) -> 
-                       CGIProg effs a -> 
-                       Cgi (InitialisedCGI TaskRunning) 
-                           (InitialisedCGI TaskRunning) a
-\end{Verbatim}
-\caption{CGI Effect}
-\label{fig:cgieffect}
-\end{figure}
-
-Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
-
-User code executes in the \texttt{TaskRunning} state.
-Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
-This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
-
-After the user code has finished execution, control returns to the library code.
-At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
-Finally, the headers and content are written which completes the process.
-Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
-
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-\begin{Verbatim}
-CGIProg : List EFFECT -> Type -> Type
-CGIProg effs a = 
- Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{Verbatim}
-This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
-
-A simple ``Hello, world!'' program would be defined as follows:
-
-\begin{Verbatim}
-module Main
-import Cgi
-
-sayHello : CGIProg [] ()
-sayHello = output "Hello, world!"
-
-main : IO ()
-main = runCGI [initCGIState] sayHello
-\end{Verbatim}
-% =================================================
-
-\subsection{Database access with SQLite}
-
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {ConnectionOpened};
-  \node[state]         (C) [below of=B] {PreparedStatementOpen};
-  \node[state]         (D) [below of=C] {PreparedStatementBinding};
-  \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExecuting};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
-        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
-\end{tikzpicture}
-}
-\caption{Database Resource Usage Protocol}
-\label{fig:sqlitestates}
-\end{figure}
-
-SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
-We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-
-The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
-Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
-Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user.
-Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.
-
-Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
-access, which we have implemented for the SQLite library. This is encapsulated
-by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
-bind and retrieve data types other than \texttt{String}, in the interest of
-brevity.
-
-\begin{figure}[h]
-\begin{Verbatim}[samepage]
-data Sqlite : Effect where
-  OpenDB : 
-    String -> 
-    Sqlite () 
-           (SQLiteRes ConnectionOpened) Bool
-  CloseDB : 
-    Sqlite (SQLiteRes ConnectionOpened) () Bool
-  PrepareStatement : 
-    String -> 
-    Sqlite (SQLiteRes ConnectionOpened) 
-           (SQLiteRes PreparedStatementOpen) Bool
-  StartBind : 
-    Sqlite (SQLiteRes PreparedStatementOpen) 
-           (SQLiteRes PreparedStatementBinding) ()
-  BindText : 
-    ArgPos -> String -> Int -> 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBinding) 
-           Bool
-  FinishBind : 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBound) 
-           Bool
-  ExecuteStatement : 
-    Sqlite (SQLiteRes PreparedStatementBound) 
-           (SQLiteRes PreparedStatementExecuting) ()
-  Finalise : 
-    Sqlite (SQLiteRes PreparedStatementExecuting) 
-           (SQLiteRes ConnectionOpened) 
-           Bool
-  GetColumnText : 
-    Int -> 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           String
-  RowStep : 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           StepResult
-\end{Verbatim}
-\caption{Database Effect}
-\label{fig:dbeffect}
-\end{figure}
-
-There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
-\begin{Verbatim}
-data SQLiteRes : Step -> Type where
-  OpenConn : DBPointer -> SQLiteRes s
-  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
-  ExecutingStmt : DBPointer -> 
-                  StmtPointer -> 
-                  StepResult -> 
-                  SQLiteRes s
-
-data DBPointer = ValidConn Ptr
-               | InvalidConn
-
-data StmtPointer = ValidStmt Ptr
-                 | InvalidStmt 
-\end{Verbatim}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
-At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
-
-%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
-In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
-Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function.
-We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-\begin{Verbatim}
-data StepResult = Unstarted
-                | StepFail
-                | StepComplete
-                | NoMoreRows
-\end{Verbatim}
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
-Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
-We may therefore statically guarantee that only calls that will return a valid result are executed.
-
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
-
-\subsubsection{Example Code}
-Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
-Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
-
-We define a function of type:
-\begin{Verbatim}
-String ->
-Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{Verbatim}
-This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-\begin{Verbatim}
-testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, Int)))
-testSelect dept = do
-  open_db <- openDB "people.db"
-  if open_db then do
-    let sql = "SELECT name, address FROM `staff` 
-                    WHERE dept = ?;"
-    sql_prep_res <- prepareStatement sql
-    if sql_prep_res then do 
-      startBind
-      bindText 1 dept
-      finishBind
-      beginExecution
-      results <- collectResults
-      finaliseStatement
-      closeDB
-      return $ Right results
-    else do err <- stmtFail
-            return $ Left err
-  else do err <- connFail
-          return $ Left err 
-\end{Verbatim}
-The program initially attempts to open a connection to the \texttt{people.db} database.
-At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
-The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
-
-A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.
-
-\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database.
-This function is then recursively called until there are no more rows to process.
-\begin{Verbatim}
-collectResults : 
-  Eff IO 
-  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-  (List (String, Int))
-collectResults = do
-  step_result <- nextRow
-  case step_result of
-      StepComplete => do name <- getColumnText 1
-                         age <- getColumnInt 2
-                         xs <- collectResults
-                         return $ (name, age) :: xs
-      NoMoreRows => return []
-      StepFail => return [] 
-\end{Verbatim}
-In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
-In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-\begin{Verbatim}
-data DBVal = DBInt Int
-           | DBText String
-           | DBFloat Float
-           | DBNull
-\end{Verbatim}
-We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-\begin{Verbatim}
-  ResultSet : Type
-  ResultSet = List (List DBVal)
-\end{Verbatim}
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-\begin{Verbatim}
-executeSelect : 
-  String -> String -> List (Int, DBVal) -> 
-  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-                 (List DBVal)) -> 
-  Eff m [SQLITE ()] (Either String ResultSet)
-\end{Verbatim}
-% =================================================
-\subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate requests.
-This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
-In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.
-
-A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects.
-By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.
-
-We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
-%TODO: Fix up the labels
-
-\begin{figure}[h]
-\begin{Verbatim}[samepage]
-data Session : Effect where
-  LoadSession : 
-  SessionID -> 
-  Session (SessionRes SessionUninitialised) 
-          (SessionRes SessionInitialised) 
-          (Maybe SessionData)
-  UpdateSession : 
-    SessionData -> 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) ()
-  CreateSession : 
-    SessionData -> 
-    Session (SessionRes SessionUninitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  DeleteSession : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool 
-  WriteToDB : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool
-  DiscardSessionChanges : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) ()
-  GetSessionID : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  GetSessionData : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionData)
-\end{Verbatim}
-\caption{Session Effect}
-\label{fig:sessioneffect}
-\end{figure}
-
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {SessionUninitialised};
-  \node[state]         (B) [below of=A] {SessionInitialised};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
-\end{tikzpicture}
-}
-\caption{Session Handler Resource Usage Protocol}
-\label{fig:sessionstates}
-\end{figure}
-
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
-with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
-algebraic effect. In this application, there exist two states:
-In \texttt{SessionUninitialised}, the user may load or create a
-session.
-In \texttt{SessionInitialised}, the user may update the
-representation of the session in memory, serialise the session and write it to
-the database, or delete the session and invalidate the user's session key. The
-introduction of these two states ensures that changes are explicitly either
-written or discarded, eliminating the possibility of a developer updating the
-session but neglecting to commit it to persistent storage. This, of course, is
-under the assumption that the process exits cleanly: we attempt to facilitate
-this by writing total functions where possible.
-
-Much like the SQLite effect, we encapsulate failure by reflecting it in the
-resource associated with the effect.
-
-\begin{Verbatim}
-data SessionStep = SessionUninitialised
-                 | SessionInitialised
-
-data SessionRes : SessionStep -> Type where
-  InvalidSession : SessionRes s  
-  ValidSession   : SessionID -> 
-                   SessionData -> 
-                   SessionRes s
-\end{Verbatim}
-The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
-If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.
-
-% =================================================
-
-\section{Type-aware form handling}
-\label{form}
-Programming web applications often involves processing user data, which may
-then be used in further effectful computations. Data submitted using a form is
-transmitted over the internet as a string as part of an HTTP request, which
-traditionally involves losing associated type information.
-
-This can in turn lead to risks; developers may assume that data is
-of a certain type, and therefore discount the possibility that it may have been
-modified by an attacker. One example would be the traversal of paginated data,
-in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be
-used in a query such as:
-
-\begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + 
-       page + ', 5';
-\end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be
-modified by an attacker to include a malicious string which would alter the
-semantics of the query, allowing an attacker to execute a blind SQL injection
-attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
-clauses here
-
-In this section, we present a mechanism by which we introduce a DSL
-for the creation of web forms which preserve type information, implemented
-as a dependent algebraic effect. Once the form has
-been submitted, retrieved information is passed directly to a
-developer-specified function for handling, without the need to manually check
-and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and
-echoes it back. Firstly, we define a form handler which echoes back a string
-provided by the form handler. It has one argument of type \texttt{Maybe
-String}, which accounts for the possibility that the user may have specified
-incorrect data within the form.
-
-\begin{Verbatim}
-sayHello : Maybe String -> 
-           FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
-\end{Verbatim}
-
-We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
-
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [(handler args=[FormString], 
-                     effects=[CgiEffect], 
-                     fn=sayHello, 
-                     name="sayHello")
-           ]
-\end{Verbatim}
-
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
-
-\begin{Verbatim}
-showHelloForm : UserForm
-showHelloForm = do
-  addTextBox "Name" FormString Nothing
-  useEffects [CgiEffect]
-  addSubmit sayHello handlers
-\end{Verbatim}
-
-Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
-
-\begin{Verbatim}
-cgiHello : CGIProg [] ()
-cgiHello = do
-  handler_set <- isHandlerSet
-  if handler_set then do
-    handleForm handlers
-    return ()
-  else do
-    addForm "nameform" "helloform" showHelloForm
-    return ()
-
-main : IO ()
-main = runCGI [initCGIState] cgiHello
-\end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the
-page, requesting a name from the user. Upon submission of the form, the form
-handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
-of the form-handling system: namely, the effect which allows the creation of
-forms, and the handling code which deserialises the data and passes it to the
-user-specified handler function.
-
-\subsection{Form Construction}
-\label{formcons}
-Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-\begin{Verbatim}
-data FormTy = FormString
-            | FormInt
-            | FormBool
-            | FormFloat
-            | FormList FormTy 
-\end{Verbatim}
-Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-\begin{Verbatim}
-interpFormTy : FormTy -> Type
-interpFormTy FormString = String
-interpFormTy FormInt = Int
-interpFormTy FormBool = Bool
-interpFormTy FormFloat = Float
-interpFormTy (FormList a) = List (interpFormTy a)
-\end{Verbatim}
-%
-In order to specify a form, we once again use \texttt{Effects}. By recording
-the type of each form element as it is added in the type of the form, we may
-statically ensure that the user-supplied handler function is of the correct
-type to handle the data supplied by the form: the specification of an
-incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code
-through \texttt{using} notation. We may therefore parameterise the 
-form data types over
-the types associated with each form element, and the effects required by the
-handler function.
-
-\begin{Verbatim}
-using (G : List FormTy, E : List WebEffect)
-  data FormRes : List FormTy -> 
-                 List WebEffect -> Type where
-    FR : Nat -> 
-         List FormTy -> 
-         List WebEffect -> 
-         String -> 
-         FormRes G E
-
-  data Form : Effect where
-    AddTextBox : (label : String) -> 
-                 (fty : FormTy) -> 
-                 (Maybe (interpFormTy fty)) -> 
-                 Form (FormRes G E) 
-                      (FormRes (fty :: G) E) () 
-   ...
-
-    Submit : (mkHandlerFn ((reverse G), E)) ->
-             String -> 
-             Form (FormRes G E) (FormRes [] []) String
-\end{Verbatim}
-The implementation of the form effect also contains other constructs for
-additional form elements such as radio buttons and check boxes, but are omitted
-in the interest of brevity.
-
-We make use of the resource associated with the effect, \texttt{FormRes}, to
-construct the form. The resource allows us to record the types associated with
-each form element and the HTML required to display the form. he constructor,
-\texttt{FR}, requires the number of elements in the form in order to allow for
-the naming of new elements, the list of element types, the list of effects
-supported by the handler function, and the currently generated HTML for the
-form.
-
-By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
-It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
-
-By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource.
-The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
-
-To specify a form instance, we define a function of type \texttt{UserForm}:
-\begin{Verbatim}
-  UserForm : Type
-  UserForm = EffM m [FORM (FormRes []) 
-                          (FormRes [])] String
-\end{Verbatim}
-All forms are required to include a submit button, as mandated by the
-requirement that the input and output resource contains an empty list of types;
-this requirement is fulfilled as per the output resource type of the
-\texttt{AddSubmit} operation. As the creation of a form is a pure function
-which does not include side effects, we do not restrict the handler to IO as
-with previously-discussed effects, instead denoting the fact that it may be run
-in any handler with the implicit variable \texttt{m}.
-
-Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
-Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
-If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
-
-Before associating a handler function with the form, we must specify the effects available to the handler.
-This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
-By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-\begin{Verbatim}
-  useEffects : (effs : List WebEffect) ->
-               EffM m [FORM (FormRes G E)] 
-                      [FORM (FormRes G effs)] ()
-  useEffects effs = (UseEffects effs)
-\end{Verbatim}
-Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
-%
-We may specify a handler function of type \texttt{FormHandler}:
-\begin{Verbatim}
-  FormHandler : List EFFECT -> Type
-  FormHandler effs = Eff IO effs ()
-\end{Verbatim}
-In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-\begin{Verbatim}
-  addSubmit : 
-    (f :  mkHandlerFn ((reverse G), E)) ->
-    (fns : HandlerList) ->
-    {default tactics 
-      { applyTactic findFn 100; solve; }
-      prf : FnElem f fns} ->
-    EffM m [FORM (FormRes G E)]
-           [FORM (FormRes [] [])] 
-           String
-  addSubmit f handlers {prf} = (Submit f name)
-    where name : String
-          name = getString' f handlers prf          
-\end{Verbatim}
-
-Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-\begin{Verbatim}
-MkHandlerFnTy : Type
-MkHandlerFnTy = (List FormTy, List WebEffect)
-
-mkHandlerFn' : List FormTy -> List WebEffect -> Type
-mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
-mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
-                              mkHandlerFn' xs effs
-
-mkHandlerFn : MkHandlerFnTy -> Type 
-mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{Verbatim}
-The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
-When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-
-To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{Verbatim}
-HandlerFn : Type
-HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{Verbatim}
-%
-The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
-We also store a unique string identifier, which is used to serialise a reference to the handler function.
-
-In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
-\begin{Verbatim}
-syntax 
-  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
-  ", name=" [name] = ((args, effs) ** (fn, name))
-\end{Verbatim}
-We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
-This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
-\begin{Verbatim}
-handler args=[FormString], 
-        effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
-\end{Verbatim}
-
-We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-\begin{Verbatim}
-HandlerList : Type
-HandlerList = List HandlerFn
-\end{Verbatim}
-To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-\begin{Verbatim}
-  data FnElem : mkHandlerFn ((reverse G), E) -> 
-                HandlerList -> Type where
-
-       FnHere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
-       FnThere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-               FnElem f xs -> FnElem f (x :: xs)
-\end{Verbatim}
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
-\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
-%TODO: this should probably be rewritten
-We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
-The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail.
-
-Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.
-
-Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this.
-Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.
-
-Running form construction is achieved as an operation of the CGI
-effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a
-hidden HTML field.
-
-\subsection{Form Handling}
-\label{formhandling}
-Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
-
-With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
-We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
-
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
-This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
-If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
-If not, the computation is abandoned and an error is shown.
-% I gave it a shot, right?
-:P
-\begin{Verbatim}
-checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
-                 (frm_fn_ty : MkHandlerFnTy) -> 
-                 mkHandlerFn reg_fn_ty -> 
-                 Maybe (mkHandlerFn frm_fn_ty)
-checkFunctions reg_ty frm_ty reg_fn with 
-                             (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
-                                            Just reg_fn
-  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{Verbatim}
-We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
-Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
-
-%-----------------------------
-%-----------------------------
-
-\section{Extended Example: Message Board}
-\label{messageboard}
-Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
-
-Firstly, we create a database schema in which to record information stored by the message board.
-We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
-
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
-Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
-\subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-\begin{Verbatim}
-main : IO ()
-main = do (runCGI [initCGIState, 
-                  InvalidSession, ()] 
-           handleRequest)
-          pure ()
-\end{Verbatim}
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-\begin{Verbatim}
-handleRequest : CGIProg [SESSION 
-  (SessionRes SessionUninitialised), 
-  SQLITE ()] ()
-handleRequest = do 
-  handler_set <- isHandlerSet
-  if handler_set then do
-    lift' (handleForm handlers)
-    Effects.pure ()
-  else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
-    handleNonFormRequest action (map strToInt thread_id)
-\end{Verbatim}
-\subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
-One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-\begin{Verbatim}
-newThreadForm : UserForm
-newThreadForm = do
-  addTextBox "Title" FormString Nothing
-  addTextBox "Post Content" FormString Nothing 
-  useEffects [CgiEffect, SessionEffect, SqliteEffect]
-  addSubmit handleNewThread handlers
-\end{Verbatim}
-The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
-Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
-The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleRegisterForm, 
-          name="handleRegisterForm"),
-
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleNewThread, 
-          name="handleNewThread"),
-     ...
-  ]
-\end{Verbatim}
-Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
-In order to do this, we make use of the session handler.
-We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
-If not, then a failure function is called instead.
-Should the form handler function be called with invalid arguments, an error is shown.
-\begin{Verbatim}
-handleNewThread : 
-  Maybe String -> Maybe String -> 
-  FormHandler [CGI (InitialisedCGI TaskRunning), 
-               SESSION (SessionRes SessionUninitialised), 
-               SQLITE ()]  
-handleNewThread (Just title) (Just content) = do 
-  withSession (addNewThread title content) notLoggedIn
-  pure ()
-handleNewThread _ _ = do 
-  outputWithPreamble "<h1>Error</h1><br />There was 
-       an error posting your thread."
-  pure ()
-\end{Verbatim}
-Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
-The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
-This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
-In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
-In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-\begin{Verbatim}
-threadInsert : Int -> 
-               String -> 
-               String -> 
-               Eff IO [SQLITE ()] Bool
-threadInsert uid title content = do
-  let query = "INSERT INTO `Threads` 
-    (`UserID`, `Title`) VALUES (?, ?)"
-  insert_res <- (executeInsert DB_NAME query 
-    [(1, DBInt uid), (2, DBText title)]
-  case insert_res of
-    Left err => pure False
-    Right thread_id => postInsert uid thread_id content
-\end{Verbatim}
-\subsection{Listing Threads}
-Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-\begin{Verbatim}
-getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = 
-  executeSelect DB_NAME query [] collectThreadResults
- where query = "SELECT `ThreadID`, `Title`, `UserID`, 
-   `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{Verbatim}
-Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
-Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-\begin{Verbatim}
-traverseThreads : ResultSet -> 
-  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = pure ()
-traverseThreads (x::xs) = do traverseRow x
-                             traverseThreads xs
-  where traverseRow : List DBVal -> 
-           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-        traverseRow ((DBInt thread_id)::
-                     (DBText title)::
-                     (DBInt user_id)::
-                     (DBText username)::[]) =
-           (output $ "<tr><td>
-            <a href=\"?action=showthread&thread_id=" ++ 
-            (show thread_id) ++ "\">" ++ 
-            title ++ "</a></td><td>" ++ 
-            username ++ "</td></tr>") 
-        traverseRow _ = pure ()
-\end{Verbatim}
-\subsection{Authentication}
-Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
-If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
-The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
-Any failures, for example with creating a new session or querying the database, are reported to the user.
-\begin{Verbatim}
-handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
-  case auth_res of
-    Right (Just uid) => do
-      set_sess_res <- setSession uid
-      if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
-        pure ()
-      else do
-        lift' (output "Could not set session")
-        pure ()
-    Right Nothing => do
-      lift' (output "Invalid username or password")
-      pure ()
-    Left err => do
-      lift' (output $ "Error: " ++ err)
-      pure ()
-\end{Verbatim}
-Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
-
-Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
-Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
-Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
-%-----------------------------
-%-----------------------------
-
-\section{Related Work}
-\citet{meijer:cgi} implemented a CGI library which was among the first
-libraries to handle web scripting monadically, and allows the user to implement
-application logic without having to consider the low-level details such as
-parsing in CGI data from the environment, or printing headers to the remote
-browser. The library also provides support for cookies and basic form handling.
-
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining
-state, and provides more sophisticated form-handling methods. In particular,
-callbacks may be associated with submit buttons, with nameless representations
-for form inputs. Due to the unavailability of full dependent types in Haskell,
-however, this implementation does not statically verify the suitability of the
-callback function for the form inputs. Both implementations of the CGI library,
-being built upon monads, mean that the use of additional effects such as
-database access is achieved either through monad transformers or the ability to
-perform arbitrary IO operations. Both of these approaches are limited---the
-former does not scale well to multiple effects, and the latter allows for the
-introduction of errors by allowing the violation of resource usage protocols.
-
-\citet{plasmeijer:idata} describe an alternative approach to type-safe form
-handling which does not require the use of dependent types through the
-\textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of
-processing being triggered by form submission, as in the approach described in
-this paper, applications created in the \idata{} toolkit are instead
-edit-driven. This means that upon a component being edited, a computation
-occurs, given the state of the current form. This is then saved to be used in
-future computations. Should a user enter invalid data, for example by entering
-text in a field designated for integers, the change will be reverted. This is
-demonstrated practically through the use of \idata{} to implement a conference
-management system \cite{plasmeijer:cms}.
-
-The concept of \idata{} is taken further by the introduction of \itasks{}
-\cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple
-\idata{} forms to interact with one another. This is achieved through the use
-of high-level combinators which allow the implementation of concepts such as
-recurseion, sequence and choice in a scalable fashion. These high-level
-abstractions are elegant, but the style and syntax differ substantially from a
-traditional web application. Our approach takes the concept of type-safe input
-handling and uses it in a more traditional fashion, whilst retaining the
-type-retention guarantees afforded by \idata{} elements.
-
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
-dependent types to enable the generation of provably-correct and secure queries
-and DOM code, without requiring overly complex types or proofs. Through the use
-of the Ur\/Web DSL, developers may generate well-typed code from records. By
-statically verifying the \textit{code generator}, it is therefore possible to
-assert that the generated code is secure. Such ideas are extremely interesting
-for the implementation of object-relational mapping, which we hope to implement
-in the near future.
-
-WebDSL \cite{webdsl} is a domain-specific language written primarily to
-introduce new abstractions which aim to reduce the amount of boilerplate code
-that must be written and maintained by developeres. WebDSL is built on top of
-Java, which often includes a large amount of redundant code such as accessor
-and mutator functions within entity classes. The Java Persistence API (JPA)
-\cite{jpa} provides an object-relational mapping through the use of Java 5
-annotations, which may then be used to construct database tables. These
-annotations soon become complex, however, and coupled with redundant
-boilerplate code, data model declarations may soon become unwieldy. Through the
-use of WebDSL, these data model declarations can be much more compactly
-declared, and elaborated into Java code by parsing the data-modelling DSL into
-an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL
-also applies similar concepts to implement a \textit{template system} for such
-objects, which allows the data to be used in code generation. We look to
-implement many of these ideas, but as effects within the IdrisWeb
-framework, as with the form construction effect.
-
-%-----------------------------
-%-----------------------------
-
-\section{Conclusions}
-Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
-
-Our framework, through the use of dependent types, provides several static guarantees.
-Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
-This immediately eliminates the possibilities of SQL injection attacks on non-string types.
-Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
-Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
-
-Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
-Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
-By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
-
-In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
-With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
-
-Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
-\subsection{Further Work}
-We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
-
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
-
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
-We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
-
-Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
-Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
-Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
-
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
-Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
-This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
-Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
-Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
-
-Dependently-typed languages provide great promise for the construction of secure and correct programs.
-Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
-% Cryptography bindings would be really good.
-% Integration with a web server instead of doing everything over CGI
-% Improvements to form handling system
-% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%-----------------------------
-%-----------------------------
-
-\acks
-This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
-The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
-% SICSA / EPSRC (grant number?
-% #idris
-% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{plainnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{refs}
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-11-11 paper/forms.tex
Added paper revisions so far Also started the rewrite of the DB section with the new DB stuff

\section{Type-aware form handling}

\label{form}
Programming web applications often involves processing user data, which may
then be used in further effectful computations. Data submitted using a form is
transmitted over the internet as a string as part of an HTTP request, which
traditionally involves losing associated type information.

This can in turn lead to risks; developers may assume that data is
of a certain type, and therefore discount the possibility that it may have been
modified by an attacker. One example would be the traversal of paginated data,
in which a form is used to make a request to retrieve the next page of data.
This may involve sending an integer detailing the current page, which could be
used in a query such as:

\begin{SaveVerbatim}{selectq}

'SELECT `name`, `address` FROM `staff` LIMIT ' + 
       page + ', 5';

\end{SaveVerbatim}
\useverb{selectq}

\noindent
The \texttt{page} variable is assumed to be an integer, but may instead be
modified by an attacker to include a malicious string which would alter the
semantics of the query, allowing an attacker to execute a blind SQL injection
attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
% clauses here

In this section, we present a mechanism by which we introduce a DSL
for the creation of web forms which preserve type information, implemented
as a dependent algebraic effect. Once the form has
been submitted, retrieved information is passed directly to a
developer-specified function for handling, without the need to manually check
and deserialise data.

We begin with a simple example of a form which requests a user's name, and
echoes it back. Firstly, we define a form handler which echoes back a string
provided by the form handler. It has one argument of type \texttt{Maybe
String}, which accounts for the possibility that the user may have specified
incorrect data within the form.

\begin{SaveVerbatim}{sayhello}

sayHello : Maybe String -> 
           FormHandler [CGI (InitialisedCGI TaskRunning)]
sayHello (Just name) = output ("Hello, " ++ name ++ "!")
sayHello _ = output "Error!"

\end{SaveVerbatim}
\useverb{sayhello}

We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:

\begin{SaveVerbatim}{handerlist}

handlers : HandlerList
handlers = [handler args=[FormString], 
                    effects=[CgiEffect], 
                    fn=sayHello, 
                    name="sayHello"]

\end{SaveVerbatim}
\useverb{handerlist}

We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.

\begin{SaveVerbatim}{showhello}

showHelloForm : UserForm
showHelloForm = do
  addTextBox "Name" FormString Nothing
  useEffects [CgiEffect]
  addSubmit sayHello handlers

\end{SaveVerbatim}
\useverb{showhello}

\noindent
Finally, we specify that if data has been submitted for processing, then it
should be passed to the form handler. If not, then the form should be shown.

\begin{SaveVerbatim}{cgihello}

cgiHello : CGIProg [] ()
cgiHello = do
  handler_set <- isHandlerSet
  if handler_set then do
    handleForm handlers
    return ()
  else do
    addForm "nameform" "helloform" showHelloForm
    return ()

main : IO ()
main = runCGI [initCGIState] cgiHello

\end{SaveVerbatim}
\useverb{cgihello}

\noindent
When this CGI application is invoked, it will begin by outputting a form to the
page, requesting a name from the user. Upon submission of the form, the form
handler will be invoked, and the name will be used in the output.

In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
of the form-handling system: namely, the effect which allows the creation of
forms, and the handling code which deserialises the data and passes it to the
user-specified handler function.

\subsection{Form Construction}
\label{formcons}
Each form element is specified  to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.

\begin{SaveVerbatim}{formty}

data FormTy = FormString | FormInt
            | FormBool   | FormFloat
            | FormList FormTy

\end{SaveVerbatim}
\useverb{formty}

\noindent
Since types in \idris{} are first-class, we may use this to convert
between abstract and concrete representations of allowed form types:

\begin{SaveVerbatim}{interpformty}

interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)

\end{SaveVerbatim}
\useverb{interpformty}

\begin{SaveVerbatim}{formeff}
using (G : List FormTy, E : List WebEffect)
  data FormRes : List FormTy -> List WebEffect -> Type where
    FR : Nat -> List FormTy -> List WebEffect -> String -> FormRes G E

  data Form : Effect where
    AddTextBox      : (label : String) -> (fty : FormTy) -> (Maybe (interpFormTy fty)) -> 
                      Form (FormRes G E) (FormRes (fty :: G) E) () 
    AddSelectionBox : (label : String) -> (fty : FormTy) -> (vals : Vect m (interpFormTy fty)) -> 
                      (names : Vect m String) ->
                      Form (FormRes G E)  (FormRes (fty :: G) E) ()
    AddRadioGroup   : (label : String) -> (fty : FormTy) -> (vals : Vect m (interpFormTy fty)) ->
                      (names : Vect m String) -> (default : Int) ->
                      Form (FormRes G E)  (FormRes (fty :: G) E) ()
    AddCheckBoxes   : (label : String) -> (fty : FormTy) -> (vals : Vect m (interpFormTy fty)) ->
                      (names : Vect m String) -> (checked_boxes : Vect m Bool) ->
                      Form (FormRes G E)  (FormRes ((FormList fty) :: G) E) ()
    Submit          : (mkHandlerFn ((reverse G), E)) -> String -> 
                      Form (FormRes G E)  (FormRes [] [])       String
\end{SaveVerbatim}

\begin{figure*}[t]
\begin{center}
\useverb{formeff}
\end{center}
\caption{Form Effect}
\label{fig:formeffect}
\end{figure*}

In order to specify a form, we once again use \texttt{Effects}. By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: the specification of an
incompatible handler will result in a compile-time type error. The \texttt{Form}
effect and associated resource \texttt{FormRes} is given in Figure 
\ref{fig:formeffect}.
The \texttt{using} notation here indicates that within
the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
the given type.

The general process of form construction is illustrated by the \texttt{AddTextBox}
and \texttt{Submit} operations of the \texttt{Form} effect:

\begin{SaveVerbatim}{formeffsmall}

data Form : Effect where
  AddTextBox : (label : String) -> (fty : FormTy) -> 
               Maybe (interpFormTy fty) -> 
               Form (FormRes G E) 
                    (FormRes (fty :: G) E) () 
     ...
  Submit : (mkHandlerFn ((reverse G), E)) -> String -> 
           Form (FormRes G E) (FormRes [] []) String

\end{SaveVerbatim}
\useverb{formeffsmall}

\noindent
These make use of the resource associated with the effect, \texttt{FormRes}, to
construct the form. Adding a field such as a text box adds a new type, \texttt{fty}
to the list of field types, carried in the resource. When the form is complete,
the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field
types, and using it to construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

\begin{SaveVerbatim}{userform}

UserForm : Type
UserForm = EffM m [FORM (FormRes []) 
                        (FormRes [])] String

\end{SaveVerbatim}
\useverb{userform}

\noindent
All forms are required to include a submit button, as mandated by the
requirement that

The input and output resource contains an empty list of types, which means that
any form which includes fields must also include a submit button. Adding fields
adds to the list of types, and only adding a submit button empties that list.
Note that there is no need to restrict this effect to running in the \texttt{IO}
monad since creating a form merely returns HTML text, with no side-effects by
default.

Handlers may only be associated with a form if they have argument types
corresponding to the types associated with the form elements. Additionally, we
wish to name the function in order for it to be serialised, whilst requiring a
proof that the specified name is associated with the function. If this were not
the case, it would be possible to specify a function which satisfies the type
requirement, without guaranteeing that it the serialised data corresponded to
that function, thus rendering the check pointless.

Before associating a handler function with the form, we must specify the
effects available to the handler. This is done through the use of the
\texttt{useEffects}, which updates the list of effects in the type of the form
resource. By doing this, we may subsequently use the effects in calculations at
the type level, in particular when calculating the type of the handler function
for the form.

\begin{SaveVerbatim}{useeffs}

useEffects : (effs : List WebEffect) ->
             EffM m [FORM (FormRes G E)] 
                    [FORM (FormRes G effs)] ()
useEffects effs = (UseEffects effs)

\end{SaveVerbatim}
\useverb{useeffs}

\noindent
A \texttt{WebEffect} is an effect which is usable in a web application, and can
be converted to an \texttt{EFFECT} using:

\begin{SaveVerbatim}{interpwebeff}

webEffect : WebEffect -> EFFECT

\end{SaveVerbatim}
\useverb{interpwebeff}

Whilst it is not possible to serialise arbitrary effects due to the associated
difficulties with serialising initial resource environments, we allow for three
effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
This is, however, not an inherent limitation as the \texttt{Effects} library
permits introduction of additional effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:

\begin{SaveVerbatim}{formhandler}

FormHandler : List EFFECT -> Type
FormHandler effs = Eff IO effs ()

\end{SaveVerbatim}
\useverb{formhandler}

\noindent
In order to associate a handler with a form, we may call the \texttt{addSubmit}
function:

\begin{SaveVerbatim}{addsubmit}

addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
            (fns : HandlerList) ->
            {default tactics 
              { applyTactic findFn 100; solve; }
              prf : FnElem f fns} ->
            EffM m [FORM (FormRes G E)]
                   [FORM (FormRes [] [])] 
                   String
addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf

\end{SaveVerbatim}
\useverb{addsubmit}

\noindent
This function takes a handler function and a list of available handlers, along with
an automatically constructed proof (using the a \texttt{default} argument) that the
handler is available.
Let us look at each aspect of this function in turn. Firstly, the
\texttt{mkHandlerFn} function calculates the required type of the handler
function from the list of types associated with the form elements, and the
effects we specified with \texttt{useEffects}. Note that since we prepend types
to the list of \texttt{FormTy}s as opposed to appending them, we must reverse
the list.

\begin{SaveVerbatim}{mkhandlerfnty}

MkHandlerFnTy : Type
MkHandlerFnTy = (List FormTy, List WebEffect)

\end{SaveVerbatim}

\begin{SaveVerbatim}{mkhandlerfn}

mkHandlerFn' : List FormTy -> List WebEffect -> Type
mkHandlerFn' [] effs = FormHandler (map webEffect effs) 
mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
                              mkHandlerFn' xs effs

mkHandlerFn : MkHandlerFnTy -> Type 
mkHandlerFn (tys, effs) = mkHandlerFn' tys effs

\end{SaveVerbatim}

\useverb{mkhandlerfnty}

\useverb{mkhandlerfn}

\noindent
The \texttt{mkHandlerFn} function takes a tuple describing the arguments and
web effects available to the handler function. When constructing the function
type, we encase all arguments withing \texttt{Maybe} type, in order to handle
failure should the supplied data fail to parse to the specified type.
%
To store a reference to a handler function, we use the \texttt{HandlerFn} type:

\begin{SaveVerbatim}{handlerfn}

HandlerFn : Type
HandlerFn = (ft ** (mkHandlerFn ft, String))

\end{SaveVerbatim}
\useverb{handlerfn}

\noindent
The \texttt{**} notation denotes a dependent pair, where the type of the second
element of the pair is parameterised over the value of the first element. It is
an existential binding;
the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
\texttt{P x} holds''.
Therefore a \texttt{HandlerFn} states that there exists a function type
\texttt{ft} such that we have a handler for it, and a unique string identifier
which is used to serialise a
reference to the handler function.

In order to abstract away from this implementation detail, we make use of
\idris{} syntax rewriting rules. This allows us to define the following syntax
rewrite rule:

\noindent
\begin{SaveVerbatim}{syntaxhandler}

syntax 
  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
  ", name=" [name] = ((args, effs) ** (fn, name))

\end{SaveVerbatim}
\useverb{syntaxhandler}

\noindent
We may then define handlers in a more intuitive fashion, without being
concerned with the implementation details. This allows us to write a handler
with one String argument, making use of the CGI effect, associated with the
\texttt{sayHello} handler function as follows:

\begin{SaveVerbatim}{handlerex}

handler args=[FormString], 
        effects=[CgiEffect], 
        fn=sayHello, 
        name="sayHello"

\end{SaveVerbatim}
\useverb{handlerex}

\noindent
We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.

\begin{SaveVerbatim}{handlerlist}

HandlerList : Type
HandlerList = List HandlerFn

\end{SaveVerbatim}
\useverb{handlerlist}

To enforce the requirement that a supplied handler function must reside in the
list of available handlers, and therefore allow us to retrieve the name with
which to serialise the handler, we require a \textit{list membership proof},
\texttt{FnElem f fns}, which statically guarantees that a given item resides in
a list.

\begin{SaveVerbatim}{fnelem}

using (xs : HanderList , f : mkHandler (reverse G, E))
  data FnElem : mkHandlerFn ((reverse G), E) -> 
                HandlerList -> Type where
       FnHere  : FnElem f (((reverse G, E) ** 
                                  (f, fStr)) :: xs)
       FnThere : FnElem f xs -> FnElem f (x :: xs)

\end{SaveVerbatim}
\useverb{fnelem}

\noindent
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the
form elements and the effects used by the handler function. \texttt{FnHere}
is a proof that the element is at the head of the current point of the list,
whereas \texttt{FnThere} is a proof that the element is in the tail of 
the list. 
%TODO: this should probably be rewritten
We then use a simple automated proof search to
automatically generate the proof at compile time, should one exist. The proof
may then be used in subsequent computations: in our case, we use it to retrieve
the unique identifier for the function. If the automated proof search fails,
compilation will fail.

Finally, we serialise the argument types, supported effects, and return
type of the handler, to allow the form data to be 
deserialised and ensure that the correct handler is executed on the
server.

Although sending details of the handler function to the client may appear to be
a security risk, we envisage that the use of symmetric encryption or a
cryptographic nonce would alleviate this. Ultimately, we hope to implement a
web server with persistent state, which would eliminate the need for
serialisation altogether.

Running form construction is achieved as an operation of the CGI
effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
The generated metadata describing the handler function is serialised as a
hidden HTML field.

\subsection{Form Handling}
\label{formhandling}
Once the form has been submitted, a web application may handle the submitted
data by invoking \texttt{HandleForm}. This will check for the
existence of the hidden \texttt{handler} field, which contains the previously
serialised metadata about the form handler, before deserialising the
data into a \texttt{MkHandlerFnTy}.

With this data, we then look up the function in the list of registered
handlers by using the unique handler identifier. In order to apply
the handler function to the data submitted in the form, we must first prove to
the type checker that the deserialised \texttt{MkHandlerFnTy} is the same as
the one retrieved from the list of registered handlers. We do this by making
use of the \texttt{decEq} function, which determines whether two types are
equal, returning a proof of equality if so, and a proof of inequality if not.

\begin{SaveVerbatim}{deceqty}

decEq : DecEq t => (x : t) -> (y : t) -> Dec (x1 = x2)

\end{SaveVerbatim}
\useverb{deceqty}

We then use the \texttt{with} construct, inspired by
\textit{views} in Epigram \cite{mcbride.mckinna:viewfromleft}, to rewrite the
arguments on the left hand side. This allows us to construct a function which,
given the stored handler, the data required to
construct the function type and the \texttt{MkHandlerFnTy} deserialised from
the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably
equal. If so, we rewrite this on the left hand side since the equality proof
demonstrates that the recorded function may also be used to handle the form
data. If not, the computation fails.

\begin{SaveVerbatim}{checkfunctionsty}

checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
                 (frm_fn_ty : MkHandlerFnTy) -> 
                 mkHandlerFn reg_fn_ty -> 
                 Maybe (mkHandlerFn frm_fn_ty)
\end{SaveVerbatim}

\begin{SaveVerbatim}{checkfunctions}
checkFunctions reg_ty frm_ty reg_fn with 
                             (decEq reg_ty frm_ty)
  checkFunctions frm_ty frm_ty reg_fn 
                        | Yes refl = Just reg_fn
  checkFunctions reg_ty frm_ty reg_fn 
                        | No _ = Nothing

\end{SaveVerbatim}

\useverb{checkfunctionsty}

\useverb{checkfunctions}

\noindent
We may then parse the arguments according to the types specified by the handler
function, and then apply the arguments to the handler function through the use
of partial application. Finally, we may run the handler function, ensuring that
all updates made to the CGI state are propagated.

%-----------------------------
%-----------------------------

### 2013-11-11 paper/messageboard.tex
Added paper revisions so far Also started the rewrite of the DB section with the new DB stuff

\section{Extended Example: Message Board}
\label{messageboard}
In this section we consider a larger example---a message board application
which allows users to register, log in, view and create threads, and list and
create new posts in threads.

Firstly, we create a database schema in which to record information stored by
the message board. We create three tables: \texttt{Users}, which contains a
unique User ID, usernames and passwords; \texttt{Threads}, which contains a
unique thread ID, a title, and the ID of the user who created the thread; and
\texttt{Posts}, which contains a unique post ID, the ID of the thread to which
each post belongs, the content of the post, and the ID of the user that created
the post.

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the
message board should be displayed, and pattern-match on these to call the
appropriate function which displays the page. Some pages, such as the page
which shows all of the posts in a thread, require a second argument,
\texttt{thread\_id}.  \subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.

\begin{SaveVerbatim}{msgmain}

main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          return ()

\end{SaveVerbatim}
\useverb{msgmain}

\noindent
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.

\begin{SaveVerbatim}{handlereq}

handleRequest : CGIProg 
     [SESSION (SessionRes SessionUninitialised), 
      SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.return ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)

\end{SaveVerbatim}
\useverb{handlereq}

\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in,
one to handle the creation of new threads, and one to handle the creation of
new posts. For example, the form used to create a new thread
contains elements for the title of the new thread and the content of the
first post of the new thread:

\begin{SaveVerbatim}{newthread}

newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers

\end{SaveVerbatim}
\useverb{newthread}

\noindent
This consists of two text boxes: one for the title of the thread, and one
for the content of the first post. Both are of type \texttt{String}, as denoted
by the \texttt{FormString} argument, and both have no default value. The
handler function may make use of the \texttt{CGI}, \texttt{SESSION} and
\texttt{SQLITE} effects, and the handler function is specified as
\texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of
form handlers, and is of the following form:

\begin{SaveVerbatim}{msghandlers}

handlers : HandlerList
handlers = [
 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleRegisterForm, 
          name="handleRegisterForm"),

 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleNewThread, 
          name="handleNewThread"),
     ...]

\end{SaveVerbatim}
\noindent
\useverb{msghandlers}

\noindent
Creating a new thread (show in Figure \ref{fig:handlethread})
requires a user to be logged in, so that the thread
starter may be recorded in the database. In order to do this, we make use of
the session handler. We define a function \texttt{withSession}, which attempts
to retrieve the session associated with the current request, and if it exists,
executes a function which is passed the associated session data. If not, then a
failure function is called instead. Should the form handler function be called
with invalid arguments, an error is shown.

\begin{SaveVerbatim}{handlenewthread}
handleNewThread : 
  Maybe String -> Maybe String -> 
  FormHandler [CGI (InitialisedCGI TaskRunning), 
               SESSION (SessionRes SessionUninitialised), 
               SQLITE ()]  
handleNewThread (Just title) (Just content) = do 
  withSession (addNewThread title content) notLoggedIn
  return ()
handleNewThread _ _ = do 
  outputWithPreamble "<h1>Error</h1><br />There was 
       an error posting your thread."
  return ()
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{handlenewthread}
\caption{Thread Creation}
\label{fig:handlethread}
\end{figure}

\noindent
Once we have loaded the session data from the database, we then check whether
the \texttt{UserID} variable is set, which demonstrates that a user has
successfully logged into the system, and allows us to use the ID in subsequent
computations. The database operation to insert the thread into the database is
performed by \texttt{threadInsert}, shown in Figure \ref{fig:threadins}.
This uses a library function \texttt{executeInsert}, which abstracts over the
low-level resource usage protocol, enabling for provably-correct database
access without the excess boilerplate code. In addition, \texttt{executeInsert}
returns the unique row ID of the last item which was inserted, which may be
used in subsequent computations. In the case of the message board, we use this
to associate the first post of the thread with the thread being inserted.

\begin{SaveVerbatim}{threadins}
threadInsert : Int -> String -> String -> 
               Eff IO [SQLITE ()] Bool
threadInsert uid title content = do
  let query = "INSERT INTO `Threads` 
    (`UserID`, `Title`) VALUES (?, ?)"
  insert_res <- (executeInsert DB_NAME query 
    [(1, DBInt uid), (2, DBText title)]
  case insert_res of
    Left err => return False
    Right thread_id => postInsert uid thread_id content
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{threadins}
\caption{Thread Insertion}
\label{fig:threadins}
\end{figure}

\subsection{Listing Threads}

Listing the threads in the database is achieved using 
\texttt{executeSelect}, which returns either a
\texttt{ResultSet} or an error:

\begin{SaveVerbatim}{getthreads}

getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
getThreads = 
    executeSelect DB_NAME query [] collectThreadResults
 where query = "SELECT `ThreadID`, `Title`, `UserID`, 
         `Username` FROM `Threads` NATURAL JOIN `Users`"

\end{SaveVerbatim}
\noindent
\useverb{getthreads}

\noindent 
Once the result set has been retrieved, we may iterate through the
results and output them to the page, including a link to a page which shows the
posts associated with the thread. This is shown in Figure \ref{fig:traverse}.
Since we know the structure of the returned
row from designing the query, we may pattern match on each returned row to make
use of the returned values.

\begin{SaveVerbatim}{traversethreads}
traverseThreads : ResultSet -> 
  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
traverseThreads [] = return ()
traverseThreads (x::xs) = do traverseRow x
                             traverseThreads xs
  where traverseRow : List DBVal -> 
           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
        traverseRow ((DBInt thread_id)::
                     (DBText title)::
                     (DBInt user_id)::
                     (DBText username)::[]) =
           (output $ "<tr><td>
            <a href=\"?action=showthread&thread_id=" ++ 
            (show thread_id) ++ "\">" ++ 
            title ++ "</a></td><td>" ++ 
            username ++ "</td></tr>") 
        traverseRow _ = return ()
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{traversethreads}
\caption{Thread Insertion}
\label{fig:traverse}
\end{figure}

\subsection{Authentication}

Once a user submits the login form, the associated handler queries the database
to ascertain whether a user with the given username and password exists through
a call to the \texttt{authUser} function. This is shown in Figure
\ref{fig:handlelogin}. If so, then the session handler is
invoked, and a session is initialised with the user ID retrieved from the
database. The session ID is then set as a cookie using the CGI effect, so that
it may be used in subsequent requests. Any failures, for example with creating
a new session or querying the database, are reported to the user.

\begin{SaveVerbatim}{handlelogin}
handleLoginForm (Just name) (Just pwd) = do
  auth_res <- lift' (authUser name pwd)
  case auth_res of
    Right (Just uid) => do
      set_sess_res <- setSession uid
      if set_sess_res then do
        lift' (output $ "Welcome, " ++ name)
        return ()
      else do
        lift' (output "Could not set session")
         return ()
    Right Nothing => do
      lift' (output "Invalid username or password")
      return ()
    Left err => do
      lift' (output $ "Error: " ++ err)
      return ()
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{handlelogin}
\caption{Thread Insertion}
\label{fig:handlelogin}
\end{figure}

Implementations for the insertion and display of posts, alongside registration,
are similar in form to those described above, and are as such omitted from this
paper.

Although we have described a relatively simple application, we have shown that
through the use of embedded domain-specific languages, we may write verified
code that fails to compile should resources be incorrectly accessed.
Additionally, we have used the form handling mechanism to simply handle the
arguments passed by the user. Importantly, we have shown that dependent types
can be used to increase confidence in an (albeit simplified) real-world
application, without requiring developers to supply proofs or indeed work
explicitly with dependent types.

%-----------------------------
%-----------------------------

### 2013-11-11 paper/protocols.tex
Added paper revisions so far Also started the rewrite of the DB section with the new DB stuff

--- 2013-08-14 paper/main.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-%\usepackage{silence}
-\usepackage{pgf}
-\usepackage{tikz}
-\usepackage{url}
-\usepackage[hidelinks]{hyperref}
-\usepackage{fancyvrb}
-\usepackage[numbers]{natbib}
-\usetikzlibrary{arrows,automata}
-%\WarningsOff
-
-\begin{document}
-\fvset{fontsize=\small}
-\newcommand{\idris}{\textsc{Idris}}
-\newcommand{\idata}{\textsf{iData}}
-\newcommand{\itasks}{\textsf{iTasks}}
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{2013} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{}        % These are ignored unless
-\preprintfooter{}   % 'preprint' option specified.
-
-\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
-%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
-
-\authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
-           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
-
-\maketitle
-
-\begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within
-programs in turn allowing easier reasoning about the operation of such
-programs. Stronger specifications do however bring with them the disadvantage
-that it becomes increasingly difficult to write programs that are accepted by
-the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in high-level languages which use dependent
-types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an
-EDSL to facilitate the creation and handling of web forms which retain type
-information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how EDSLs may be constructed to enforce resource usage
-protocols associated with common web operations such as CGI, database access
-and session handling.
-
-\end{abstract}
-
-%\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-%\keywords
-%keyword1, keyword2
-
-\section{Introduction}
-
-Web applications, whilst ubiquitous, are also prone to incorrect construction
-and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
-scripting \cite{owasp:xss}. Security breaches using such exploits are
-far-reaching, and high profile cases involve large corporations such as Sony,
-who suffered a well-publicised and extremely costly SQL Injection breach in
-2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
-\cite{imperva:yahoo}.
-
-Many web applications are written in dynamically-checked scripting languages
-such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. A significant drawback, however, is that such languages
-do not provide  the same static guarantees about runtime behaviour afforded by
-programs with more expressive, static type systems, instead relying on
-extensive unit testing to ensure correctness and security.
-
-%%------- EXAMPLE
-Let us consider a simple database access routine, written in
-PHP, where we wish to obtain the name and address of every employee working in
-a given department, \texttt{\$dept}. We firstly construct an object
-representing a database connection, where the arguments are the database host,
-user, password and name respectively:
-
-%
-\begin{Verbatim}
-$conn = new mysqli("localhost", "username", 
-                    "password", "db");
-\end{Verbatim}
-%
-We then check to see if the connection was successful, and exit
-if not.  This check is optional, so
-it would be possible to omit it: this would, however, cause
-problems with later steps.
-
-%
-\begin{Verbatim}
-  if (mysqli_connect_errno()) {
-      exit();
-  }
-\end{Verbatim}
-%
-We then create a prepared statement detailing our query, and bind the
-\texttt{`dept'} value:
-
-\begin{Verbatim}
-  $stmt = $conn->prepare("SELECT `name`, `address` 
-     FROM `staff` WHERE `dept` = ?);
-  $stmt->bind_param('s', $dept);
-\end{Verbatim}
-%
-After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. At
-this point, a developer may forget to execute the statement before beginning to
-fetch results or may have forgotten to bind the parameters to the statement:
-
-%
-\begin{Verbatim}
-  $stmt->execute();
-  $stmt->bind_result($name, $address);
-  while ($stmt->fetch()) {
-    printf("Name: %s, Age: %s", $name, $age);
-  }
-\end{Verbatim}
-%
-Finally, once the statement and connection are no longer needed, they should be
-closed in order to discard the associated resources:
-
-%
-\begin{Verbatim}
-  $stmt->close();
-  $conn->close();
-\end{Verbatim}
-%
-Even in this small example, there exists a specific resource usage protocol.
-Firstly, a connection to the database must be opened. The object-oriented style
-used in the example encapsulates this to an extent, as the object must be
-created in order for operations to be performed, however it is less obvious in
-a procedural version of the code. Secondly, a prepared statement is created,
-using the raw SQL and placeholders to which variables are later bound. The
-statement is then executed, and each row is retrieved from the database.
-Finally, the resources are freed.
-
-Problems may arise if the protocol is not followed correctly.
-A developer may, for example, accidentally close a statement whilst still
-retrieving rows, which would cause a runtime error. Similarly, a programmer may
-omit closing the statement or connection, which can lead to
-problems such as resource leaks in longer-running server applications.
-However, in conventional programming languages, there is no way to check
-automatically that a protocol is followed.
-
-In contrast, the use of \textit{dependent types} makes it possible
-to specify a program's behaviour precisely, and to check that a 
-specification is followed.
-%
-The difficulty is 
-that automatic verification by a compiler can be difficult or
-often impossible, requiring additional proofs to be given by the programmer.
-
-This complexity can be ameliorated through the use of \textit{embedded
-domain-specific languages} (EDSLs), which aim to abstract away the
-complexity of the underlying type system. EDSLs allow domain experts to
-focus on a problem domain, rather than the specific details of their
-solution.
-
-\idris{} \cite{brady2011idris} is a language with full dependent types, and
-extensive support for EDSLs through overloading and syntax macros. Through the
-use of \idris{}, and a framework for describing resource protocols using
-\emph{algebraic effects}~\cite{brady:effects}, we
-present a dependently-typed web framework, which allows the construction of
-programs with additional guarantees about correctness and security, whilst
-minimising the increase in development complexity.
-
-\subsection{Contributions}
-The primary contribution of this paper is the application of 
-dependent types may to provide strong static guarantees
-about the correctness and security of web applications, whilst minimising
-additional development complexity. In particular, we present:
-
-\begin{itemize}
-\item A form-handling mechanism, which preserves type information and
-manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection.
-
-\item Representations of CGI, Databases and sessions as
-\textit{algebraic effects}, allowing programs to be accepted only when they
-follow clearly defined resource usage protocols.
-
-\item A practical demonstration of the framework, a message board application.
-
-\end{itemize}
-
-We structure the remainder of this paper as follows.
-We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
-
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
-% =================================================
-% =================================================
-
-\section{An overview of the \texttt{Effects} framework}
-\label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access
-as \textit{algebraic effects}, allowing greater code modularity. This section
-is intended to serve as a brief introduction to the framework; for more
-information, see \cite{brady:effects}.
-
-An effect is represented as an instance of the \texttt{EFFECT} type. Such
-instances may be associated with a \textit{resource}, Describing a program in
-the \texttt{Effects} library is achieved using the following data type:
-
-\begin{Verbatim}
-EffM : (m : Type -> Type) ->
-       (es : List EFFECT) ->
-       (es’ : List EFFECT) ->
-       (a : Type) -> Type
-\end{Verbatim}
-\texttt{EffM} is a type that is parameterised over an execution context
-(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
-of effects at the end of the program's execution (\texttt{es'}), and a return
-type (\texttt{a}). The list of input types can be used to ensure that an
-operation may only be called when the program is in a particular state, and
-since the list of input effects may differ from the list of output effects, we
-may modify the resource associated with the state after each operation. This
-allows us to simply enforce a resource usage protocol, and deviation from this
-protocol will result in a compile-time error.
-
-Implementing a new effect involves creating a new algebraic data type, of kind
-\texttt{Effect}, which describes the available operations and how they modify
-the resource associated with the effect. Each operation consists of an initial
-resource \texttt{res}, the resource at the end of the operation \texttt{res'},
-and the output type of the operation \texttt{t}.
-
-%TODO: Execution
-\begin{Verbatim}
-Effect : Type
-Effect = (res : Type) -> (res’ : Type) ->
-         (t : Type) -> Type
-\end{Verbatim}
-
-Operations are made available to users by writing functions with the
-corresponding \texttt{EffM} type, which may then be used to invoke the
-operation from within an effectful program.
-
-\begin{Verbatim}
-queryPostVar : 
- String -> 
- Eff m [CGI (InitialisedCGI TaskRunning)] 
-   (Maybe String)
-queryPostVar query = (QueryPostVar query)
-\end{Verbatim}
-
-Additionally, we define \textit{handlers}, which describe how each operation is
-interpreted in a given execution context. Finally, \texttt{Effect} instances
-are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
-constructor, which is provided with the initial resource value.
-
-% =================================================
-% =================================================
-
 \section{Modelling resource usage protocols}
@@
 We represent an uninitialised CGI process as the unit type, ().
-\begin{Verbatim}[samepage]
-data CGIStep = Initialised 
-             | TaskRunning 
-             | TaskCompleted 
-             | HeadersWritten 
+
+\begin{SaveVerbatim}{cgistep}
+
+data CGIStep = Initialised   | TaskRunning 
+             | TaskCompleted | HeadersWritten 
              | ContentWritten
@@
      ICgi : CGIInfo -> InitialisedCGI s
-\end{Verbatim}
-Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
-Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.
 
-Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved.
-The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
-The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
+\end{SaveVerbatim}
+\useverb{cgistep}
 
-\begin{Verbatim}
+\noindent
+Figure~\ref{fig:cgistates} shows the states through which the CGI program
+progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
+as a resource-dependent algebraic effect. Each operation performed in an effectful
+program requires the resource to be of a certain type, and the completion of
+the operation may alter the type or value of the resource. The \texttt{Cgi}
+effect declaration shows these resource updates in the types of each operation,
+effectively specifying a state machine.
+
+Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
+that can be performed in this state is initialisation: by calling
+\texttt{initialise}, a CGIInfo record is populated, and the state transitions
+to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
+\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
+the initialised state.
+
+\begin{SaveVerbatim}{cgieff}
+{-                     { Input resource type }         { Output resource type }        { Value } -}
+
 data Cgi : Effect where
     Init         : Cgi ()                              (InitialisedCGI Initialised)    ()
-    StartRun : Cgi (InitialisedCGI Initialised) 
-                   (InitialisedCGI TaskRunning) ()
-    FinishRun : Cgi (InitialisedCGI TaskRunning) 
-                    (InitialisedCGI TaskCompleted) ()
-    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
-                       (InitialisedCGI HeadersWritten) ()
-    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
-                       (InitialisedCGI ContentWritten) ()
+    StartRun     : Cgi (InitialisedCGI Initialised)    (InitialisedCGI TaskRunning)    ()
+    FinishRun    : Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskCompleted)  ()
+    WriteHeaders : Cgi (InitialisedCGI TaskCompleted)  (InitialisedCGI HeadersWritten) ()
+    WriteContent : Cgi (InitialisedCGI HeadersWritten) (InitialisedCGI ContentWritten) ()
     OutputData   : String -> 
-                 Cgi (InitialisedCGI TaskRunning) 
-                     (InitialisedCGI TaskRunning) ()
-    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
-                               :: effs) -> 
-                       CGIProg effs a -> 
-                       Cgi (InitialisedCGI TaskRunning) 
-                           (InitialisedCGI TaskRunning) a
-\end{Verbatim}
-Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
+                   Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskRunning)    ()
+    RunAction    : Env IO (CGI (InitialisedCGI TaskRunning) :: effs) -> CGIProg effs a -> 
+                   Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskRunning)    a
+\end{SaveVerbatim}
 
-User code executes in the \texttt{TaskRunning} state.
-Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
-This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
+\begin{figure*}[t]
+\begin{center}
+\useverb{cgieff}
+\end{center}
+\caption{CGI Effect}
+\label{fig:cgieffect}
+\end{figure*}
 
-After the user code has finished execution, control returns to the library code.
-At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
-Finally, the headers and content are written which completes the process.
-Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
+Additional operations, including those to query POST and GET variables, are
+omitted in the interest of brevity.
 
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-\begin{Verbatim}
+User code executes in the \texttt{TaskRunning} state. Several operations, such
+as querying the POST and GET variables, are available in this state, alongside
+functions to output data to the web page and append data to the response
+headers. It is important to note, that at this stage nothing is written to the
+page, with the \texttt{output} and \texttt{addHeader} functions instead
+modifying the CGIInfo record. This data may then be printed at the end of the
+program's execution, in accordance with the resource usage protocol.
+
+After the user code has finished execution, control returns to the library
+code. At this point, the state transitions to TaskCompleted, and the headers
+are written.  Finally, the headers and content are written which completes the
+process. Since we parameterise the resource over a state, we may ensure that
+certain operations only happen in a particular prescribed order.
+
+In \idris{}, types are first-class, meaning that they may be treated like other
+terms in computations. We may therefore define the following type synonym,
+which we may use in order to make use of the CGI framework:
+
+\begin{SaveVerbatim}{cgiprog}
+
 CGIProg : List EFFECT -> Type -> Type
@@
   Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{Verbatim}
-This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
 
+\end{SaveVerbatim}
+\useverb{cgiprog}
+
+\noindent
+This is then passed, along with initial values for other effects that the user
+may wish to use, to the runAction function, which invokes the RunAction
+operation and executes the user-specified action.
+%
 A simple ``Hello, world!'' program would be defined as follows:
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{hworldcgi}
+
 module Main
@@
 main = runCGI [initCGIState] sayHello
-\end{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{hworldcgi}
 % =================================================
@@
 
+SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database
+engine often used as simple, structured storage for larger applications. We
+make use of SQLite due to its simplicity, although we envisage that these
+concepts would be applicable to more complex database management systems.
+
+The creation, preparation and execution of SQL queries has a specific usage
+protocol, with several possible points of failure. Failure is handled in
+traditional web applications by the generation of exceptions, which may be
+handled in the program.  Handling such exceptions is often optional, however,
+and in some cases unhandled errors may cause a deployed web application to
+display an error to the user. Such errors can be used to determine the
+structure of an insecure SQL query, and are often used by attackers to
+determine attack vectors for SQL injection attacks.
+
+Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
+access, which we have implemented for the SQLite library. This is encapsulated
+by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
+bind and retrieve data types other than \texttt{String}, in the interest of
+brevity).
+
+\begin{SaveVerbatim}{databaseeff}
+{-                           { Input resource type }    { Output resource type }             { Value }           -}
+data Sqlite : Effect where
+  OpenDB            : DBName -> 
+                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
+  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
+  PrepareStatement  : QueryString -> 
+                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
+                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
+  BindInt           : ArgPos -> Int -> 
+                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
+  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
+                                                          (SQLitePSSuccess Bound))           (Either BindError ())
+  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
+                                                          (SQLiteExecuting ValidRow)         StepResult
+  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
+                                                          (SQLiteExecuting ValidRow))        StepResult
+  GetColumnText     : Column -> 
+                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String
+
+\end{SaveVerbatim}
+
+\begin{figure*}[t]
+\begin{center}
+\useverb{databaseeff}
+\end{center}
+\caption{Database Effect}
+\label{fig:dbeffect}
+\end{figure*}
+
 %TODO: Fix up the labels
-\begin{figure}[htpb!]
+\begin{figure}[h]
 \centering
@@
   \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExecuting};
+  \node[state]         (F) [below of=E] {PreparedStatementExec};
 
@@
 
-SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
-We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
+\noindent
+There are three main phases involved in the usage of the SQLite protocol:
+connection to the database, preparation of a query, and execution of the query.
+We define several resources to encapsulate the state at any given point during the protocol.
 
-The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
-Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
-Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user.
-Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.
+We first define the \texttt{SQLiteConnected} resource, which signifies that a successful connection has been made to the database.
+This resource contains a pointer to the database structure which is used in further computations.
 
-Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
-This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.
+\begin{SaveVerbatim}{sqliteconnected}
 
-\begin{Verbatim}[samepage]
-data Sqlite : Effect where
-  OpenDB : 
-    String -> 
-    Sqlite () 
-           (SQLiteRes ConnectionOpened) Bool
-  CloseDB : 
-    Sqlite (SQLiteRes ConnectionOpened) () Bool
-  PrepareStatement : 
-    String -> 
-    Sqlite (SQLiteRes ConnectionOpened) 
-           (SQLiteRes PreparedStatementOpen) Bool
-  StartBind : 
-    Sqlite (SQLiteRes PreparedStatementOpen) 
-           (SQLiteRes PreparedStatementBinding) ()
-  BindText : 
-    ArgPos -> String -> Int -> 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBinding) 
-           Bool
-  FinishBind : 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBound) 
-           Bool
-  ExecuteStatement : 
-    Sqlite (SQLiteRes PreparedStatementBound) 
-           (SQLiteRes PreparedStatementExecuting) ()
-  Finalise : 
-    Sqlite (SQLiteRes PreparedStatementExecuting) 
-           (SQLiteRes ConnectionOpened) 
-           Bool
-  GetColumnText : 
-    Int -> 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
+data SQLiteConnected : Type where
+  SQLConnection : ConnectionPtr -> SQLiteConnected
+
+\end{SaveVerbatim}  
+\useverb{sqliteconnected}
+
+\noindent
+We secondly define the \texttt{SQLitePSSuccess}, \texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail} resources to represent the process of creating and binding variables to a prepared statement.
+
+\begin{SaveVerbatim}{sqliteconnected}
+
+data BindStep = Binding | Bound
+
+data SQLitePSSuccess : BindStep -> Type where
+  SQLitePS       : ConnectionPtr -> 
+                   StmtPtr -> SQLitePSSuccess a
+  SQLiteBindFail : ConnectionPtr -> 
+                   StmtPtr -> 
+                   BindError -> SQLitePSSuccess a
+
+data SQLitePSFail : Type where
+  PSFail : ConnectionPtr -> SQLitePSFail
+
+data SQLiteFinishBindFail : Type where
+  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
+                 SQLiteFinishBindFail
+
+\end{SaveVerbatim}
+\useverb{sqliteconnected}
+
+\noindent
+The \texttt{SQLitePSSuccess} resource indicates that a prepared statement has been correctly created by the underlying library, given a query string.
+The resource is parameterised by the \texttt{BindStep} type, which indicates whether the binding process has been completed.
+Should the creation of a prepared statement fail, the resource will be set to \texttt{SQLitePSFail}.
+
+To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor.
+Should any binding operations be called whilst in this state, the operations will not be passed to the underlying library.
+After the binding stage has completed, the \texttt{FinishBind} operation is called.
+This operation inspects the resource and either transitions to the \texttt{SQLitePSSuccess Bound} state, indicating that all bind operations were successful, or to the \texttt{SQLiteFinishBindFail} state, should a binding operation have been unsuccessful.
+
+SQLite operates by loading database rows into a buffer, the contents of which may then be accessed through several column access functions.
+After all rows returned by a query have been processed, no further calls to fetch more rows may be made.
+Additionally, no calls to column access functions may be made whilst there is no row being currently processed.
+
+Through the use of dependent types, we may encode these invariants statically, using the \texttt{SQLiteExecuting} resource.
+This is parameterised by the \texttt{ExecutionResult} data type, which indicates whether there is currently a valid row in the buffer.
+
+\begin{SaveVerbatim}{sqliteexecuting}
+
+data ExecutionResult = ValidRow
+                     | InvalidRow
+
+data SQLiteExecuting : ExecutionResult -> Type where
+  SQLiteE : ConnectionPtr -> 
+            StmtPtr -> SQLiteExecuting a
+
+\end{SaveVerbatim}
+\useverb{sqliteexecuting}
+
+Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.
+
+\begin{SaveVerbatim}{getcoltext}
+
+GetColumnText : Column -> 
+                Sqlite (SQLiteExecuting ValidRow) 
+                       (SQLiteExecuting ValidRow)
                        String
-  RowStep : 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           StepResult
-\end{Verbatim}
-There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
-\begin{Verbatim}
-data SQLiteRes : Step -> Type where
-  OpenConn : DBPointer -> SQLiteRes s
-  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
-  ExecutingStmt : DBPointer -> 
-                  StmtPointer -> 
-                  StepResult -> 
-                  SQLiteRes s
 
-data DBPointer = ValidConn Ptr
-               | InvalidConn
+\end{SaveVerbatim}
+\useverb{getcoltext}
 
-data StmtPointer = ValidStmt Ptr
-                 | InvalidStmt 
-\end{Verbatim}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
-At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
+\noindent
+In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct. The \texttt{nextRow} function 
+%If a failure happens at any point during the computation, the resource is
+%updated to reflect the failure: if, for example, the library failed to create a
+%connection to the database, the resource value would be updated to
+%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
+%requests are made to the underlying SQLite library, in order to ensure safety.
+%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
+%\texttt{executeFail} utility functions allow for failures, once detected, to be
+%handled by executing the appropriate sequence of state transition functions to
+%dispose of any open resources and return to the initial protocol state.
 
 %Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
-In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
-Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function.
-We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-\begin{Verbatim}
-data StepResult = Unstarted
-                | StepFail
-                | StepComplete
-                | NoMoreRows
-\end{Verbatim}
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
-Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
-We may therefore statically guarantee that only calls that will return a valid result are executed.
+%SQL queries are evaluated in SQLite upon a call to the C library function
+%\texttt{sqlite3\_step()}. In the case that a statement returns a result set,
+%each subsequent call retrieves another row for processing using a column access
+%function. Once all rows have been retrieved, the library returns
+%\texttt{SQLITE\_DONE}, meaning that no further calls should be made without
+%resetting the function. We encapsulate this requirement through the
+%\texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
 
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
+%\begin{SaveVerbatim}{stepres}
 
-\subsubsection{Example Code}
-Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
-Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
+%data StepResult = Unstarted    | StepFail
+%                | StepComplete | NoMoreRows
 
-We define a function of type:
-\begin{Verbatim}
+%\end{SaveVerbatim}
+%\useverb{stepres}
+
+\noindent
+Each call to \texttt{nextRow}, which is a wrapper around the underlying
+\texttt{sqlite3\_step()} C library function, returns a result of type
+\texttt{StepResult}, which is then reflected in the resource. Calls to
+\texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult}
+is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore
+statically guarantee that only calls that will return a valid result are
+executed.
+
+By incorporating pointers to open connections and prepared statements into the
+resource associated with the effect, we introduce a further layer of
+abstraction, in order to hide implementation details from the developer and
+encourage cleaner, less error-prone code.
+
+\subsubsection{Example}
+
+Programs making use of the DSL should look familiar to developers even without
+a background in functional programming. To demonstrate the functionality of the
+library, we return to the previous example of selecting the names and addresses
+of all staff working in a given department. Due to the fact that the
+\texttt{Effects} library overloads the bind operator, we may make use of
+\texttt{do}-notation, facilitating the usage of an imperative style.
+We define a function \texttt{textSel} of type:
+
+\noindent
+\begin{SaveVerbatim}{sqleffty}
+
 String ->
 Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{Verbatim}
-This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-\begin{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{sqleffty}
+
+\noindent
+This means that the program will be run in \texttt{IO}, and must
+start and end with no active resources. The return type indicates that either a
+list of \texttt{(String, String)} pairs, representing names and addresses in the
+database, or an error will be returned.
+%This example is shown in Figure
+%\ref{fig:testsel}.
+
+\begin{SaveVerbatim}{testsel}
 testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, Int)))
+             (Either String (List (String, String)))
 testSelect dept = do
@@
           return $ Left err 
-\end{Verbatim}
-The program initially attempts to open a connection to the \texttt{people.db} database.
-At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
-The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
+\end{SaveVerbatim}
 
-A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.
+\begin{figure}[h]
+\useverb{testsel}
+\caption{Example SQL program}
+\label{fig:testsel}
+\end{figure}
 
-\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database.
-This function is then recursively called until there are no more rows to process.
-\begin{Verbatim}
-collectResults : 
-  Eff IO 
-  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
+The program initially attempts to open a connection to the \texttt{people.db}
+database. At this point, since the \texttt{OpenDB} operation has been invoked,
+the program transitions to the \texttt{ConnectionOpened} state. The
+\texttt{openDB} function returns a Boolean value indicating whether or not the
+operation is successful. If not, then the \texttt{connFail} function is called
+to generate an appropriate error and dispose of the resources, as shown in
+~\ref{fig:sqlitestates}.
+
+A call to \texttt{prepareStatement} attempts to create a prepared statement,
+and a subsequent call to \texttt{beginExecution} allows data to be retrieved
+from the database.
+
+\texttt{CollectResults} makes a call to
+\texttt{nextRow} in order to make the next row of the result set available for
+processing, and uses \texttt{getColumnText} and \texttt{getColumnInt}
+to retrieve the data from the database. This function is then
+recursively called until there are no more rows to process.
+
+\begin{SaveVerbatim}{collectresty}
+
+collectResults : Eff IO 
+            [SQLITE (SQLiteRes PreparedStatementExec)] 
             (List (String, Int))
+\end{SaveVerbatim}
+
+\begin{SaveVerbatim}{collectres}
 collectResults = do
@@
       StepFail => return []
-\end{Verbatim}
-In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
-In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-\begin{Verbatim}
-data DBVal = DBInt Int
-           | DBText String
-           | DBFloat Float
-           | DBNull
-\end{Verbatim}
-We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-\begin{Verbatim}
+
+\end{SaveVerbatim}
+
+\noindent
+\useverb{collectresty}
+
+\noindent
+\useverb{collectres}
+
+In order to decrease unnecessary boilerplate code in user applications, we
+provide functions which abstract out unnecessary parts of this process. In
+order to do this, we define the algebraic data type \texttt{DBVal}, which is a
+tagged union over simple primitive types:
+
+\begin{SaveVerbatim}{dbvals}
+
+data DBVal = DBInt Int     | DBText String
+           | DBFloat Float | DBNull
+
+\end{SaveVerbatim}
+\useverb{dbvals}
+
+\noindent
+We also make use of the \texttt{ResultSet} type, which is a list of returned
+database rows.
+
+\begin{SaveVerbatim}{resultset}
+
 ResultSet : Type
 ResultSet = List (List DBVal)
-\end{Verbatim}
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-\begin{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{resultset}
+
+\noindent
+Using these, we may implement a function, \texttt{ExecuteSelect}, which, given
+a query, a list of variables to bind and their associated indices within the
+query, and a function which is used to extract information out of each returned
+database row, returns either a \texttt{ResultSet} or an error.
+
+\begin{SaveVerbatim}{execsel}
+
 executeSelect : 
   String -> String -> List (Int, DBVal) -> 
-  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-                 (List DBVal)) -> 
+  Eff m [SQLITE (SQLiteRes PreparedStatementExec)] 
+                 (List DBVal) -> 
   Eff m [SQLITE ()] (Either String ResultSet)
-\end{Verbatim}
+
+\end{SaveVerbatim}
+\useverb{execsel}
+
 % =================================================
 \subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate requests.
-This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
-In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.
+More complex web applications require some persistent state across separate
+requests. This is often done through an abstraction of a \textit{session},
+wherein a cookie is set on the remote host containing a unique session ID,
+which is in turn used to retrieve data. In this section, we describe the
+implementation of a simple session handler, and the resource usage protocol
+involved.
 
-A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects.
-By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.
+A major strength of the \texttt{Effects} library is that it allows for simple
+composition of individual, fine-grained effects. By combining the individual
+CGI and SQLite components, we may construct a simple session handler to provide
+a notion of state across separate web requests.
 
-We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
+We implement this by having a SQLite database, containing two tables:
+\texttt{session}, which stores session keys and their associated expiry dates,
+and \texttt{sessiondata}, which contains the data associated with each session.
+A datum associated with the session is described as a tagged union containing
+one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int},
+which is serialised alongside a type tag for storage in the database.
 %TODO: Fix up the labels
-\begin{Verbatim}[samepage]
+
+\begin{SaveVerbatim}{sessioneff}
+{-                        { Input resource type }            { Output resource type }   { Value }        -}
+
 data Session : Effect where
-  LoadSession : 
-  SessionID -> 
-  Session (SessionRes SessionUninitialised) 
-          (SessionRes SessionInitialised) 
-          (Maybe SessionData)
-  UpdateSession : 
-    SessionData -> 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) ()
-  CreateSession : 
-    SessionData -> 
-    Session (SessionRes SessionUninitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  DeleteSession : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool 
-  WriteToDB : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool
-  DiscardSessionChanges : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) ()
-  GetSessionID : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  GetSessionData : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionData)
-\end{Verbatim}
+  LoadSession           : SessionID -> 
+                          Session (SessionRes SessionClosed) (SessionRes SessionOpen)   (Maybe SessionData)
+  UpdateSession         : SessionData -> 
+                          Session (SessionRes SessionOpen)   (SessionRes SessionOpen)   ()
+  CreateSession         : SessionData -> 
+                          Session (SessionRes SessionClosed) (SessionRes SessionOpen)   (Maybe SessionID)
+  DeleteSession         : Session (SessionRes SessionOpen)   (SessionRes SessionClosed) Bool 
+  WriteToDB             : Session (SessionRes SessionOpen)   (SessionRes SessionClosed) Bool
+  DiscardSessionChanges : Session (SessionRes SessionOpen)   (SessionRes SessionClosed) ()
+  GetSessionID          : Session (SessionRes SessionOpen)   (SessionRes SessionOpen)   (Maybe SessionID)
+  GetSessionData        : Session (SessionRes SessionOpen)   (SessionRes SessionOpen)   (Maybe SessionData)
+\end{SaveVerbatim}
+
+\begin{figure*}[t]
+\begin{center}
+\useverb{sessioneff}
+\end{center}
+\caption{Session Effect}
+\label{fig:sessioneffect}
+\end{figure*}
+
 \begin{figure}[htpb!]
@@
 
-  \node[initial,state] (A)              {SessionUninitialised};
-  \node[state]         (B) [below of=A] {SessionInitialised};
+  \node[initial,state] (A)              {SessionClosed};
+  \node[state]         (B) [below of=A] {SessionOpen};
 
@@
 \end{figure}
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler.
-In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key.
-The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.
-This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.
 
-Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
-\begin{Verbatim}
-data SessionStep = SessionUninitialised
-                 | SessionInitialised
+Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
+with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
+algebraic effect. In this application, there exist two states:
+In \texttt{SessionClosed}, the user may load or create a
+session.
+In \texttt{SessionOpen}, the user may update the
+representation of the session in memory, serialise the session and write it to
+the database, or delete the session and invalidate the user's session key. The
+introduction of these two states ensures that changes are explicitly either
+written or discarded, eliminating the possibility of a developer updating the
+session but neglecting to commit it to persistent storage. This, of course, is
+under the assumption that the process exits cleanly: we attempt to facilitate
+this by writing total functions where possible.
 
+Much like the SQLite effect, we encapsulate failure by reflecting it in the
+resource associated with the effect.
+
+\begin{SaveVerbatim}{sstep}
+
+data SessionStep = SessionClosed | SessionOpen
+\end{SaveVerbatim}
+
+\begin{SaveVerbatim}{sres}
 data SessionRes : SessionStep -> Type where
@@
                    SessionRes s
-\end{Verbatim}
-The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
-If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.
+
+\end{SaveVerbatim}
+
+\useverb{sstep}
+
+\useverb{sres}
+
+The \texttt{SessionRes} data type is parameterised over the current state,
+which determines which operations may be performed, and has two constructors:
+\texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as
+creating a new session fails, no further side-effecting calls will be made, in
+turn preserving integrity.
 
 % =================================================
-
-\section{Type-aware form handling}
-\label{form}
-Programming web applications often involves processing user data, which may then be used in further effectful computations.
-Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.
-
-This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker.
-One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be used in a query such as:
-
-\begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + 
-       page + ', 5';
-\end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
-% Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here
-
-In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information.
-Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and echoes it back.
-Firstly, we define a form handler which echoes back a string provided by the form handler.
-It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.
-
-\begin{Verbatim}
-sayHello : Maybe String -> 
-           FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
-\end{Verbatim}
-
-We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
-
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [(handler args=[FormString], 
-                     effects=[CgiEffect], 
-                     fn=sayHello, 
-                     name="sayHello")
-           ]
-\end{Verbatim}
-
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
-
-\begin{Verbatim}
-showHelloForm : UserForm
-showHelloForm = do
-  addTextBox "Name" FormString Nothing
-  useEffects [CgiEffect]
-  addSubmit sayHello handlers
-\end{Verbatim}
-
-Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
-
-\begin{Verbatim}
-cgiHello : CGIProg [] ()
-cgiHello = do
-  handler_set <- isHandlerSet
-  if handler_set then do
-    handleForm handlers
-    return ()
-  else do
-    addForm "nameform" "helloform" showHelloForm
-    return ()
-
-main : IO ()
-main = runCGI [initCGIState] cgiHello
-\end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user.
-Upon submission of the form, the form handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
-\subsection{Form Construction}
-\label{formcons}
-Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-\begin{Verbatim}
-data FormTy = FormString
-            | FormInt
-            | FormBool
-            | FormFloat
-            | FormList FormTy 
-\end{Verbatim}
-Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-\begin{Verbatim}
-interpFormTy : FormTy -> Type
-interpFormTy FormString = String
-interpFormTy FormInt = Int
-interpFormTy FormBool = Bool
-interpFormTy FormFloat = Float
-interpFormTy (FormList a) = List (interpFormTy a)
-\end{Verbatim}
-%
-In order to specify a form, we once again use \texttt{Effects}.
-By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation.
-We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.
-
-\begin{Verbatim}
-using (G : List FormTy, E : List WebEffect)
-  data FormRes : List FormTy -> 
-                 List WebEffect -> Type where
-    FR : Nat -> 
-         List FormTy -> 
-         List WebEffect -> 
-         String -> 
-         FormRes G E
-
-  data Form : Effect where
-    AddTextBox : (label : String) -> 
-                 (fty : FormTy) -> 
-                 (Maybe (interpFormTy fty)) -> 
-                 Form (FormRes G E) 
-                      (FormRes (fty :: G) E) () 
-   ...
-
-    Submit : (mkHandlerFn ((reverse G), E)) ->
-             String -> 
-             Form (FormRes G E) (FormRes [] []) String
-\end{Verbatim}
-The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
-
-We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form.
-The resource allows us to record the types associated with each form element and the HTML required to display the form.
-he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
-
-By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
-It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
-
-By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource.
-The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
-
-To specify a form instance, we define a function of type \texttt{UserForm}:
-\begin{Verbatim}
-  UserForm : Type
-  UserForm = EffM m [FORM (FormRes []) 
-                          (FormRes [])] String
-\end{Verbatim}
-All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
-As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.
-
-Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
-Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
-If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
-
-Before associating a handler function with the form, we must specify the effects available to the handler.
-This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
-By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-\begin{Verbatim}
-  useEffects : (effs : List WebEffect) ->
-               EffM m [FORM (FormRes G E)] 
-                      [FORM (FormRes G effs)] ()
-  useEffects effs = (UseEffects effs)
-\end{Verbatim}
-Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
-%
-We may specify a handler function of type \texttt{FormHandler}:
-\begin{Verbatim}
-  FormHandler : List EFFECT -> Type
-  FormHandler effs = Eff IO effs ()
-\end{Verbatim}
-In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-\begin{Verbatim}
-  addSubmit : 
-    (f :  mkHandlerFn ((reverse G), E)) ->
-    (fns : HandlerList) ->
-    {default tactics 
-      { applyTactic findFn 100; solve; }
-      prf : FnElem f fns} ->
-    EffM m [FORM (FormRes G E)]
-           [FORM (FormRes [] [])] 
-           String
-  addSubmit f handlers {prf} = (Submit f name)
-    where name : String
-          name = getString' f handlers prf          
-\end{Verbatim}
-
-Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-\begin{Verbatim}
-MkHandlerFnTy : Type
-MkHandlerFnTy = (List FormTy, List WebEffect)
-
-mkHandlerFn' : List FormTy -> List WebEffect -> Type
-mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
-mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
-                              mkHandlerFn' xs effs
-
-mkHandlerFn : MkHandlerFnTy -> Type 
-mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{Verbatim}
-The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
-When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-
-To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{Verbatim}
-HandlerFn : Type
-HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{Verbatim}
-%
-The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
-We also store a unique string identifier, which is used to serialise a reference to the handler function.
-
-In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
-\begin{Verbatim}
-syntax 
-  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
-  ", name=" [name] = ((args, effs) ** (fn, name))
-\end{Verbatim}
-We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
-This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
-\begin{Verbatim}
-handler args=[FormString], 
-        effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
-\end{Verbatim}
-
-We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-\begin{Verbatim}
-HandlerList : Type
-HandlerList = List HandlerFn
-\end{Verbatim}
-To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-\begin{Verbatim}
-  data FnElem : mkHandlerFn ((reverse G), E) -> 
-                HandlerList -> Type where
-
-       FnHere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
-       FnThere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-               FnElem f xs -> FnElem f (x :: xs)
-\end{Verbatim}
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
-\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
-%TODO: this should probably be rewritten
-We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
-The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail.
-
-Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.
-
-Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this.
-Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.
-
-Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a hidden HTML field.
-
-\subsection{Form Handling}
-\label{formhandling}
-Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
-
-With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
-We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
-
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
-This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
-If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
-If not, the computation is abandoned and an error is shown.
-% I gave it a shot, right?
-:P
-\begin{Verbatim}
-checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
-                 (frm_fn_ty : MkHandlerFnTy) -> 
-                 mkHandlerFn reg_fn_ty -> 
-                 Maybe (mkHandlerFn frm_fn_ty)
-checkFunctions reg_ty frm_ty reg_fn with 
-                             (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
-                                            Just reg_fn
-  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{Verbatim}
-We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
-Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
-
-%-----------------------------
-%-----------------------------
-
-\section{Extended Example: Message Board}
-\label{messageboard}
-Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
-
-Firstly, we create a database schema in which to record information stored by the message board.
-We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
-
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
-Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
-\subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-\begin{Verbatim}
-main : IO ()
-main = do (runCGI [initCGIState, 
-                  InvalidSession, ()] 
-           handleRequest)
-          pure ()
-\end{Verbatim}
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-\begin{Verbatim}
-handleRequest : CGIProg [SESSION 
-  (SessionRes SessionUninitialised), 
-  SQLITE ()] ()
-handleRequest = do 
-  handler_set <- isHandlerSet
-  if handler_set then do
-    lift' (handleForm handlers)
-    Effects.pure ()
-  else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
-    handleNonFormRequest action (map strToInt thread_id)
-\end{Verbatim}
-\subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
-One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-\begin{Verbatim}
-newThreadForm : UserForm
-newThreadForm = do
-  addTextBox "Title" FormString Nothing
-  addTextBox "Post Content" FormString Nothing 
-  useEffects [CgiEffect, SessionEffect, SqliteEffect]
-  addSubmit handleNewThread handlers
-\end{Verbatim}
-The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
-Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
-The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleRegisterForm, 
-          name="handleRegisterForm"),
-
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleNewThread, 
-          name="handleNewThread"),
-     ...
-  ]
-\end{Verbatim}
-Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
-In order to do this, we make use of the session handler.
-We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
-If not, then a failure function is called instead.
-Should the form handler function be called with invalid arguments, an error is shown.
-\begin{Verbatim}
-handleNewThread : 
-  Maybe String -> Maybe String -> 
-  FormHandler [CGI (InitialisedCGI TaskRunning), 
-               SESSION (SessionRes SessionUninitialised), 
-               SQLITE ()]  
-handleNewThread (Just title) (Just content) = do 
-  withSession (addNewThread title content) notLoggedIn
-  pure ()
-handleNewThread _ _ = do 
-  outputWithPreamble "<h1>Error</h1><br />There was 
-       an error posting your thread."
-  pure ()
-\end{Verbatim}
-Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
-The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
-This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
-In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
-In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-\begin{Verbatim}
-threadInsert : Int -> 
-               String -> 
-               String -> 
-               Eff IO [SQLITE ()] Bool
-threadInsert uid title content = do
-  let query = "INSERT INTO `Threads` 
-    (`UserID`, `Title`) VALUES (?, ?)"
-  insert_res <- (executeInsert DB_NAME query 
-    [(1, DBInt uid), (2, DBText title)]
-  case insert_res of
-    Left err => pure False
-    Right thread_id => postInsert uid thread_id content
-\end{Verbatim}
-\subsection{Listing Threads}
-Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-\begin{Verbatim}
-getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = 
-  executeSelect DB_NAME query [] collectThreadResults
- where query = "SELECT `ThreadID`, `Title`, `UserID`, 
-   `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{Verbatim}
-Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
-Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-\begin{Verbatim}
-traverseThreads : ResultSet -> 
-  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = pure ()
-traverseThreads (x::xs) = do traverseRow x
-                             traverseThreads xs
-  where traverseRow : List DBVal -> 
-           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-        traverseRow ((DBInt thread_id)::
-                     (DBText title)::
-                     (DBInt user_id)::
-                     (DBText username)::[]) =
-           (output $ "<tr><td>
-            <a href=\"?action=showthread&thread_id=" ++ 
-            (show thread_id) ++ "\">" ++ 
-            title ++ "</a></td><td>" ++ 
-            username ++ "</td></tr>") 
-        traverseRow _ = pure ()
-\end{Verbatim}
-\subsection{Authentication}
-Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
-If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
-The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
-Any failures, for example with creating a new session or querying the database, are reported to the user.
-\begin{Verbatim}
-handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
-  case auth_res of
-    Right (Just uid) => do
-      set_sess_res <- setSession uid
-      if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
-        pure ()
-      else do
-        lift' (output "Could not set session")
-        pure ()
-    Right Nothing => do
-      lift' (output "Invalid username or password")
-      pure ()
-    Left err => do
-      lift' (output $ "Error: " ++ err)
-      pure ()
-\end{Verbatim}
-Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
-
-Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
-Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
-Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
-%-----------------------------
-%-----------------------------
-
-\section{Related Work}
-\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser.
-The library also provides support for cookies and basic form handling.
-
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods.
-In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs.
-Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs.
-Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations.
-Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.
-
-\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction.
-Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven.
-This means that upon a component being edited, a computation occurs, given the state of the current form.
-This is then saved to be used in future computations.
-Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted.
-This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.
-
-The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another.
-This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion.
-These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application.
-Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.
-
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs.
-Through the use of the Ur\/Web DSL, developers may generate well-typed code from records.
-By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure.
-Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.
-
-WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres.
-WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes.
-The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables.
-These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy.
-Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing.
-WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation.
-We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.
-
-%-----------------------------
-%-----------------------------
-
-\section{Conclusions}
-Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
-
-Our framework, through the use of dependent types, provides several static guarantees.
-Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
-This immediately eliminates the possibilities of SQL injection attacks on non-string types.
-Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
-Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
-
-Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
-Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
-By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
-
-In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
-With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
-
-Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
-\subsection{Further Work}
-We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
-
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
-
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
-We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
-
-Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
-Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
-Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
-
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
-Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
-This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
-Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
-Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
-
-Dependently-typed languages provide great promise for the construction of secure and correct programs.
-Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
-% Cryptography bindings would be really good.
-% Integration with a web server instead of doing everything over CGI
-% Improvements to form handling system
-% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%-----------------------------
-%-----------------------------
-
-\acks
-This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
-The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
-% SICSA / EPSRC (grant number?
-% #idris
-% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{plainnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{refs}
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-11-11 paper/conclusion.tex

--- 2013-08-21 paper/main.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-%\usepackage{silence}
-\usepackage{pgf}
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata}
-\usepackage{url}
-\usepackage[hidelinks]{hyperref}
-\usepackage{fancyvrb}
-\usepackage[numbers]{natbib}
-%\WarningsOff
-
-\begin{document}
-\fvset{fontsize=\small}
-\newcommand{\idris}{\textsc{Idris}}
-\newcommand{\idata}{\textsf{iData}}
-\newcommand{\itasks}{\textsf{iTasks}}
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
-\copyrightyear{2013} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{}        % These are ignored unless
-\preprintfooter{}   % 'preprint' option specified.
-
-\title{Practical Dependently-Typed Programming for the Web}
-\subtitle{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
-%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
-
-\authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
-           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
-
-\maketitle
-
-\begin{abstract}
-Dependently-typed languages allow very expressive types to be used during development, in turn facilitating easier reasoning about the operation of programs written in such languages.
-Stronger type specifications do however bring with them the disadvantage
-that it becomes increasingly difficult to write programs that are accepted by
-the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in high-level languages which use dependent
-types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an
-EDSL to facilitate the creation and handling of web forms which retain type
-information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how to enforce resource usage
-protocols associated with common web operations such as CGI, database access
-and session handling.
-
-\end{abstract}
-
-%\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-%\keywords
-%keyword1, keyword2
-
-\section{Introduction}
-
-Web applications, whilst ubiquitous, are also prone to incorrect construction
-and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
-scripting \cite{owasp:xss}. Security breaches using such exploits are
-far-reaching, and high profile cases involve large corporations such as Sony,
-who suffered a well-publicised and extremely costly SQL Injection breach in
-2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
-\cite{imperva:yahoo}.
-
-Many web applications are written in dynamically-checked scripting languages
-such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. A significant drawback, however, is that such languages
-do not provide  the same static guarantees about runtime behaviour afforded by
-programs with more expressive, static type systems, instead relying on
-extensive unit testing to ensure correctness and security.
-
-%%------- EXAMPLE
-Let us consider a simple database access routine, written in
-PHP, where we wish to obtain the name and address of every employee working in
-a given department, \texttt{\$dept}. We firstly construct an object
-representing a database connection, where the arguments are the database host,
-user, password and name respectively:
-
-%
-\begin{Verbatim}
-  $conn = new mysqli("localhost", "username", 
-                     "password", "db");
-\end{Verbatim}
-%
-We then check to see if the connection was successful, and exit
-if not.  This check is optional, so
-it would be possible to omit it. However, this would cause
-problems with later steps.
-
-%
-\begin{Verbatim}
-  if (mysqli_connect_errno()) {
-      exit();
-  }
-\end{Verbatim}
-%
-We then create a prepared statement detailing our query, and bind the
-\texttt{`dept'} value:
-
-\begin{Verbatim}
-  $stmt = $conn->prepare("SELECT `name`, `address` 
-     FROM `staff` WHERE `dept` = ?);
-  $stmt->bind_param('s', $dept);
-\end{Verbatim}
-%
-After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. 
-Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.
-
-%
-\begin{Verbatim}
-  $stmt->execute();
-  $stmt->bind_result($name, $address);
-  while ($stmt->fetch()) {
-    printf("Name: %s, Age: %s", $name, $age);
-  }
-\end{Verbatim}
-%
-Finally, once the statement and connection are no longer needed, they should be
-closed in order to discard the associated resources:
-
-%
-\begin{Verbatim}
-  $stmt->close();
-  $conn->close();
-\end{Verbatim}
-%
-Even in this small example, there exists a specific resource usage protocol.
-Firstly, a connection to the database must be opened. The object-oriented style
-used in the example encapsulates this to an extent, as the object must be
-created in order for operations to be performed, however it is less obvious in
-a procedural version of the code. Secondly, a prepared statement is created,
-using the raw SQL and placeholders to which variables are later bound. The
-statement is then executed, and each row is retrieved from the database.
-Finally, the resources are freed.
-
-Problems may arise if the protocol is not followed correctly.
-A developer may, for example, accidentally close a statement whilst still
-retrieving rows, which would cause a runtime error. Similarly, a programmer may
-omit closing the statement or connection, which can lead to
-problems such as resource leaks in longer-running server applications.
-However, in conventional programming languages, there is no way to check
-automatically that a protocol is followed.
-
-In contrast, the use of \textit{dependent types} makes it possible
-to specify a program's behaviour precisely, and to check that a 
-specification is followed.
-%
-The difficulty is 
-that automatic verification by a compiler can be difficult or
-often impossible, requiring additional proofs to be given by the programmer.
-
-This complexity can be ameliorated through the use of \textit{embedded
-domain-specific languages} (EDSLs), which aim to abstract away the
-complexity of the underlying type system. EDSLs allow domain experts to
-write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.
-
-\idris{} \cite{brady2011idris} is a language with full dependent types, and
-extensive support for EDSLs through overloading and syntax macros. Through the
-use of \idris{}, and a framework for describing resource protocols using
-\emph{algebraic effects}~\cite{brady:effects}, we
-present a dependently-typed web framework, which allows the construction of
-programs with additional guarantees about correctness and security, whilst
-minimising the increase in development complexity.
-
-\subsection{Contributions}
-The primary contribution of this paper is the application of 
-dependent types to provide strong static guarantees
-about the correctness and security of web applications, whilst minimising
-additional development complexity. In particular, we present:
-
-\begin{itemize}
-\item A form-handling mechanism, which preserves type information and
-manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection and cross-site scripting.
-
-\item Representations of CGI, Databases and sessions as
-\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
-follow clearly defined resource usage protocols.
-
-\item A message board application, demonstrating the usage of the framework.
-
-\end{itemize}
-
-We structure the remainder of this paper as follows. We provide a brief
-overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
-how this may be used to ensure adherence to resource usage protocols for CGI,
-SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
-type-safe form handling in Section ~\ref{form}, implemented using
-\texttt{Effects}; and discuss the larger example
-of a message board system making use of these components in Section
-~\ref{messageboard}.
-
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
-% =================================================
-% =================================================
-
-\section{An overview of the \texttt{Effects} framework}
-\label{effects}
-
-\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
-side-effects such as state, exceptions, and I/O as \emph{algebraic
-effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
-by an input and output \emph{state}, which permits effectful programs to track
-the progress of a resource usage protocol. Effectful programs are written
-in a monadic style, with \texttt{do}-notation, with their type stating which
-specific effects are available.
-Effectful
-programs are described using the following data type,
-in the simplest case:
-
-\begin{Verbatim}
-Eff : (m : Type -> Type) ->
-      (es : List EFFECT) -> (a : Type) -> Type
-\end{Verbatim}
-
-\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
-which describes the context in which the effectful program will be run, a
-list of side effects \texttt{es} that the program is permitted to use, and the
-program’s return type \texttt{a}. The name \texttt{m} for the computation context is
-suggestive of a monad, but there is no requirement for it to be so.
-
-For example, the following type carries an integer state,
-throws an exception of type \texttt{String} if the state reaches 100, 
-and runs in a \texttt{Maybe} context:
-
-\begin{Verbatim}
-addState : Eff Maybe [STATE Int, EXCEPTION String] ()
-addState = do val <- get
-              when (val == 100) (raise "State too big")
-              put (val + 1)
-\end{Verbatim}
-
-\subsection{Implementing Effects}
-
-Effects such as state and exception are described as algebraic data types,
-and run by giving \emph{handlers} for specific computation contexts. 
-Effects have a corresponding \emph{resource} (in the case of state, the
-resource is simply the current state). Executing an effectful operation may
-change the resource and return a value:
-
-\begin{Verbatim}
-Effect : Type
-Effect = (in_res : Type) -> (out_res : Type) -> 
-         (val : Type) -> Type
-\end{Verbatim}
-
-For example, the state effect is described as follows:
-
-\begin{Verbatim}
-data State : Effect where
-     Get :      State a a a
-     Put : b -> State a b ()
-\end{Verbatim}
-
-That is, \texttt{Get} returns a value of type \texttt{a} without updating
-the resource type. \texttt{Put} returns nothing, but has the effect of updating
-the resource. In order to make an effect usable, we implement a handler
-for a computation context by making an instance of the following class:
-
-\begin{Verbatim}
-class Handler (e : Effect) (m : Type -> Type) where
-     handle : res -> 
-              (eff : e res res' t) -> 
-              (k: res' -> t -> m a) -> m a
-\end{Verbatim}
-
-The \texttt{handle} function takes the input resource, an effect which may
-update that resource and execute a side-effect, and a continuation \texttt{k}
-which takes the updated resource and the return value of the effect. We use
-a continuation here primarily because there is no restriction on the number of
-times a handler may invoke the continuation (raising an exception, for example,
-will not call the continuation). Reading and updating states in handled
-for all computation contexts \texttt{m} as follows:
-
-\begin{Verbatim}
-instance Handler State m where
-     handle st Get     k = k st st
-     handle st (Put n) k = k n ()
-\end{Verbatim}
-
-Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
-the \texttt{Get} and \texttt{Put} operations into functions in \texttt{Eff}, as
-follows:
-
-\begin{Verbatim}
-STATE : Type -> EFFECT
-STATE t = MkEff t State
-
-get : Eff m [STATE x] x
-get = Get
-
-put : x -> Eff m [STATE x] ()
-put val = Put val
-\end{Verbatim}
-
-A concrete effect is simply an algebraic effect type paired with its current
-resource type. This, and other
-technical details, are explained in full elsewhere~\cite{brady:effects}.
-For the purposes of this paper, it suffices to know how to describe and
-handle new algebraic effects.
-
-\subsection{Resource Protocols as Effects}
-
-More generally, a program might modify the set of effects available.
-This might be desirable for several reasons, such as adding a new
-effect, or to update an index of a dependently typed state. In this
-case, we describe programs using the \texttt{EffM} data type:
-
-\begin{Verbatim}
-EffM : (m : Type -> Type) ->
-       (es : List EFFECT) -> (es' : List EFFECT) ->
-       (a : Type) -> Type
-\end{Verbatim}
-
-\texttt{EffM} is parameterised over the context and type as before, but
-separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
-In fact, \texttt{Eff}
-is defined in terms of \texttt{EffM}, with equal input/output effects.
-We can use this to describe how effects follow resource protocols. A simple
-example is a file access protocol, where a file must be opened before it
-is read or written, and a file must be closed on exit:
-
-\begin{Verbatim}
-data FileIO : Effect where
-     Open  : String -> (m : Mode) -> 
-             FileIO () (OpenFile m) ()
-     Close : FileIO (OpenFile m) () ()
-
-     ReadLine  : FileIO (OpenFile Read)  
-                        (OpenFile Read) String
-     WriteLine : String -> 
-                 FileIO (OpenFile Write) 
-                        (OpenFile Write) ()
-     EOF       : FileIO (OpenFile Read)  
-                        (OpenFile Read) Bool
-\end{Verbatim}
-
-Note that the types of the input and output resources describes how resource
-state changes in each operation: opening a file changes an empty resource to
-a resource containing an open file; reading a line is only possible if the
-resoource is a file open for reading, etc.
-The handler for this effect for an \texttt{IO} computation context will
-execute the required primitive I/O actions, as well as throwing an exception
-if any operation fails.
-
-The following program type checks, and therefore implicitly carries 
-a proof that the file resource protocol is followed correctly:
-
-\begin{Verbatim}
-testFile : Eff IO [FILE_IO (), STDIO] () 
-testFile = catch (do open "testFile" Read
-                     str <- readLine
-                     close
-                     putStrLn str)
-                 (\err => putStrLn ("Error: " ++ show err))
-\end{Verbatim}
-
-The type of \texttt{testFile} states
-that File I/O and console I/O are available effects, and in particular that
-the resource associated with the File I/O will be in the same state on entry
-and exit.
-Therefore, attempting to write to the file, or failing to open or close the
-file, would cause a \emph{compile-time} error.
-
-We will use this technique extensively
-throughout this paper: describe a resource usage protocol in terms of
-state transitions; implement an effect which captures that protocol; implement
-programs which, by using this effect, implicitly carry a proof that the resource
-protocol has been correctly followed.
-
-% =================================================
-% =================================================
-
-\section{Modelling resource usage protocols}
-\label{rup}
-In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
-
-% =================================================
-
-\subsection{CGI}
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {Initialised};
-  \node[state]         (C) [below of=B] {TaskRunning};
-  \node[state]         (D) [below of=C] {TaskCompleted};
-  \node[state]         (E) [below of=D] {HeadersWritten};
-  \node[state]         (F) [below of=E] {ContentWritten};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
-\end{tikzpicture}
-}
-\caption{CGI States}
-\label{fig:cgistates}
-\end{figure}
-
-CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error being shown.
-
-A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
-Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
-If this were to happen, then the program would fail due to the fact that headers had not been written to the client.
-
-By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
-In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
-This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI process as the unit type, ().
-\begin{Verbatim}[samepage]
-data CGIStep = Initialised 
-             | TaskRunning 
-             | TaskCompleted 
-             | HeadersWritten 
-             | ContentWritten
-
-data InitialisedCGI : CGIStep -> Type where
-  ICgi : CGIInfo -> InitialisedCGI s
-\end{Verbatim}
-Figure~\ref{fig:cgistates} shows the states through which the CGI program
-progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
-as a resource-dependent algebraic effect. Each operation performed in an effectful
-program requires the resource to be of a certain type, and the completion of
-the operation may alter the type or value of the resource.
-
-Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
-that can be performed in this state is initialisation: by calling
-\texttt{initialise}, a CGIInfo record is populated, and the state transitions
-to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
-\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
-the initialised state.
-
-\begin{figure}[h]
-\begin{Verbatim}
-data Cgi : Effect where
-    Init : Cgi () (InitialisedCGI Initialised) ()
-    StartRun : Cgi (InitialisedCGI Initialised) 
-                   (InitialisedCGI TaskRunning) ()
-    FinishRun : Cgi (InitialisedCGI TaskRunning) 
-                    (InitialisedCGI TaskCompleted) ()
-    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
-                       (InitialisedCGI HeadersWritten) ()
-    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
-                       (InitialisedCGI ContentWritten) ()
-    OutputData : String -> 
-                 Cgi (InitialisedCGI TaskRunning) 
-                     (InitialisedCGI TaskRunning) ()
-    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
-                               :: effs) -> 
-                       CGIProg effs a -> 
-                       Cgi (InitialisedCGI TaskRunning) 
-                           (InitialisedCGI TaskRunning) a
-\end{Verbatim}
-\caption{CGI Effect}
-\label{fig:cgieffect}
-\end{figure}
-
-Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
-
-User code executes in the \texttt{TaskRunning} state.
-Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
-This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
-
-After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, and the headers are written.
-Finally, the headers and content are written which completes the process.
-Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
-
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-\begin{Verbatim}
-CGIProg : List EFFECT -> Type -> Type
-CGIProg effs a = 
- Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{Verbatim}
-This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
-
-A simple ``Hello, world!'' program would be defined as follows:
-
-\begin{Verbatim}
-module Main
-import Cgi
-
-sayHello : CGIProg [] ()
-sayHello = output "Hello, world!"
-
-main : IO ()
-main = runCGI [initCGIState] sayHello
-\end{Verbatim}
-% =================================================
-
-\subsection{Database access with SQLite}
-
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {ConnectionOpened};
-  \node[state]         (C) [below of=B] {PreparedStatementOpen};
-  \node[state]         (D) [below of=C] {PreparedStatementBinding};
-  \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExecuting};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
-        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
-\end{tikzpicture}
-}
-\caption{Database Resource Usage Protocol}
-\label{fig:sqlitestates}
-\end{figure}
-
-SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
-We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-
-The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
-Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
-Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user.
-Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.
-
-Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
-access, which we have implemented for the SQLite library. This is encapsulated
-by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
-bind and retrieve data types other than \texttt{String}, in the interest of
-brevity).
-
-\begin{figure}[h]
-\begin{Verbatim}[samepage]
-data Sqlite : Effect where
-  OpenDB : 
-    String -> 
-    Sqlite () 
-           (SQLiteRes ConnectionOpened) Bool
-  CloseDB : 
-    Sqlite (SQLiteRes ConnectionOpened) () Bool
-  PrepareStatement : 
-    String -> 
-    Sqlite (SQLiteRes ConnectionOpened) 
-           (SQLiteRes PreparedStatementOpen) Bool
-  StartBind : 
-    Sqlite (SQLiteRes PreparedStatementOpen) 
-           (SQLiteRes PreparedStatementBinding) ()
-  BindText : 
-    ArgPos -> String -> Int -> 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBinding) 
-           Bool
-  FinishBind : 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBound) 
-           Bool
-  ExecuteStatement : 
-    Sqlite (SQLiteRes PreparedStatementBound) 
-           (SQLiteRes PreparedStatementExecuting) ()
-  Finalise : 
-    Sqlite (SQLiteRes PreparedStatementExecuting) 
-           (SQLiteRes ConnectionOpened) 
-           Bool
-  GetColumnText : 
-    Int -> 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           String
-  RowStep : 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           StepResult
-\end{Verbatim}
-\caption{Database Effect}
-\label{fig:dbeffect}
-\end{figure}
-
-There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
-\begin{Verbatim}
-data SQLiteRes : Step -> Type where
-  OpenConn : DBPointer -> SQLiteRes s
-  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
-  ExecutingStmt : DBPointer -> 
-                  StmtPointer -> 
-                  StepResult -> 
-                  SQLiteRes s
-
-data DBPointer = ValidConn Ptr
-               | InvalidConn
-
-data StmtPointer = ValidStmt Ptr
-                 | InvalidStmt 
-\end{Verbatim}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
-At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
-
-%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
-In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
-Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function.
-We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-\begin{Verbatim}
-data StepResult = Unstarted
-                | StepFail
-                | StepComplete
-                | NoMoreRows
-\end{Verbatim}
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
-Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
-We may therefore statically guarantee that only calls that will return a valid result are executed.
-
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
-
-\subsubsection{Example Code}
-Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
-Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
-
-We define a function of type:
-\begin{Verbatim}
-String ->
-Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{Verbatim}
-This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-\begin{Verbatim}
-testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, Int)))
-testSelect dept = do
-  open_db <- openDB "people.db"
-  if open_db then do
-    let sql = "SELECT name, address FROM `staff` 
-                    WHERE dept = ?;"
-    sql_prep_res <- prepareStatement sql
-    if sql_prep_res then do 
-      startBind
-      bindText 1 dept
-      finishBind
-      beginExecution
-      results <- collectResults
-      finaliseStatement
-      closeDB
-      return $ Right results
-    else do err <- stmtFail
-            return $ Left err
-  else do err <- connFail
-          return $ Left err 
-\end{Verbatim}
-The program initially attempts to open a connection to the \texttt{people.db} database.
-At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
-The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
-
-A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.
-
-\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database.
-This function is then recursively called until there are no more rows to process.
-\begin{Verbatim}
-collectResults : 
-  Eff IO 
-  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-  (List (String, Int))
-collectResults = do
-  step_result <- nextRow
-  case step_result of
-      StepComplete => do name <- getColumnText 1
-                         age <- getColumnInt 2
-                         xs <- collectResults
-                         return $ (name, age) :: xs
-      NoMoreRows => return []
-      StepFail => return [] 
-\end{Verbatim}
-In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
-In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-\begin{Verbatim}
-data DBVal = DBInt Int
-           | DBText String
-           | DBFloat Float
-           | DBNull
-\end{Verbatim}
-We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-\begin{Verbatim}
-  ResultSet : Type
-  ResultSet = List (List DBVal)
-\end{Verbatim}
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-\begin{Verbatim}
-executeSelect : 
-  String -> String -> List (Int, DBVal) -> 
-  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-                 (List DBVal)) -> 
-  Eff m [SQLITE ()] (Either String ResultSet)
-\end{Verbatim}
-% =================================================
-\subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate requests.
-This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
-In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.
-
-A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects.
-By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.
-
-We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
-%TODO: Fix up the labels
-
-\begin{figure}[h]
-\begin{Verbatim}[samepage]
-data Session : Effect where
-  LoadSession : 
-  SessionID -> 
-  Session (SessionRes SessionUninitialised) 
-          (SessionRes SessionInitialised) 
-          (Maybe SessionData)
-  UpdateSession : 
-    SessionData -> 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) ()
-  CreateSession : 
-    SessionData -> 
-    Session (SessionRes SessionUninitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  DeleteSession : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool 
-  WriteToDB : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool
-  DiscardSessionChanges : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) ()
-  GetSessionID : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  GetSessionData : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionData)
-\end{Verbatim}
-\caption{Session Effect}
-\label{fig:sessioneffect}
-\end{figure}
-
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {SessionUninitialised};
-  \node[state]         (B) [below of=A] {SessionInitialised};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
-\end{tikzpicture}
-}
-\caption{Session Handler Resource Usage Protocol}
-\label{fig:sessionstates}
-\end{figure}
-
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
-with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
-algebraic effect. In this application, there exist two states:
-In \texttt{SessionUninitialised}, the user may load or create a
-session.
-In \texttt{SessionInitialised}, the user may update the
-representation of the session in memory, serialise the session and write it to
-the database, or delete the session and invalidate the user's session key. The
-introduction of these two states ensures that changes are explicitly either
-written or discarded, eliminating the possibility of a developer updating the
-session but neglecting to commit it to persistent storage. This, of course, is
-under the assumption that the process exits cleanly: we attempt to facilitate
-this by writing total functions where possible.
-
-Much like the SQLite effect, we encapsulate failure by reflecting it in the
-resource associated with the effect.
-
-\begin{Verbatim}
-data SessionStep = SessionUninitialised
-                 | SessionInitialised
-
-data SessionRes : SessionStep -> Type where
-  InvalidSession : SessionRes s  
-  ValidSession   : SessionID -> 
-                   SessionData -> 
-                   SessionRes s
-\end{Verbatim}
-The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
-If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.
-
-% =================================================
-
-\section{Type-aware form handling}
-\label{form}
-Programming web applications often involves processing user data, which may
-then be used in further effectful computations. Data submitted using a form is
-transmitted over the internet as a string as part of an HTTP request, which
-traditionally involves losing associated type information.
-
-This can in turn lead to risks; developers may assume that data is
-of a certain type, and therefore discount the possibility that it may have been
-modified by an attacker. One example would be the traversal of paginated data,
-in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be
-used in a query such as:
-
-\begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + 
-       page + ', 5';
-\end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be
-modified by an attacker to include a malicious string which would alter the
-semantics of the query, allowing an attacker to execute a blind SQL injection
-attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT
-% clauses here
-
-In this section, we present a mechanism by which we introduce a DSL
-for the creation of web forms which preserve type information, implemented
-as a dependent algebraic effect. Once the form has
-been submitted, retrieved information is passed directly to a
-developer-specified function for handling, without the need to manually check
-and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and
-echoes it back. Firstly, we define a form handler which echoes back a string
-provided by the form handler. It has one argument of type \texttt{Maybe
-String}, which accounts for the possibility that the user may have specified
-incorrect data within the form.
-
-\begin{Verbatim}
-sayHello : Maybe String -> 
-           FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
-\end{Verbatim}
-
-We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
-
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [(handler args=[FormString], 
-                     effects=[CgiEffect], 
-                     fn=sayHello, 
-                     name="sayHello")
-           ]
-\end{Verbatim}
-
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
-
-\begin{Verbatim}
-showHelloForm : UserForm
-showHelloForm = do
-  addTextBox "Name" FormString Nothing
-  useEffects [CgiEffect]
-  addSubmit sayHello handlers
-\end{Verbatim}
-
-Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
-
-\begin{Verbatim}
-cgiHello : CGIProg [] ()
-cgiHello = do
-  handler_set <- isHandlerSet
-  if handler_set then do
-    handleForm handlers
-    return ()
-  else do
-    addForm "nameform" "helloform" showHelloForm
-    return ()
-
-main : IO ()
-main = runCGI [initCGIState] cgiHello
-\end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the
-page, requesting a name from the user. Upon submission of the form, the form
-handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
-of the form-handling system: namely, the effect which allows the creation of
-forms, and the handling code which deserialises the data and passes it to the
-user-specified handler function.
-
-\subsection{Form Construction}
-\label{formcons}
-Each form element is specified  to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-\begin{Verbatim}
-data FormTy = FormString
-            | FormInt
-            | FormBool
-            | FormFloat
-            | FormList FormTy 
-\end{Verbatim}
-Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-\begin{Verbatim}
-interpFormTy : FormTy -> Type
-interpFormTy FormString = String
-interpFormTy FormInt = Int
-interpFormTy FormBool = Bool
-interpFormTy FormFloat = Float
-interpFormTy (FormList a) = List (interpFormTy a)
-\end{Verbatim}
-%
-In order to specify a form, we once again use \texttt{Effects}. By recording
-the type of each form element as it is added in the type of the form, we may
-statically ensure that the user-supplied handler function is of the correct
-type to handle the data supplied by the form: the specification of an
-incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code
-through \texttt{using} notation. We may therefore parameterise the 
-form data types over
-the types associated with each form element, and the effects required by the
-handler function.
-
-\begin{Verbatim}
-using (G : List FormTy, E : List WebEffect)
-  data FormRes : List FormTy -> 
-                 List WebEffect -> Type where
-    FR : Nat -> 
-         List FormTy -> 
-         List WebEffect -> 
-         String -> 
-         FormRes G E
-
-  data Form : Effect where
-    AddTextBox : (label : String) -> 
-                 (fty : FormTy) -> 
-                 (Maybe (interpFormTy fty)) -> 
-                 Form (FormRes G E) 
-                      (FormRes (fty :: G) E) () 
-   ...
-
-    Submit : (mkHandlerFn ((reverse G), E)) ->
-             String -> 
-             Form (FormRes G E) (FormRes [] []) String
-\end{Verbatim}
-The implementation of the form effect also contains other constructs for
-additional form elements such as radio buttons and check boxes, but are omitted
-in the interest of brevity.
-
-We make use of the resource associated with the effect, \texttt{FormRes}, to
-construct the form. The resource allows us to record the types associated with
-each form element and the HTML required to display the form. The constructor,
-\texttt{FR}, requires the number of elements in the form in order to allow for
-the naming of new elements, the list of element types, the list of effects
-supported by the handler function, and the currently generated HTML for the
-form.
-
-By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
-It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
-
-By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource.
-The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
-
-To specify a form instance, we define a function of type \texttt{UserForm}:
-\begin{Verbatim}
-  UserForm : Type
-  UserForm = EffM m [FORM (FormRes []) 
-                          (FormRes [])] String
-\end{Verbatim}
-All forms are required to include a submit button, as mandated by the
-requirement that the input and output resource contains an empty list of types;
-this requirement is fulfilled as per the output resource type of the
-\texttt{AddSubmit} operation. As the creation of a form is a return ()function
-which does not include side effects, we do not restrict the handler to IO as
-with previously-discussed effects, instead denoting the fact that it may be run
-in any handler with the implicit variable \texttt{m}.
-
-Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
-Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
-If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
-
-Before associating a handler function with the form, we must specify the effects available to the handler.
-This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
-By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-\begin{Verbatim}
-  useEffects : (effs : List WebEffect) ->
-               EffM m [FORM (FormRes G E)] 
-                      [FORM (FormRes G effs)] ()
-  useEffects effs = (UseEffects effs)
-\end{Verbatim}
-Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library permits introduction of additional effects within an effectful computation.
-%
-We may specify a handler function of type \texttt{FormHandler}:
-\begin{Verbatim}
-  FormHandler : List EFFECT -> Type
-  FormHandler effs = Eff IO effs ()
-\end{Verbatim}
-In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-\begin{Verbatim}
-  addSubmit : 
-    (f :  mkHandlerFn ((reverse G), E)) ->
-    (fns : HandlerList) ->
-    {default tactics 
-      { applyTactic findFn 100; solve; }
-      prf : FnElem f fns} ->
-    EffM m [FORM (FormRes G E)]
-           [FORM (FormRes [] [])] 
-           String
-  addSubmit f handlers {prf} = (Submit f name)
-    where name : String
-          name = getString' f handlers prf          
-\end{Verbatim}
-
-Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-\begin{Verbatim}
-MkHandlerFnTy : Type
-MkHandlerFnTy = (List FormTy, List WebEffect)
-
-mkHandlerFn' : List FormTy -> List WebEffect -> Type
-mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
-mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
-                              mkHandlerFn' xs effs
-
-mkHandlerFn : MkHandlerFnTy -> Type 
-mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{Verbatim}
-The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
-When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-
-To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{Verbatim}
-HandlerFn : Type
-HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{Verbatim}
-%
-The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
-We also store a unique string identifier, which is used to serialise a reference to the handler function.
-
-In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
-\begin{Verbatim}
-syntax 
-  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
-  ", name=" [name] = ((args, effs) ** (fn, name))
-\end{Verbatim}
-We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
-This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
-\begin{Verbatim}
-handler args=[FormString], 
-        effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
-\end{Verbatim}
-
-We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-\begin{Verbatim}
-HandlerList : Type
-HandlerList = List HandlerFn
-\end{Verbatim}
-To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-\begin{Verbatim}
-  data FnElem : mkHandlerFn ((reverse G), E) -> 
-                HandlerList -> Type where
-
-       FnHere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
-       FnThere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-               FnElem f xs -> FnElem f (x :: xs)
-\end{Verbatim}
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
-\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
-%TODO: this should probably be rewritten
-We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
-The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail.
-
-Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.
-
-Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this.
-Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.
-
-Running form construction is achieved as an operation of the CGI
-effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a
-hidden HTML field.
-
-\subsection{Form Handling}
-\label{formhandling}
-Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
-
-With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
-We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
-
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
-This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
-If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
-If not, the computation is abandoned and an error is shown.
-% I gave it a shot, right?
-:P
-\begin{Verbatim}
-checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
-                 (frm_fn_ty : MkHandlerFnTy) -> 
-                 mkHandlerFn reg_fn_ty -> 
-                 Maybe (mkHandlerFn frm_fn_ty)
-checkFunctions reg_ty frm_ty reg_fn with 
-                             (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
-                                            Just reg_fn
-  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{Verbatim}
-We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
-Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
-
-%-----------------------------
-%-----------------------------
-
-\section{Extended Example: Message Board}
-\label{messageboard}
-Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
-
-Firstly, we create a database schema in which to record information stored by the message board.
-We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
-
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
-Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
-\subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-\begin{Verbatim}
-main : IO ()
-main = do (runCGI [initCGIState, 
-                  InvalidSession, ()] 
-           handleRequest)
-          return ()
-\end{Verbatim}
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-\begin{Verbatim}
-handleRequest : CGIProg [SESSION 
-  (SessionRes SessionUninitialised), 
-  SQLITE ()] ()
-handleRequest = do 
-  handler_set <- isHandlerSet
-  if handler_set then do
-    lift' (handleForm handlers)
-    Effects.return ()
-  else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
-    handleNonFormRequest action (map strToInt thread_id)
-\end{Verbatim}
-\subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
-One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-\begin{Verbatim}
-newThreadForm : UserForm
-newThreadForm = do
-  addTextBox "Title" FormString Nothing
-  addTextBox "Post Content" FormString Nothing 
-  useEffects [CgiEffect, SessionEffect, SqliteEffect]
-  addSubmit handleNewThread handlers
-\end{Verbatim}
-The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
-Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
-The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleRegisterForm, 
-          name="handleRegisterForm"),
-
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleNewThread, 
-          name="handleNewThread"),
-     ...
-  ]
-\end{Verbatim}
-Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
-In order to do this, we make use of the session handler.
-We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
-If not, then a failure function is called instead.
-Should the form handler function be called with invalid arguments, an error is shown.
-\begin{Verbatim}
-handleNewThread : 
-  Maybe String -> Maybe String -> 
-  FormHandler [CGI (InitialisedCGI TaskRunning), 
-               SESSION (SessionRes SessionUninitialised), 
-               SQLITE ()]  
-handleNewThread (Just title) (Just content) = do 
-  withSession (addNewThread title content) notLoggedIn
-  return ()
-handleNewThread _ _ = do 
-  outputWithPreamble "<h1>Error</h1><br />There was 
-       an error posting your thread."
-  return ()
-\end{Verbatim}
-Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
-The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
-This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
-In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
-In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-\begin{Verbatim}
-threadInsert : Int -> 
-               String -> 
-               String -> 
-               Eff IO [SQLITE ()] Bool
-threadInsert uid title content = do
-  let query = "INSERT INTO `Threads` 
-    (`UserID`, `Title`) VALUES (?, ?)"
-  insert_res <- (executeInsert DB_NAME query 
-    [(1, DBInt uid), (2, DBText title)]
-  case insert_res of
-    Left err => return False
-    Right thread_id => postInsert uid thread_id content
-\end{Verbatim}
-\subsection{Listing Threads}
-Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-\begin{Verbatim}
-getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = 
-  executeSelect DB_NAME query [] collectThreadResults
- where query = "SELECT `ThreadID`, `Title`, `UserID`, 
-   `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{Verbatim}
-Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
-Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-\begin{Verbatim}
-traverseThreads : ResultSet -> 
-  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = return ()
-traverseThreads (x::xs) = do traverseRow x
-                             traverseThreads xs
-  where traverseRow : List DBVal -> 
-           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-        traverseRow ((DBInt thread_id)::
-                     (DBText title)::
-                     (DBInt user_id)::
-                     (DBText username)::[]) =
-           (output $ "<tr><td>
-            <a href=\"?action=showthread&thread_id=" ++ 
-            (show thread_id) ++ "\">" ++ 
-            title ++ "</a></td><td>" ++ 
-            username ++ "</td></tr>") 
-        traverseRow _ = return ()
-\end{Verbatim}
-\subsection{Authentication}
-Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
-If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
-The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
-Any failures, for example with creating a new session or querying the database, are reported to the user.
-\begin{Verbatim}
-handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
-  case auth_res of
-    Right (Just uid) => do
-      set_sess_res <- setSession uid
-      if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
-        return ()
-      else do
-        lift' (output "Could not set session")
-         return ()
-    Right Nothing => do
-      lift' (output "Invalid username or password")
-      return ()
-    Left err => do
-      lift' (output $ "Error: " ++ err)
-      return ()
-\end{Verbatim}
-Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
-
-Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
-Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
-Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
-%-----------------------------
-%-----------------------------
-
 \section{Related Work}
@@
 
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining
+\citet{thiemann:wash} adds the notion of a CGI Session for maintaining
 state, and provides more sophisticated form-handling methods. In particular,
@@
 being built upon monads, mean that the use of additional effects such as
-database access is achieved either through monad transformers or the ability to
-perform arbitrary IO operations. Both of these approaches are limited---the
+database access is achieved either through monad transformers or 
+performing arbitrary IO operations. Both of these approaches are limited---the
 former does not scale well to multiple effects, and the latter allows for the
@@
 \citet{plasmeijer:idata} describe an alternative approach to type-safe form
-handling which does not require the use of dependent types through the
+handling through the
 \textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of
 processing being triggered by form submission, as in the approach described in
-this paper, applications created in the \idata{} toolkit are instead
+this paper, applications created in the \idata{} toolkit are 
 edit-driven. This means that upon a component being edited, a computation
-occurs, given the state of the current form. This is then saved to be used in
+occurs, given the state of the current form. This is saved for
 future computations. Should a user enter invalid data, for example by entering
@@
 \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple
-\idata{} forms to interact with one another. This is achieved through the use
-of high-level combinators which allow the implementation of concepts such as
-recurseion, sequence and choice in a scalable fashion. These high-level
+\idata{} forms to interact with one another. This is achieved using
+high-level combinators which allow the implementation of concepts such as
+recursion, sequence and choice in a scalable fashion. These high-level
 abstractions are elegant, but the style and syntax differ substantially from a
@@
 
-Ur\/Web \cite{urweb} is a library built for the Ur language, which does not use dependent types but does have a more expressive type system, making extensive use of record types and type-level computation.
-By using these concepts, Ur\/Web may generate provably correct and unexploitable DOM code and SQL queries from records, without requiring developers to supply proofs. 
-In contrast to using runtime code generation, which is prone to obscure code generation errors, Ur\/Web makes use of its static type system to guarantee that metaprograms---in this case, generated SQL and DOM code---must be correct and secure.
-Such ideas regarding the use of static checking of metaprogram generation will be extremely useful when considering an object-relational mapping system, which we hope to implement in the near future.
-It will also be interesting to see how such concepts may be applied with a yet more expressive type system involving full dependent types.
+Ur\/Web \cite{urweb} is a library built for the Ur language, which does not use
+full dependent types but does have an expressive type system with 
+record types and type-level computation.  By using these concepts,
+Ur\/Web may generate provably correct and unexploitable DOM code and SQL
+queries from records, without requiring developers to supply proofs.  In
+contrast to using runtime code generation, which is prone to obscure code
+generation errors, Ur\/Web makes use of its static type system to guarantee
+that metaprograms---in this case, generated SQL and DOM code---must be correct
+and secure.  Such ideas regarding the use of static checking of metaprogram
+generation will be extremely useful when considering an object-relational
+mapping system, which we hope to implement in the near future. It will also be
+interesting to see how such concepts may be applied with a yet more expressive
+type system involving full dependent types.
+
 %Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
@@
 \section{Conclusions}
-Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
 
-Our framework, through the use of dependent types, provides several static guarantees.
-Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
-This immediately eliminates the possibilities of SQL injection attacks on non-string types.
-Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
-Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
+To date,
+dependently-typed languages have remained relatively unused for practical
+purposes. Advances in embedded domain-specific languages
+have allowed many of the advantages of such languages, including the additional
+guarantees given by the successful compilation of a program (assuming the use
+of specific enough types) to be adopted without additional proofs.
+
+Our framework, through the use of dependent types, provides several static
+guarantees. Data submitted by users is inherently unsafe due to the possibility
+of attacks such as SQL injection, and this particular threat is ameliorated due
+to elements to be associated with specific types during form construction. This
+immediately eliminates the possibilities of SQL injection attacks on non-string
+types. Since failures are handled transparently, no runtime errors are output
+to the browser, meaning that attackers may not use such information to aid
+attacks. Additionally, since checking is performed on the types of the form
+elements and the types of arguments accepted by the handler, it is impossible
+to associate a form with a handler incompatible with the submitted data.
 
@@
 % Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%-----------------------------
-%-----------------------------
-
-\acks
-This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
-The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
-% SICSA / EPSRC (grant number?
-% #idris
-% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{plainnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{refs}
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-11-11 paper/introduction.tex

--- 2013-08-14 paper/main.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-%\usepackage{silence}
-\usepackage{pgf}
-\usepackage{tikz}
-\usepackage{url}
-\usepackage[hidelinks]{hyperref}
-\usepackage{fancyvrb}
-\usepackage[numbers]{natbib}
-\usetikzlibrary{arrows,automata}
-%\WarningsOff
-
-\begin{document}
-\fvset{fontsize=\small}
-\newcommand{\idris}{\textsc{Idris}}
-\newcommand{\idata}{\textsf{iData}}
-\newcommand{\itasks}{\textsf{iTasks}}
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{2013} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{}        % These are ignored unless
-\preprintfooter{}   % 'preprint' option specified.
-
-\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
-%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
-
-\authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
-           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
-
-\maketitle
-
-\begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within
-programs in turn allowing easier reasoning about the operation of such
-programs. Stronger specifications do however bring with them the disadvantage
-that it becomes increasingly difficult to write programs that are accepted by
-the type checker and additional proofs may have to be specified by a user.
-
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in high-level languages which use dependent
-types to enforce certain invariants without additional proof obligations.
-
-In this paper, we apply this technique to web programming, and introduce an
-EDSL to facilitate the creation and handling of web forms which retain type
-information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how EDSLs may be constructed to enforce resource usage
-protocols associated with common web operations such as CGI, database access
-and session handling.
-
-\end{abstract}
-
-%\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%\terms
-%term1, term2
-
-%\keywords
-%keyword1, keyword2
-
 \section{Introduction}
@@
 
-%
-\begin{Verbatim}
+\begin{SaveVerbatim}{conn}
+
   $conn = new mysqli("localhost", "username", 
                      "password", "db");
-\end{Verbatim}
-%
+
+\end{SaveVerbatim}
+\useverb{conn}
+
+\noindent
 We then check to see if the connection was successful, and exit
 if not.  This check is optional, so
-it would be possible to omit it: this would, however, cause
+it would be possible to omit it. However, this would cause
 problems with later steps.
 
-%
-\begin{Verbatim}
+\begin{SaveVerbatim}{connerr}
+
   if (mysqli_connect_errno()) {
@@
   }
-\end{Verbatim}
-%
+
+\end{SaveVerbatim}
+\useverb{connerr}
+
+\noindent
 We then create a prepared statement detailing our query, and bind the
@@
 
-\begin{Verbatim}
+\begin{SaveVerbatim}{prepare}
+
   $stmt = $conn->prepare("SELECT `name`, `address` 
@@
   $stmt->bind_param('s', $dept);
-\end{Verbatim}
-%
+
+\end{SaveVerbatim}
+\useverb{prepare}
+
+\noindent
 After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. At
-this point, a developer may forget to execute the statement before beginning to
-fetch results or may have forgotten to bind the parameters to the statement:
+variables into which results will be stored, and fetch each row in turn. 
+Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.
 
-%
-\begin{Verbatim}
+\begin{SaveVerbatim}{stmtexec}
+
 $stmt->execute();
@@
   }
-\end{Verbatim}
-%
+
+\end{SaveVerbatim}
+\useverb{stmtexec}
+
+\noindent
 Finally, once the statement and connection are no longer needed, they should be
@@
 
-%
-\begin{Verbatim}
+\begin{SaveVerbatim}{connclose}
+
   $stmt->close();
   $conn->close();
-\end{Verbatim}
-%
+
+\end{SaveVerbatim}
+\useverb{connclose}
+
+\noindent
 Even in this small example, there exists a specific resource usage protocol.
@@
 complexity of the underlying type system. EDSLs allow domain experts to
-focus on a problem domain, rather than the specific details of their
-solution.
+write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.
 
-\idris{} \cite{brady2011idris} is a language with full dependent types, and
+\idris{} \cite{brady2013idris} is a language with full dependent types, and
 extensive support for EDSLs through overloading and syntax macros. Through the
@@
 
-We structure the remainder of this paper as follows.
-We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
+We structure the remainder of this paper as follows. We provide a brief
+overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
+how this may be used to ensure adherence to resource usage protocols for CGI,
+SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
+type-safe form handling in Section ~\ref{form}, implemented using
+\texttt{Effects}; and discuss the larger example
+of a message board system making use of these components in Section
+~\ref{messageboard}.
 
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
+The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/idris-hackers/IdrisWeb}.
 % =================================================
 % =================================================
-
-\section{An overview of the \texttt{Effects} framework}
-\label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access
-as \textit{algebraic effects}, allowing greater code modularity. This section
-is intended to serve as a brief introduction to the framework; for more
-information, see \cite{brady:effects}.
-
-An effect is represented as an instance of the \texttt{EFFECT} type. Such
-instances may be associated with a \textit{resource}, Describing a program in
-the \texttt{Effects} library is achieved using the following data type:
-
-\begin{Verbatim}
-EffM : (m : Type -> Type) ->
-       (es : List EFFECT) ->
-       (es’ : List EFFECT) ->
-       (a : Type) -> Type
-\end{Verbatim}
-\texttt{EffM} is a type that is parameterised over an execution context
-(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
-of effects at the end of the program's execution (\texttt{es'}), and a return
-type (\texttt{a}). The list of input types can be used to ensure that an
-operation may only be called when the program is in a particular state, and
-since the list of input effects may differ from the list of output effects, we
-may modify the resource associated with the state after each operation. This
-allows us to simply enforce a resource usage protocol, and deviation from this
-protocol will result in a compile-time error.
-
-Implementing a new effect involves creating a new algebraic data type, of kind
-\texttt{Effect}, which describes the available operations and how they modify
-the resource associated with the effect. Each operation consists of an initial
-resource \texttt{res}, the resource at the end of the operation \texttt{res'},
-and the output type of the operation \texttt{t}.
-
-%TODO: Execution
-\begin{Verbatim}
-Effect : Type
-Effect = (res : Type) -> (res’ : Type) ->
-         (t : Type) -> Type
-\end{Verbatim}
-
-Operations are made available to users by writing functions with the
-corresponding \texttt{EffM} type, which may then be used to invoke the
-operation from within an effectful program.
-
-\begin{Verbatim}
-queryPostVar : 
- String -> 
- Eff m [CGI (InitialisedCGI TaskRunning)] 
-   (Maybe String)
-queryPostVar query = (QueryPostVar query)
-\end{Verbatim}
-
-Additionally, we define \textit{handlers}, which describe how each operation is
-interpreted in a given execution context. Finally, \texttt{Effect} instances
-are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
-constructor, which is provided with the initial resource value.
-
-% =================================================
-% =================================================
-
-\section{Modelling resource usage protocols}
-\label{rup}
-In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
-
-% =================================================
-
-\subsection{CGI}
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {Initialised};
-  \node[state]         (C) [below of=B] {TaskRunning};
-  \node[state]         (D) [below of=C] {TaskCompleted};
-  \node[state]         (E) [below of=D] {HeadersWritten};
-  \node[state]         (F) [below of=E] {ContentWritten};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
-\end{tikzpicture}
-}
-\caption{CGI States}
-\label{fig:cgistates}
-\end{figure}
-
-CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.
-
-A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
-Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
-If this were to happen, then the program would fail due to the fact that headers had not been written to the client.
-
-By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
-In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
-This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI process as the unit type, ().
-\begin{Verbatim}[samepage]
-data CGIStep = Initialised 
-             | TaskRunning 
-             | TaskCompleted 
-             | HeadersWritten 
-             | ContentWritten
-
-data InitialisedCGI : CGIStep -> Type where
-  ICgi : CGIInfo -> InitialisedCGI s
-\end{Verbatim}
-Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
-Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.
-
-Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved.
-The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
-The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
-
-\begin{Verbatim}
-data Cgi : Effect where
-    Init : Cgi () (InitialisedCGI Initialised) ()
-    StartRun : Cgi (InitialisedCGI Initialised) 
-                   (InitialisedCGI TaskRunning) ()
-    FinishRun : Cgi (InitialisedCGI TaskRunning) 
-                    (InitialisedCGI TaskCompleted) ()
-    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
-                       (InitialisedCGI HeadersWritten) ()
-    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
-                       (InitialisedCGI ContentWritten) ()
-    OutputData : String -> 
-                 Cgi (InitialisedCGI TaskRunning) 
-                     (InitialisedCGI TaskRunning) ()
-    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
-                               :: effs) -> 
-                       CGIProg effs a -> 
-                       Cgi (InitialisedCGI TaskRunning) 
-                           (InitialisedCGI TaskRunning) a
-\end{Verbatim}
-Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
-
-User code executes in the \texttt{TaskRunning} state.
-Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
-This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
-
-After the user code has finished execution, control returns to the library code.
-At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
-Finally, the headers and content are written which completes the process.
-Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
-
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-\begin{Verbatim}
-CGIProg : List EFFECT -> Type -> Type
-CGIProg effs a = 
- Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{Verbatim}
-This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
-
-A simple ``Hello, world!'' program would be defined as follows:
-
-\begin{Verbatim}
-module Main
-import Cgi
-
-sayHello : CGIProg [] ()
-sayHello = output "Hello, world!"
-
-main : IO ()
-main = runCGI [initCGIState] sayHello
-\end{Verbatim}
-% =================================================
-
-\subsection{Database access with SQLite}
-
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {ConnectionOpened};
-  \node[state]         (C) [below of=B] {PreparedStatementOpen};
-  \node[state]         (D) [below of=C] {PreparedStatementBinding};
-  \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExecuting};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
-        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
-\end{tikzpicture}
-}
-\caption{Database Resource Usage Protocol}
-\label{fig:sqlitestates}
-\end{figure}
-
-SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
-We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-
-The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
-Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
-Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user.
-Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.
-
-Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
-This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.
-
-\begin{Verbatim}[samepage]
-data Sqlite : Effect where
-  OpenDB : 
-    String -> 
-    Sqlite () 
-           (SQLiteRes ConnectionOpened) Bool
-  CloseDB : 
-    Sqlite (SQLiteRes ConnectionOpened) () Bool
-  PrepareStatement : 
-    String -> 
-    Sqlite (SQLiteRes ConnectionOpened) 
-           (SQLiteRes PreparedStatementOpen) Bool
-  StartBind : 
-    Sqlite (SQLiteRes PreparedStatementOpen) 
-           (SQLiteRes PreparedStatementBinding) ()
-  BindText : 
-    ArgPos -> String -> Int -> 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBinding) 
-           Bool
-  FinishBind : 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBound) 
-           Bool
-  ExecuteStatement : 
-    Sqlite (SQLiteRes PreparedStatementBound) 
-           (SQLiteRes PreparedStatementExecuting) ()
-  Finalise : 
-    Sqlite (SQLiteRes PreparedStatementExecuting) 
-           (SQLiteRes ConnectionOpened) 
-           Bool
-  GetColumnText : 
-    Int -> 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           String
-  RowStep : 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           StepResult
-\end{Verbatim}
-There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
-\begin{Verbatim}
-data SQLiteRes : Step -> Type where
-  OpenConn : DBPointer -> SQLiteRes s
-  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
-  ExecutingStmt : DBPointer -> 
-                  StmtPointer -> 
-                  StepResult -> 
-                  SQLiteRes s
-
-data DBPointer = ValidConn Ptr
-               | InvalidConn
-
-data StmtPointer = ValidStmt Ptr
-                 | InvalidStmt 
-\end{Verbatim}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
-At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
-
-%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
-In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
-Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function.
-We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-\begin{Verbatim}
-data StepResult = Unstarted
-                | StepFail
-                | StepComplete
-                | NoMoreRows
-\end{Verbatim}
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
-Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
-We may therefore statically guarantee that only calls that will return a valid result are executed.
-
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
-
-\subsubsection{Example Code}
-Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
-Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
-
-We define a function of type:
-\begin{Verbatim}
-String ->
-Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{Verbatim}
-This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-\begin{Verbatim}
-testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, Int)))
-testSelect dept = do
-  open_db <- openDB "people.db"
-  if open_db then do
-    let sql = "SELECT name, address FROM `staff` 
-                    WHERE dept = ?;"
-    sql_prep_res <- prepareStatement sql
-    if sql_prep_res then do 
-      startBind
-      bindText 1 dept
-      finishBind
-      beginExecution
-      results <- collectResults
-      finaliseStatement
-      closeDB
-      return $ Right results
-    else do err <- stmtFail
-            return $ Left err
-  else do err <- connFail
-          return $ Left err 
-\end{Verbatim}
-The program initially attempts to open a connection to the \texttt{people.db} database.
-At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
-The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
-
-A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.
-
-\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database.
-This function is then recursively called until there are no more rows to process.
-\begin{Verbatim}
-collectResults : 
-  Eff IO 
-  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-  (List (String, Int))
-collectResults = do
-  step_result <- nextRow
-  case step_result of
-      StepComplete => do name <- getColumnText 1
-                         age <- getColumnInt 2
-                         xs <- collectResults
-                         return $ (name, age) :: xs
-      NoMoreRows => return []
-      StepFail => return [] 
-\end{Verbatim}
-In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
-In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-\begin{Verbatim}
-data DBVal = DBInt Int
-           | DBText String
-           | DBFloat Float
-           | DBNull
-\end{Verbatim}
-We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-\begin{Verbatim}
-  ResultSet : Type
-  ResultSet = List (List DBVal)
-\end{Verbatim}
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-\begin{Verbatim}
-executeSelect : 
-  String -> String -> List (Int, DBVal) -> 
-  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-                 (List DBVal)) -> 
-  Eff m [SQLITE ()] (Either String ResultSet)
-\end{Verbatim}
-% =================================================
-\subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate requests.
-This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
-In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.
-
-A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects.
-By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.
-
-We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
-%TODO: Fix up the labels
-\begin{Verbatim}[samepage]
-data Session : Effect where
-  LoadSession : 
-  SessionID -> 
-  Session (SessionRes SessionUninitialised) 
-          (SessionRes SessionInitialised) 
-          (Maybe SessionData)
-  UpdateSession : 
-    SessionData -> 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) ()
-  CreateSession : 
-    SessionData -> 
-    Session (SessionRes SessionUninitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  DeleteSession : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool 
-  WriteToDB : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool
-  DiscardSessionChanges : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) ()
-  GetSessionID : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  GetSessionData : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionData)
-\end{Verbatim}
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {SessionUninitialised};
-  \node[state]         (B) [below of=A] {SessionInitialised};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
-\end{tikzpicture}
-}
-\caption{Session Handler Resource Usage Protocol}
-\label{fig:sessionstates}
-\end{figure}
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler.
-In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key.
-The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.
-This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.
-
-Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
-\begin{Verbatim}
-data SessionStep = SessionUninitialised
-                 | SessionInitialised
-
-data SessionRes : SessionStep -> Type where
-  InvalidSession : SessionRes s  
-  ValidSession   : SessionID -> 
-                   SessionData -> 
-                   SessionRes s
-\end{Verbatim}
-The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
-If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.
-
-% =================================================
-
-\section{Type-aware form handling}
-\label{form}
-Programming web applications often involves processing user data, which may then be used in further effectful computations.
-Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.
-
-This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker.
-One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be used in a query such as:
-
-\begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + 
-       page + ', 5';
-\end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
-% Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here
-
-In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information.
-Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and echoes it back.
-Firstly, we define a form handler which echoes back a string provided by the form handler.
-It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.
-
-\begin{Verbatim}
-sayHello : Maybe String -> 
-           FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
-\end{Verbatim}
-
-We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
-
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [(handler args=[FormString], 
-                     effects=[CgiEffect], 
-                     fn=sayHello, 
-                     name="sayHello")
-           ]
-\end{Verbatim}
-
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
-
-\begin{Verbatim}
-showHelloForm : UserForm
-showHelloForm = do
-  addTextBox "Name" FormString Nothing
-  useEffects [CgiEffect]
-  addSubmit sayHello handlers
-\end{Verbatim}
-
-Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
-
-\begin{Verbatim}
-cgiHello : CGIProg [] ()
-cgiHello = do
-  handler_set <- isHandlerSet
-  if handler_set then do
-    handleForm handlers
-    return ()
-  else do
-    addForm "nameform" "helloform" showHelloForm
-    return ()
-
-main : IO ()
-main = runCGI [initCGIState] cgiHello
-\end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user.
-Upon submission of the form, the form handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
-\subsection{Form Construction}
-\label{formcons}
-Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-\begin{Verbatim}
-data FormTy = FormString
-            | FormInt
-            | FormBool
-            | FormFloat
-            | FormList FormTy 
-\end{Verbatim}
-Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-\begin{Verbatim}
-interpFormTy : FormTy -> Type
-interpFormTy FormString = String
-interpFormTy FormInt = Int
-interpFormTy FormBool = Bool
-interpFormTy FormFloat = Float
-interpFormTy (FormList a) = List (interpFormTy a)
-\end{Verbatim}
-%
-In order to specify a form, we once again use \texttt{Effects}.
-By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation.
-We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.
-
-\begin{Verbatim}
-using (G : List FormTy, E : List WebEffect)
-  data FormRes : List FormTy -> 
-                 List WebEffect -> Type where
-    FR : Nat -> 
-         List FormTy -> 
-         List WebEffect -> 
-         String -> 
-         FormRes G E
-
-  data Form : Effect where
-    AddTextBox : (label : String) -> 
-                 (fty : FormTy) -> 
-                 (Maybe (interpFormTy fty)) -> 
-                 Form (FormRes G E) 
-                      (FormRes (fty :: G) E) () 
-   ...
-
-    Submit : (mkHandlerFn ((reverse G), E)) ->
-             String -> 
-             Form (FormRes G E) (FormRes [] []) String
-\end{Verbatim}
-The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
-
-We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form.
-The resource allows us to record the types associated with each form element and the HTML required to display the form.
-he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
-
-By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
-It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
-
-By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource.
-The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
-
-To specify a form instance, we define a function of type \texttt{UserForm}:
-\begin{Verbatim}
-  UserForm : Type
-  UserForm = EffM m [FORM (FormRes []) 
-                          (FormRes [])] String
-\end{Verbatim}
-All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
-As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.
-
-Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
-Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
-If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
-
-Before associating a handler function with the form, we must specify the effects available to the handler.
-This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
-By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-\begin{Verbatim}
-  useEffects : (effs : List WebEffect) ->
-               EffM m [FORM (FormRes G E)] 
-                      [FORM (FormRes G effs)] ()
-  useEffects effs = (UseEffects effs)
-\end{Verbatim}
-Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
-%
-We may specify a handler function of type \texttt{FormHandler}:
-\begin{Verbatim}
-  FormHandler : List EFFECT -> Type
-  FormHandler effs = Eff IO effs ()
-\end{Verbatim}
-In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-\begin{Verbatim}
-  addSubmit : 
-    (f :  mkHandlerFn ((reverse G), E)) ->
-    (fns : HandlerList) ->
-    {default tactics 
-      { applyTactic findFn 100; solve; }
-      prf : FnElem f fns} ->
-    EffM m [FORM (FormRes G E)]
-           [FORM (FormRes [] [])] 
-           String
-  addSubmit f handlers {prf} = (Submit f name)
-    where name : String
-          name = getString' f handlers prf          
-\end{Verbatim}
-
-Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-\begin{Verbatim}
-MkHandlerFnTy : Type
-MkHandlerFnTy = (List FormTy, List WebEffect)
-
-mkHandlerFn' : List FormTy -> List WebEffect -> Type
-mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
-mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
-                              mkHandlerFn' xs effs
-
-mkHandlerFn : MkHandlerFnTy -> Type 
-mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{Verbatim}
-The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
-When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-
-To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{Verbatim}
-HandlerFn : Type
-HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{Verbatim}
-%
-The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
-We also store a unique string identifier, which is used to serialise a reference to the handler function.
-
-In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
-\begin{Verbatim}
-syntax 
-  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
-  ", name=" [name] = ((args, effs) ** (fn, name))
-\end{Verbatim}
-We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
-This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
-\begin{Verbatim}
-handler args=[FormString], 
-        effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
-\end{Verbatim}
-
-We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-\begin{Verbatim}
-HandlerList : Type
-HandlerList = List HandlerFn
-\end{Verbatim}
-To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-\begin{Verbatim}
-  data FnElem : mkHandlerFn ((reverse G), E) -> 
-                HandlerList -> Type where
-
-       FnHere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
-       FnThere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-               FnElem f xs -> FnElem f (x :: xs)
-\end{Verbatim}
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
-\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
-%TODO: this should probably be rewritten
-We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
-The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail.
-
-Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.
-
-Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this.
-Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.
-
-Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a hidden HTML field.
-
-\subsection{Form Handling}
-\label{formhandling}
-Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
-
-With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
-We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
-
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
-This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
-If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
-If not, the computation is abandoned and an error is shown.
-% I gave it a shot, right?
-:P
-\begin{Verbatim}
-checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
-                 (frm_fn_ty : MkHandlerFnTy) -> 
-                 mkHandlerFn reg_fn_ty -> 
-                 Maybe (mkHandlerFn frm_fn_ty)
-checkFunctions reg_ty frm_ty reg_fn with 
-                             (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
-                                            Just reg_fn
-  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{Verbatim}
-We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
-Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
-
-%-----------------------------
-%-----------------------------
-
-\section{Extended Example: Message Board}
-\label{messageboard}
-Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
-
-Firstly, we create a database schema in which to record information stored by the message board.
-We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
-
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
-Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
-\subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-\begin{Verbatim}
-main : IO ()
-main = do (runCGI [initCGIState, 
-                  InvalidSession, ()] 
-           handleRequest)
-          pure ()
-\end{Verbatim}
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-\begin{Verbatim}
-handleRequest : CGIProg [SESSION 
-  (SessionRes SessionUninitialised), 
-  SQLITE ()] ()
-handleRequest = do 
-  handler_set <- isHandlerSet
-  if handler_set then do
-    lift' (handleForm handlers)
-    Effects.pure ()
-  else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
-    handleNonFormRequest action (map strToInt thread_id)
-\end{Verbatim}
-\subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
-One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-\begin{Verbatim}
-newThreadForm : UserForm
-newThreadForm = do
-  addTextBox "Title" FormString Nothing
-  addTextBox "Post Content" FormString Nothing 
-  useEffects [CgiEffect, SessionEffect, SqliteEffect]
-  addSubmit handleNewThread handlers
-\end{Verbatim}
-The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
-Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
-The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleRegisterForm, 
-          name="handleRegisterForm"),
-
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleNewThread, 
-          name="handleNewThread"),
-     ...
-  ]
-\end{Verbatim}
-Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
-In order to do this, we make use of the session handler.
-We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
-If not, then a failure function is called instead.
-Should the form handler function be called with invalid arguments, an error is shown.
-\begin{Verbatim}
-handleNewThread : 
-  Maybe String -> Maybe String -> 
-  FormHandler [CGI (InitialisedCGI TaskRunning), 
-               SESSION (SessionRes SessionUninitialised), 
-               SQLITE ()]  
-handleNewThread (Just title) (Just content) = do 
-  withSession (addNewThread title content) notLoggedIn
-  pure ()
-handleNewThread _ _ = do 
-  outputWithPreamble "<h1>Error</h1><br />There was 
-       an error posting your thread."
-  pure ()
-\end{Verbatim}
-Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
-The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
-This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
-In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
-In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-\begin{Verbatim}
-threadInsert : Int -> 
-               String -> 
-               String -> 
-               Eff IO [SQLITE ()] Bool
-threadInsert uid title content = do
-  let query = "INSERT INTO `Threads` 
-    (`UserID`, `Title`) VALUES (?, ?)"
-  insert_res <- (executeInsert DB_NAME query 
-    [(1, DBInt uid), (2, DBText title)]
-  case insert_res of
-    Left err => pure False
-    Right thread_id => postInsert uid thread_id content
-\end{Verbatim}
-\subsection{Listing Threads}
-Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-\begin{Verbatim}
-getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = 
-  executeSelect DB_NAME query [] collectThreadResults
- where query = "SELECT `ThreadID`, `Title`, `UserID`, 
-   `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{Verbatim}
-Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
-Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-\begin{Verbatim}
-traverseThreads : ResultSet -> 
-  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = pure ()
-traverseThreads (x::xs) = do traverseRow x
-                             traverseThreads xs
-  where traverseRow : List DBVal -> 
-           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-        traverseRow ((DBInt thread_id)::
-                     (DBText title)::
-                     (DBInt user_id)::
-                     (DBText username)::[]) =
-           (output $ "<tr><td>
-            <a href=\"?action=showthread&thread_id=" ++ 
-            (show thread_id) ++ "\">" ++ 
-            title ++ "</a></td><td>" ++ 
-            username ++ "</td></tr>") 
-        traverseRow _ = pure ()
-\end{Verbatim}
-\subsection{Authentication}
-Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
-If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
-The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
-Any failures, for example with creating a new session or querying the database, are reported to the user.
-\begin{Verbatim}
-handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
-  case auth_res of
-    Right (Just uid) => do
-      set_sess_res <- setSession uid
-      if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
-        pure ()
-      else do
-        lift' (output "Could not set session")
-        pure ()
-    Right Nothing => do
-      lift' (output "Invalid username or password")
-      pure ()
-    Left err => do
-      lift' (output $ "Error: " ++ err)
-      pure ()
-\end{Verbatim}
-Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
-
-Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
-Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
-Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
-%-----------------------------
-%-----------------------------
-
-\section{Related Work}
-\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser.
-The library also provides support for cookies and basic form handling.
-
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods.
-In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs.
-Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs.
-Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations.
-Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.
-
-\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction.
-Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven.
-This means that upon a component being edited, a computation occurs, given the state of the current form.
-This is then saved to be used in future computations.
-Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted.
-This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.
-
-The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another.
-This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion.
-These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application.
-Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.
-
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs.
-Through the use of the Ur\/Web DSL, developers may generate well-typed code from records.
-By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure.
-Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.
-
-WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres.
-WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes.
-The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables.
-These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy.
-Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing.
-WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation.
-We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.
-
-%-----------------------------
-%-----------------------------
-
-\section{Conclusions}
-Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
-
-Our framework, through the use of dependent types, provides several static guarantees.
-Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
-This immediately eliminates the possibilities of SQL injection attacks on non-string types.
-Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
-Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
-
-Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
-Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
-By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
-
-In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
-With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
-
-Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
-\subsection{Further Work}
-We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
-
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
-
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
-We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
-
-Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
-Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
-Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
-
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
-Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
-This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
-Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
-Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
-
-Dependently-typed languages provide great promise for the construction of secure and correct programs.
-Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
-% Cryptography bindings would be really good.
-% Integration with a web server instead of doing everything over CGI
-% Improvements to form handling system
-% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%-----------------------------
-%-----------------------------
-
-\acks
-This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
-The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
-% SICSA / EPSRC (grant number?
-% #idris
-% Idris contributors, in particular Melissa for the SQLite bindings and whoever wrote Network.Cgi
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{plainnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{refs}
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2013-11-11 paper/main.tex

--- 2013-08-14 paper/main.tex
@@
 \usepackage[numbers]{natbib}
-\usetikzlibrary{arrows,automata}
 %\WarningsOff
+
+\newcommand{\useverbtb}[1]{\begin{tiny}\BUseVerbatim{#1}\end{tiny}}
+\newcommand{\useverb}[1]{\begin{small}\BUseVerbatim{#1}\end{small}}
+\newcommand{\useverbb}[1]{\begin{small}\BUseVerbatim{#1}\end{small}}
 
@@
 \newcommand{\itasks}{\textsf{iTasks}}
+\newcommand{\Idris}{\textsc{Idris}}
 \special{papersize=8.5in,11in}
@@
 
-\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
-%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}
-
+\title{Dependent Types for Safe and Secure Web Programming}
 \authorinfo{Simon Fowler \and Edwin Brady}
-           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
+           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
            {Email: \{sf37, ecb10\}@st-andrews.ac.uk}
@@
 \begin{abstract}
-Dependently-typed languages allow for much expressive types to be used within
-programs in turn allowing easier reasoning about the operation of such
-programs. Stronger specifications do however bring with them the disadvantage
-that it becomes increasingly difficult to write programs that are accepted by
-the type checker and additional proofs may have to be specified by a user.
+Dependently-typed languages allow precise types to be used during development,
+facilitating easier reasoning about the operation of programs. However,
+stronger types bring a disadvantage that it becomes increasingly difficult to
+write programs that are accepted by a type checker and additional proofs may
+have to be specified by a programmer.
 
@@
 a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in high-level languages which use dependent
-types to enforce certain invariants without additional proof obligations.
+domain-specific code to be written in a high-level languages which uses
+dependent types to enforce certain invariants without additional proof
+obligations.
 
-In this paper, we apply this technique to web programming, and introduce an
-EDSL to facilitate the creation and handling of web forms which retain type
-information, reducing the scope for programmer error and attacks such as SQL
-injection. We also show how EDSLs may be constructed to enforce resource usage
-protocols associated with common web operations such as CGI, database access
-and session handling.
+In this paper, we apply this technique to web programming.  Using the
+dependently typed programming language \Idris{}, we introduce an EDSL to
+facilitate the creation and handling of \emph{statically checked} web forms,
+reducing the scope for programmer error and attacks such as SQL injection. We
+also show how to enforce resource usage protocols associated with common web
+operations such as CGI, database access and session handling.
 
@@
 
-\section{Introduction}
-
-Web applications, whilst ubiquitous, are also prone to incorrect construction
-and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
-scripting \cite{owasp:xss}. Security breaches using such exploits are
-far-reaching, and high profile cases involve large corporations such as Sony,
-who suffered a well-publicised and extremely costly SQL Injection breach in
-2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
-\cite{imperva:yahoo}.
-
-Many web applications are written in dynamically-checked scripting languages
-such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. A significant drawback, however, is that such languages
-do not provide  the same static guarantees about runtime behaviour afforded by
-programs with more expressive, static type systems, instead relying on
-extensive unit testing to ensure correctness and security.
-
-%%------- EXAMPLE
-Let us consider a simple database access routine, written in
-PHP, where we wish to obtain the name and address of every employee working in
-a given department, \texttt{\$dept}. We firstly construct an object
-representing a database connection, where the arguments are the database host,
-user, password and name respectively:
-
-%
-\begin{Verbatim}
-$conn = new mysqli("localhost", "username", 
-                    "password", "db");
-\end{Verbatim}
-%
-We then check to see if the connection was successful, and exit
-if not.  This check is optional, so
-it would be possible to omit it: this would, however, cause
-problems with later steps.
-
-%
-\begin{Verbatim}
-  if (mysqli_connect_errno()) {
-      exit();
-  }
-\end{Verbatim}
-%
-We then create a prepared statement detailing our query, and bind the
-\texttt{`dept'} value:
-
-\begin{Verbatim}
-  $stmt = $conn->prepare("SELECT `name`, `address` 
-     FROM `staff` WHERE `dept` = ?);
-  $stmt->bind_param('s', $dept);
-\end{Verbatim}
-%
-After the parameters have been bound, we execute the statement, assign
-variables into which results will be stored, and fetch each row in turn. At
-this point, a developer may forget to execute the statement before beginning to
-fetch results or may have forgotten to bind the parameters to the statement:
-
-%
-\begin{Verbatim}
-  $stmt->execute();
-  $stmt->bind_result($name, $address);
-  while ($stmt->fetch()) {
-    printf("Name: %s, Age: %s", $name, $age);
-  }
-\end{Verbatim}
-%
-Finally, once the statement and connection are no longer needed, they should be
-closed in order to discard the associated resources:
-
-%
-\begin{Verbatim}
-  $stmt->close();
-  $conn->close();
-\end{Verbatim}
-%
-Even in this small example, there exists a specific resource usage protocol.
-Firstly, a connection to the database must be opened. The object-oriented style
-used in the example encapsulates this to an extent, as the object must be
-created in order for operations to be performed, however it is less obvious in
-a procedural version of the code. Secondly, a prepared statement is created,
-using the raw SQL and placeholders to which variables are later bound. The
-statement is then executed, and each row is retrieved from the database.
-Finally, the resources are freed.
-
-Problems may arise if the protocol is not followed correctly.
-A developer may, for example, accidentally close a statement whilst still
-retrieving rows, which would cause a runtime error. Similarly, a programmer may
-omit closing the statement or connection, which can lead to
-problems such as resource leaks in longer-running server applications.
-However, in conventional programming languages, there is no way to check
-automatically that a protocol is followed.
-
-In contrast, the use of \textit{dependent types} makes it possible
-to specify a program's behaviour precisely, and to check that a 
-specification is followed.
-%
-The difficulty is 
-that automatic verification by a compiler can be difficult or
-often impossible, requiring additional proofs to be given by the programmer.
-
-This complexity can be ameliorated through the use of \textit{embedded
-domain-specific languages} (EDSLs), which aim to abstract away the
-complexity of the underlying type system. EDSLs allow domain experts to
-focus on a problem domain, rather than the specific details of their
-solution.
-
-\idris{} \cite{brady2011idris} is a language with full dependent types, and
-extensive support for EDSLs through overloading and syntax macros. Through the
-use of \idris{}, and a framework for describing resource protocols using
-\emph{algebraic effects}~\cite{brady:effects}, we
-present a dependently-typed web framework, which allows the construction of
-programs with additional guarantees about correctness and security, whilst
-minimising the increase in development complexity.
-
-\subsection{Contributions}
-The primary contribution of this paper is the application of 
-dependent types may to provide strong static guarantees
-about the correctness and security of web applications, whilst minimising
-additional development complexity. In particular, we present:
-
-\begin{itemize}
-\item A form-handling mechanism, which preserves type information and
-manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection.
-
-\item Representations of CGI, Databases and sessions as
-\textit{algebraic effects}, allowing programs to be accepted only when they
-follow clearly defined resource usage protocols.
-
-\item A practical demonstration of the framework, a message board application.
-
-\end{itemize}
-
-We structure the remainder of this paper as follows.
-We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.
-
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
-% =================================================
-% =================================================
-
-\section{An overview of the \texttt{Effects} framework}
-\label{effects}
-\texttt{Effects} is an EDSL which handles side-effects such as database access
-as \textit{algebraic effects}, allowing greater code modularity. This section
-is intended to serve as a brief introduction to the framework; for more
-information, see \cite{brady:effects}.
-
-An effect is represented as an instance of the \texttt{EFFECT} type. Such
-instances may be associated with a \textit{resource}, Describing a program in
-the \texttt{Effects} library is achieved using the following data type:
-
-\begin{Verbatim}
-EffM : (m : Type -> Type) ->
-       (es : List EFFECT) ->
-       (es’ : List EFFECT) ->
-       (a : Type) -> Type
-\end{Verbatim}
-\texttt{EffM} is a type that is parameterised over an execution context
-(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
-of effects at the end of the program's execution (\texttt{es'}), and a return
-type (\texttt{a}). The list of input types can be used to ensure that an
-operation may only be called when the program is in a particular state, and
-since the list of input effects may differ from the list of output effects, we
-may modify the resource associated with the state after each operation. This
-allows us to simply enforce a resource usage protocol, and deviation from this
-protocol will result in a compile-time error.
-
-Implementing a new effect involves creating a new algebraic data type, of kind
-\texttt{Effect}, which describes the available operations and how they modify
-the resource associated with the effect. Each operation consists of an initial
-resource \texttt{res}, the resource at the end of the operation \texttt{res'},
-and the output type of the operation \texttt{t}.
-
-%TODO: Execution
-\begin{Verbatim}
-Effect : Type
-Effect = (res : Type) -> (res’ : Type) ->
-         (t : Type) -> Type
-\end{Verbatim}
-
-Operations are made available to users by writing functions with the
-corresponding \texttt{EffM} type, which may then be used to invoke the
-operation from within an effectful program.
-
-\begin{Verbatim}
-queryPostVar : 
- String -> 
- Eff m [CGI (InitialisedCGI TaskRunning)] 
-   (Maybe String)
-queryPostVar query = (QueryPostVar query)
-\end{Verbatim}
-
-Additionally, we define \textit{handlers}, which describe how each operation is
-interpreted in a given execution context. Finally, \texttt{Effect} instances
-are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
-constructor, which is provided with the initial resource value.
-
-% =================================================
-% =================================================
-
-\section{Modelling resource usage protocols}
-\label{rup}
-In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
-
-% =================================================
-
-\subsection{CGI}
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {Initialised};
-  \node[state]         (C) [below of=B] {TaskRunning};
-  \node[state]         (D) [below of=C] {TaskCompleted};
-  \node[state]         (E) [below of=D] {HeadersWritten};
-  \node[state]         (F) [below of=E] {ContentWritten};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
-\end{tikzpicture}
-}
-\caption{CGI States}
-\label{fig:cgistates}
-\end{figure}
-
-CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.
-
-A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
-Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
-If this were to happen, then the program would fail due to the fact that headers had not been written to the client.
-
-By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
-In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
-This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI process as the unit type, ().
-\begin{Verbatim}[samepage]
-data CGIStep = Initialised 
-             | TaskRunning 
-             | TaskCompleted 
-             | HeadersWritten 
-             | ContentWritten
-
-data InitialisedCGI : CGIStep -> Type where
-  ICgi : CGIInfo -> InitialisedCGI s
-\end{Verbatim}
-Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses.
-Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.
-
-Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved.
-The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}.
-The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.
-
-\begin{Verbatim}
-data Cgi : Effect where
-    Init : Cgi () (InitialisedCGI Initialised) ()
-    StartRun : Cgi (InitialisedCGI Initialised) 
-                   (InitialisedCGI TaskRunning) ()
-    FinishRun : Cgi (InitialisedCGI TaskRunning) 
-                    (InitialisedCGI TaskCompleted) ()
-    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
-                       (InitialisedCGI HeadersWritten) ()
-    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
-                       (InitialisedCGI ContentWritten) ()
-    OutputData : String -> 
-                 Cgi (InitialisedCGI TaskRunning) 
-                     (InitialisedCGI TaskRunning) ()
-    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
-                               :: effs) -> 
-                       CGIProg effs a -> 
-                       Cgi (InitialisedCGI TaskRunning) 
-                           (InitialisedCGI TaskRunning) a
-\end{Verbatim}
-Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.
-
-User code executes in the \texttt{TaskRunning} state.
-Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers.
-It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record.
-This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.
-
-After the user code has finished execution, control returns to the library code.
-At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request.
-Finally, the headers and content are written which completes the process.
-Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.
-
-In \idris{}, types are first-class, meaning that they may be treated like other terms in computations.
-We may therefore define the following type synonym, which we may use in order to make use of the CGI framework:
-\begin{Verbatim}
-CGIProg : List EFFECT -> Type -> Type
-CGIProg effs a = 
- Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
-\end{Verbatim}
-This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.
-
-A simple ``Hello, world!'' program would be defined as follows:
-
-\begin{Verbatim}
-module Main
-import Cgi
-
-sayHello : CGIProg [] ()
-sayHello = output "Hello, world!"
-
-main : IO ()
-main = runCGI [initCGIState] sayHello
-\end{Verbatim}
-% =================================================
-
-\subsection{Database access with SQLite}
-
-%TODO: Fix up the labels
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {ConnectionOpened};
-  \node[state]         (C) [below of=B] {PreparedStatementOpen};
-  \node[state]         (D) [below of=C] {PreparedStatementBinding};
-  \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExecuting};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
-        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
-\end{tikzpicture}
-}
-\caption{Database Resource Usage Protocol}
-\label{fig:sqlitestates}
-\end{figure}
-
-SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications.
-We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.
-
-The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure.
-Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
-Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user.
-Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.
-
-Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library.
-This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.
-
-\begin{Verbatim}[samepage]
-data Sqlite : Effect where
-  OpenDB : 
-    String -> 
-    Sqlite () 
-           (SQLiteRes ConnectionOpened) Bool
-  CloseDB : 
-    Sqlite (SQLiteRes ConnectionOpened) () Bool
-  PrepareStatement : 
-    String -> 
-    Sqlite (SQLiteRes ConnectionOpened) 
-           (SQLiteRes PreparedStatementOpen) Bool
-  StartBind : 
-    Sqlite (SQLiteRes PreparedStatementOpen) 
-           (SQLiteRes PreparedStatementBinding) ()
-  BindText : 
-    ArgPos -> String -> Int -> 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBinding) 
-           Bool
-  FinishBind : 
-    Sqlite (SQLiteRes PreparedStatementBinding) 
-           (SQLiteRes PreparedStatementBound) 
-           Bool
-  ExecuteStatement : 
-    Sqlite (SQLiteRes PreparedStatementBound) 
-           (SQLiteRes PreparedStatementExecuting) ()
-  Finalise : 
-    Sqlite (SQLiteRes PreparedStatementExecuting) 
-           (SQLiteRes ConnectionOpened) 
-           Bool
-  GetColumnText : 
-    Int -> 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           String
-  RowStep : 
-    Sqlite (SQLiteRes PreparedStatementExecuting)
-           (SQLiteRes PreparedStatementExecuting)
-           StepResult
-\end{Verbatim}
-There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query.
-This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
-\begin{Verbatim}
-data SQLiteRes : Step -> Type where
-  OpenConn : DBPointer -> SQLiteRes s
-  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
-  ExecutingStmt : DBPointer -> 
-                  StmtPointer -> 
-                  StepResult -> 
-                  SQLiteRes s
-
-data DBPointer = ValidConn Ptr
-               | InvalidConn
-
-data StmtPointer = ValidStmt Ptr
-                 | InvalidStmt 
-\end{Verbatim}
-If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}.
-At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety.
-The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.
-
-%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
-SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}.
-In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function.
-Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function.
-We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.
-\begin{Verbatim}
-data StepResult = Unstarted
-                | StepFail
-                | StepComplete
-                | NoMoreRows
-\end{Verbatim}
-Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource.
-Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}.
-We may therefore statically guarantee that only calls that will return a valid result are executed.
-
-By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.
-
-\subsubsection{Example Code}
-Programs making use of the DSL should look familiar to developers even without a background in functional programming.
-To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department.
-Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.
-
-We define a function of type:
-\begin{Verbatim}
-String ->
-Eff IO [SQLITE ()] (Either String (List (String, String)))
-\end{Verbatim}
-This means that the program will be run in the IO execution context, and must start and end with no active resources.
-The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned.
-\begin{Verbatim}
-testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, Int)))
-testSelect dept = do
-  open_db <- openDB "people.db"
-  if open_db then do
-    let sql = "SELECT name, address FROM `staff` 
-                    WHERE dept = ?;"
-    sql_prep_res <- prepareStatement sql
-    if sql_prep_res then do 
-      startBind
-      bindText 1 dept
-      finishBind
-      beginExecution
-      results <- collectResults
-      finaliseStatement
-      closeDB
-      return $ Right results
-    else do err <- stmtFail
-            return $ Left err
-  else do err <- connFail
-          return $ Left err 
-\end{Verbatim}
-The program initially attempts to open a connection to the \texttt{people.db} database.
-At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state.
-The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful.
-If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.
-
-A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.
-
-\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database.
-This function is then recursively called until there are no more rows to process.
-\begin{Verbatim}
-collectResults : 
-  Eff IO 
-  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-  (List (String, Int))
-collectResults = do
-  step_result <- nextRow
-  case step_result of
-      StepComplete => do name <- getColumnText 1
-                         age <- getColumnInt 2
-                         xs <- collectResults
-                         return $ (name, age) :: xs
-      NoMoreRows => return []
-      StepFail => return [] 
-\end{Verbatim}
-In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process.
-In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
-\begin{Verbatim}
-data DBVal = DBInt Int
-           | DBText String
-           | DBFloat Float
-           | DBNull
-\end{Verbatim}
-We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
-\begin{Verbatim}
-  ResultSet : Type
-  ResultSet = List (List DBVal)
-\end{Verbatim}
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
-\begin{Verbatim}
-executeSelect : 
-  String -> String -> List (Int, DBVal) -> 
-  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
-                 (List DBVal)) -> 
-  Eff m [SQLITE ()] (Either String ResultSet)
-\end{Verbatim}
-% =================================================
-\subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate requests.
-This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data.
-In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.
-
-A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects.
-By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.
-
-We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
-%TODO: Fix up the labels
-\begin{Verbatim}[samepage]
-data Session : Effect where
-  LoadSession : 
-  SessionID -> 
-  Session (SessionRes SessionUninitialised) 
-          (SessionRes SessionInitialised) 
-          (Maybe SessionData)
-  UpdateSession : 
-    SessionData -> 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) ()
-  CreateSession : 
-    SessionData -> 
-    Session (SessionRes SessionUninitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  DeleteSession : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool 
-  WriteToDB : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) Bool
-  DiscardSessionChanges : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionUninitialised) ()
-  GetSessionID : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionID)
-  GetSessionData : 
-    Session (SessionRes SessionInitialised) 
-            (SessionRes SessionInitialised) 
-            (Maybe SessionData)
-\end{Verbatim}
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {SessionUninitialised};
-  \node[state]         (B) [below of=A] {SessionInitialised};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
-\end{tikzpicture}
-}
-\caption{Session Handler Resource Usage Protocol}
-\label{fig:sessionstates}
-\end{figure}
-Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler.
-In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key.
-The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage.
-This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.
-
-Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
-\begin{Verbatim}
-data SessionStep = SessionUninitialised
-                 | SessionInitialised
-
-data SessionRes : SessionStep -> Type where
-  InvalidSession : SessionRes s  
-  ValidSession   : SessionID -> 
-                   SessionData -> 
-                   SessionRes s
-\end{Verbatim}
-The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}.
-If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.
-
-% =================================================
-
-\section{Type-aware form handling}
-\label{form}
-Programming web applications often involves processing user data, which may then be used in further effectful computations.
-Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.
-
-This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker.
-One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data.
-This may involve sending an integer detailing the current page, which could be used in a query such as:
-
-\begin{Verbatim}
-'SELECT `name`, `address` FROM `staff` LIMIT ' + 
-       page + ', 5';
-\end{Verbatim}
-The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack.
-% Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here
-
-In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information.
-Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.
-
-We begin with a simple example of a form which requests a user's name, and echoes it back.
-Firstly, we define a form handler which echoes back a string provided by the form handler.
-It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.
-
-\begin{Verbatim}
-sayHello : Maybe String -> 
-           FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
-\end{Verbatim}
-
-We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
-
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [(handler args=[FormString], 
-                     effects=[CgiEffect], 
-                     fn=sayHello, 
-                     name="sayHello")
-           ]
-\end{Verbatim}
-
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
-
-\begin{Verbatim}
-showHelloForm : UserForm
-showHelloForm = do
-  addTextBox "Name" FormString Nothing
-  useEffects [CgiEffect]
-  addSubmit sayHello handlers
-\end{Verbatim}
-
-Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.
-
-\begin{Verbatim}
-cgiHello : CGIProg [] ()
-cgiHello = do
-  handler_set <- isHandlerSet
-  if handler_set then do
-    handleForm handlers
-    return ()
-  else do
-    addForm "nameform" "helloform" showHelloForm
-    return ()
-
-main : IO ()
-main = runCGI [initCGIState] cgiHello
-\end{Verbatim}
-When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user.
-Upon submission of the form, the form handler will be invoked, and the name will be used in the output.
-
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
-\subsection{Form Construction}
-\label{formcons}
-Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
-\begin{Verbatim}
-data FormTy = FormString
-            | FormInt
-            | FormBool
-            | FormFloat
-            | FormList FormTy 
-\end{Verbatim}
-Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
-\begin{Verbatim}
-interpFormTy : FormTy -> Type
-interpFormTy FormString = String
-interpFormTy FormInt = Int
-interpFormTy FormBool = Bool
-interpFormTy FormFloat = Float
-interpFormTy (FormList a) = List (interpFormTy a)
-\end{Verbatim}
-%
-In order to specify a form, we once again use \texttt{Effects}.
-By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.
-
-\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation.
-We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.
-
-\begin{Verbatim}
-using (G : List FormTy, E : List WebEffect)
-  data FormRes : List FormTy -> 
-                 List WebEffect -> Type where
-    FR : Nat -> 
-         List FormTy -> 
-         List WebEffect -> 
-         String -> 
-         FormRes G E
-
-  data Form : Effect where
-    AddTextBox : (label : String) -> 
-                 (fty : FormTy) -> 
-                 (Maybe (interpFormTy fty)) -> 
-                 Form (FormRes G E) 
-                      (FormRes (fty :: G) E) () 
-   ...
-
-    Submit : (mkHandlerFn ((reverse G), E)) ->
-             String -> 
-             Form (FormRes G E) (FormRes [] []) String
-\end{Verbatim}
-The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.
-
-We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form.
-The resource allows us to record the types associated with each form element and the HTML required to display the form.
-he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.
-
-By parameterising the resource over a list of the types associated with each element and the effects supported by the handler function, we may ensure that only a handler function that is compatible with the submitted data may be specified.
-It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.
-
-By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}.
-Additionally, HTML for the form element is generated, and stored in the resource.
-The generated HTML is subsequently returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.
-
-To specify a form instance, we define a function of type \texttt{UserForm}:
-\begin{Verbatim}
-  UserForm : Type
-  UserForm = EffM m [FORM (FormRes []) 
-                          (FormRes [])] String
-\end{Verbatim}
-All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation.
-As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.
-
-Handlers may only be associated with a form if they have argument types corresponding to the types associated with the form elements.
-Additionally, we wish to name the function in order for it to be serialised, whilst requiring a proof that the specified name is associated with the function.
-If this were not the case, it would be possible to specify a function which satisfies the type requirement, without guaranteeing that it the serialised data corresponded to that function, thus rendering the check pointless.
-
-Before associating a handler function with the form, we must specify the effects available to the handler.
-This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource.
-By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form.
-\begin{Verbatim}
-  useEffects : (effs : List WebEffect) ->
-               EffM m [FORM (FormRes G E)] 
-                      [FORM (FormRes G effs)] ()
-  useEffects effs = (UseEffects effs)
-\end{Verbatim}
-Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}.
-This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of additional effects within an effectful computation.
-%
-We may specify a handler function of type \texttt{FormHandler}:
-\begin{Verbatim}
-  FormHandler : List EFFECT -> Type
-  FormHandler effs = Eff IO effs ()
-\end{Verbatim}
-In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
-\begin{Verbatim}
-  addSubmit : 
-    (f :  mkHandlerFn ((reverse G), E)) ->
-    (fns : HandlerList) ->
-    {default tactics 
-      { applyTactic findFn 100; solve; }
-      prf : FnElem f fns} ->
-    EffM m [FORM (FormRes G E)]
-           [FORM (FormRes [] [])] 
-           String
-  addSubmit f handlers {prf} = (Submit f name)
-    where name : String
-          name = getString' f handlers prf          
-\end{Verbatim}
-
-Let us look at each aspect of this function in turn.
-Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with \texttt{useEffects}.
-Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
-\begin{Verbatim}
-MkHandlerFnTy : Type
-MkHandlerFnTy = (List FormTy, List WebEffect)
-
-mkHandlerFn' : List FormTy -> List WebEffect -> Type
-mkHandlerFn' [] effs = FormHandler (interpWebEffects effs) 
-mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
-                              mkHandlerFn' xs effs
-
-mkHandlerFn : MkHandlerFnTy -> Type 
-mkHandlerFn (tys, effs) = mkHandlerFn' tys effs 
-\end{Verbatim}
-The \texttt{mkHandlerFn} function takes a tuple describing the arguments and web effects available to the handler function.
-When constructing the function type, we encase all arguments withing \texttt{Maybe} type, in order to handle failure should the supplied data fail to parse to the specified type.
-
-To store a reference to a handler function, we use the \texttt{HandlerFn} type:
-\begin{Verbatim}
-HandlerFn : Type
-HandlerFn = (ft ** (mkHandlerFn ft, String))
-\end{Verbatim}
-%
-The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function.
-We also store a unique string identifier, which is used to serialise a reference to the handler function.
-
-In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
-\begin{Verbatim}
-syntax 
-  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
-  ", name=" [name] = ((args, effs) ** (fn, name))
-\end{Verbatim}
-We may then define handlers in a more intuitive fashion, without being concerned with the implementation details.
-This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
-\begin{Verbatim}
-handler args=[FormString], 
-        effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
-\end{Verbatim}
-
-We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
-\begin{Verbatim}
-HandlerList : Type
-HandlerList = List HandlerFn
-\end{Verbatim}
-To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
-\begin{Verbatim}
-  data FnElem : mkHandlerFn ((reverse G), E) -> 
-                HandlerList -> Type where
-
-       FnHere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
-       FnThere : {xs : HandlerList, f : 
-                 mkHandlerFn ((reverse G), E)} ->
-               FnElem f xs -> FnElem f (x :: xs)
-\end{Verbatim}
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function.
-\texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list.
-%TODO: this should probably be rewritten
-We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist.
-The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function.
-If the automated proof search fails, compilation will fail.
-
-Finally, we must serialise the argument types, supported effects, and return type of the handler function, in order to allow the form data to be correctly deserialised and ensure that the correct handler function is executed on the server.
-
-Although sending details of the handler function to the client may appear to be a security risk, we envisage that the use of symmetric encryption or a cryptographic nonce would alleviate this.
-Ultimately, we hope to implement a web server with persistent state, which would eliminate the need for serialisation altogether.
-
-Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
-The generated metadata describing the handler function is serialised as a hidden HTML field.
-
-\subsection{Form Handling}
-\label{formhandling}
-Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation.
-This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.
-
-With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier.
-In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers.
-We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.
-
-With this proof, we may then use the \texttt{with} construct, inspired by \textit{views} in Agda \cite{mcbride.mckinna:viewfromleft}, to rewrite the arguments on the left hand side.
-This allows us to construct a function which, given the handler stored in the list of handlers, the data required to construct the function type and the \texttt{MkHandlerFnTy} deserialised from the form, determines whether the two \texttt{MkHandlerFnTy}s are decidably equal.
-If so, we may rewrite this on the left hand side, and therefore demonstrate that the recorded function may also be used to handle the form data.
-If not, the computation is abandoned and an error is shown.
-% I gave it a shot, right?
-:P
-\begin{Verbatim}
-checkFunctions : (reg_fn_ty : MkHandlerFnTy) -> 
-                 (frm_fn_ty : MkHandlerFnTy) -> 
-                 mkHandlerFn reg_fn_ty -> 
-                 Maybe (mkHandlerFn frm_fn_ty)
-checkFunctions reg_ty frm_ty reg_fn with 
-                             (decEq reg_ty frm_ty)
-  checkFunctions frm_ty frm_ty reg_fn | Yes refl = 
-                                            Just reg_fn
-  checkFunctions reg_ty frm_ty reg_fn | No _ = Nothing
-\end{Verbatim}
-We may then parse the arguments according to the types specified by the handler function, and then apply the arguments to the handler function through the use of partial application.
-Finally, we may run the handler function, ensuring that all updates made to the CGI state are propagated.
-
-%-----------------------------
-%-----------------------------
-
-\section{Extended Example: Message Board}
-\label{messageboard}
-Using these components, we may now consider a larger example of a message board which allows users to register, log in, view and create threads, and list and create new posts in threads.
-
-Firstly, we create a database schema in which to record information stored by the message board.
-We create three tables: \texttt{Users}, which contains a unique User ID, usernames and passwords; \texttt{Threads}, which contains a unique thread ID, a title, and the ID of the user who created the thread; and \texttt{Posts}, which contains a unique post ID, the ID of the thread to which each post belongs, the content of the post, and the ID of the user that created the post.
-
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page.
-Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}.
-\subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
-\begin{Verbatim}
-main : IO ()
-main = do (runCGI [initCGIState, 
-                  InvalidSession, ()] 
-           handleRequest)
-          pure ()
-\end{Verbatim}
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
-\begin{Verbatim}
-handleRequest : CGIProg [SESSION 
-  (SessionRes SessionUninitialised), 
-  SQLITE ()] ()
-handleRequest = do 
-  handler_set <- isHandlerSet
-  if handler_set then do
-    lift' (handleForm handlers)
-    Effects.pure ()
-  else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
-    handleNonFormRequest action (map strToInt thread_id)
-\end{Verbatim}
-\subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts.
-One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
-\begin{Verbatim}
-newThreadForm : UserForm
-newThreadForm = do
-  addTextBox "Title" FormString Nothing
-  addTextBox "Post Content" FormString Nothing 
-  useEffects [CgiEffect, SessionEffect, SqliteEffect]
-  addSubmit handleNewThread handlers
-\end{Verbatim}
-The form consists of two text boxes: one for the title of the thread, and one for the content of the first post.
-Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value.
-The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}.
-The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
-\begin{Verbatim}
-handlers : HandlerList
-handlers = [
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleRegisterForm, 
-          name="handleRegisterForm"),
-
- (handler args=[FormString, FormString], 
-          effects=[CgiEffect, SessionEffect, SqliteEffect], 
-          fn=handleNewThread, 
-          name="handleNewThread"),
-     ...
-  ]
-\end{Verbatim}
-Creating a new thread requires a user to be logged in, so that the thread starter may be recorded in the database.
-In order to do this, we make use of the session handler.
-We define a function \texttt{withSession}, which attempts to retrieve the session associated with the current request, and if it exists, executes a function which is passed the associated session data.
-If not, then a failure function is called instead.
-Should the form handler function be called with invalid arguments, an error is shown.
-\begin{Verbatim}
-handleNewThread : 
-  Maybe String -> Maybe String -> 
-  FormHandler [CGI (InitialisedCGI TaskRunning), 
-               SESSION (SessionRes SessionUninitialised), 
-               SQLITE ()]  
-handleNewThread (Just title) (Just content) = do 
-  withSession (addNewThread title content) notLoggedIn
-  pure ()
-handleNewThread _ _ = do 
-  outputWithPreamble "<h1>Error</h1><br />There was 
-       an error posting your thread."
-  pure ()
-\end{Verbatim}
-Once we have loaded the session data from the database, we then check whether the \texttt{UserID} variable is set, which demonstrates that a user has successfully logged into the system, and allows us to use the ID in subsequent computations.
-The database operation to insert the thread into the database is performed by the \texttt{threadInsert} function, which makes use of the \texttt{executeInsert} library function.
-This function abstracts over the low-level resource usage protocol, enabling for provably-correct database access without the excess boilerplate code.
-In addition, the function also returns the unique row ID of the last item which was inserted, which may be used in subsequent computations.
-In the case of the message board, we use this to associate the first post of the thread with the thread being inserted.
-\begin{Verbatim}
-threadInsert : Int -> 
-               String -> 
-               String -> 
-               Eff IO [SQLITE ()] Bool
-threadInsert uid title content = do
-  let query = "INSERT INTO `Threads` 
-    (`UserID`, `Title`) VALUES (?, ?)"
-  insert_res <- (executeInsert DB_NAME query 
-    [(1, DBInt uid), (2, DBText title)]
-  case insert_res of
-    Left err => pure False
-    Right thread_id => postInsert uid thread_id content
-\end{Verbatim}
-\subsection{Listing Threads}
-Listing the threads in the database is achieved through the use of the \texttt{executeSelect} library function, which in turn returns either a \texttt{ResultSet} or an error. 
-\begin{Verbatim}
-getThreads : Eff IO [SQLITE ()] (Either String ResultSet)
-getThreads = 
-  executeSelect DB_NAME query [] collectThreadResults
- where query = "SELECT `ThreadID`, `Title`, `UserID`, 
-   `Username` FROM `Threads` NATURAL JOIN `Users`"
-\end{Verbatim}
-Once the result set has been retrieved, we may iterate through the results and output them to the page, including a link to a page which shows the posts associated with the thread.
-Since we know the structure of the returned row from designing the query, we may pattern match on each returned row to make use of the returned values.
-\begin{Verbatim}
-traverseThreads : ResultSet -> 
-  Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-traverseThreads [] = pure ()
-traverseThreads (x::xs) = do traverseRow x
-                             traverseThreads xs
-  where traverseRow : List DBVal -> 
-           Eff IO [CGI (InitialisedCGI TaskRunning)] ()
-        traverseRow ((DBInt thread_id)::
-                     (DBText title)::
-                     (DBInt user_id)::
-                     (DBText username)::[]) =
-           (output $ "<tr><td>
-            <a href=\"?action=showthread&thread_id=" ++ 
-            (show thread_id) ++ "\">" ++ 
-            title ++ "</a></td><td>" ++ 
-            username ++ "</td></tr>") 
-        traverseRow _ = pure ()
-\end{Verbatim}
-\subsection{Authentication}
-Once a user submits the login form, the associated handler queries the database to ascertain whether a user with the given username and password exists through a call to the \texttt{authUser} function.
-If so, then the session handler is invoked, and a session is initialised with the user ID retrieved from the database.
-The session ID is then set as a cookie using the CGI effect, so that it may be used in subsequent requests.
-Any failures, for example with creating a new session or querying the database, are reported to the user.
-\begin{Verbatim}
-handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
-  case auth_res of
-    Right (Just uid) => do
-      set_sess_res <- setSession uid
-      if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
-        pure ()
-      else do
-        lift' (output "Could not set session")
-        pure ()
-    Right Nothing => do
-      lift' (output "Invalid username or password")
-      pure ()
-    Left err => do
-      lift' (output $ "Error: " ++ err)
-      pure ()
-\end{Verbatim}
-Implementations for the insertion and display of posts, alongside registration, are similar in form to those described above, and are as such omitted from this paper.
-
-Although we have described a relatively simple application, we have shown that through the use of embedded domain-specific languages, we may write verified code that fails to compile should resources be incorrectly accessed.
-Additionally, we have used the form handling mechanism to simply handle the arguments passed by the user.
-Importantly, we have shown that dependent types can be used to increase confidence in an (albeit simplified) real-world application, without requiring developers to supply proofs or indeed work explicitly with dependent types.
-%-----------------------------
-%-----------------------------
-
-\section{Related Work}
-\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser.
-The library also provides support for cookies and basic form handling.
-
-\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods.
-In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs.
-Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs.
-Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations.
-Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.
-
-\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction.
-Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven.
-This means that upon a component being edited, a computation occurs, given the state of the current form.
-This is then saved to be used in future computations.
-Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted.
-This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.
-
-The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another.
-This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion.
-These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application.
-Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.
-
-Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs.
-Through the use of the Ur\/Web DSL, developers may generate well-typed code from records.
-By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure.
-Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.
-
-WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres.
-WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes.
-The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables.
-These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy.
-Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing.
-WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation.
-We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.
-
-%-----------------------------
-%-----------------------------
-
-\section{Conclusions}
-Dependently-typed languages have, remained relatively unused for practical purposes outside of academia.
-Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.
-
-Our framework, through the use of dependent types, provides several static guarantees.
-Data submitted by users is inherently unsafe due to the possibility of attacks such as SQL injection, and this particular threat is ameliorated due to elements to be associated with specific types during form construction.
-This immediately eliminates the possibilities of SQL injection attacks on non-string types.
-Since failures are handled transparently, no runtime errors are output to the browser, meaning that attackers may not use such information to aid attacks.
-Additionally, since checking is performed on the types of the form elements and the types of arguments accepted by the handler, it is impossible to associate a form with a handler incompatible with the submitted data.
-
-Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
-Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
-By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
-
-In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
-With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error.
-
-Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
-\subsection{Further Work}
-We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
-
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
-
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
-We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
-
-Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
-Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
-Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
-
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation.
-Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error.
-This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
-Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
-Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
-
-Dependently-typed languages provide great promise for the construction of secure and correct programs.
-Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
-% Cryptography bindings would be really good.
-% Integration with a web server instead of doing everything over CGI
-% Improvements to form handling system
-% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors
+\input{introduction}
+\input{effects}
+\input{protocols}
+\input{forms}
+\input{messageboard}
+\input{conclusion}
 

### 2013-11-15 paper/protocols.tex
Paper revisions

@@
 In \idris{}, types are first-class, meaning that they may be treated like other
-terms in computations. We may therefore define the following type synonym,
-which we may use in order to make use of the CGI framework:
+terms in computations. We may therefore define the following type synonym, used within the CGI section of the framework to denote an effectual CGI program:
 
@@
 engine often used as simple, structured storage for larger applications. We
-make use of SQLite due to its simplicity, although we envisage that these
+make use of SQLite to demonstrate a resource usage protocol for database access due to its simplicity, although we envisage that these
 concepts would be applicable to more complex database management systems.
@@
 Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
-access, which we have implemented for the SQLite library. This is encapsulated
+access, which we have implemented for the SQLite library. Although some additional states are used to capture failing computations, these are omitted from the diagram. The protocol is encapsulated
 by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
@@
   FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
-                                                          (SQLitePSSuccess Bound))           (Either BindError ())
+                                                          (SQLitePSSuccess Bound))           (Maybe BindError)
   ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
@@
                       Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String
+
+  CleanupPSFail     : Sqlite (SQLitePSFail)             ()                                   ()
+  CleanupBindFail   : Sqlite (SQLiteFinishBindFail)     ()                                   ()
 
@@
 %TODO: Fix up the labels
-\begin{figure}[h]
+\begin{figure}
 \centering
@@
   \tikzset{every loop/.style={min distance=10mm,looseness=5}}
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]
+  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, align=center, node distance=7em, font={\sffamily\bfseries}]
 
@@
   \node[initial,state] (A)              {Uninitialised};
-  \node[state]         (B) [below of=A] {ConnectionOpened};
-  \node[state]         (C) [below of=B] {PreparedStatementOpen};
-  \node[state]         (D) [below of=C] {PreparedStatementBinding};
-  \node[state]         (E) [below of=D] {PreparedStatementBound};
-  \node[state]         (F) [below of=E] {PreparedStatementExec};
+  \node[state]         (B) [below of=A] {SQLiteConnected};
+  \node[state]         (C) [below of=B] {SQLitePSSuccess Binding};
+  \node[state]         (D) [below of=C] {SQLitePSSuccess Bound};
+  \node[state]         (E) [below left of=D, xshift=-4em] {SQLiteExecuting \\ ValidRow};
+  \node[state]         (F) [below right of=D, xshift=4em] {SQLiteExecuting \\ InvalidRow};
 
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
-        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
-        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
-        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
-        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
-        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
-        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
-        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
+  \path (A) edge[stateEdge]   node[edgeLabel, xshift=2.5em] {\texttt{openDB}} (B)
+        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=4.5em] {\texttt{closeDB, connFail}} (A)
+        (B) edge[stateEdge]   node[edgeLabel, xshift=5em] {\texttt{prepareStatement}} (C)
+        (C) edge[stateEdge]   node[edgeLabel, xshift=3.25em, yshift=-0.75em] {\texttt{finishBind}} (D)
+        (D) edge[stateEdge]   node[edgeLabel, text width=2cm, xshift=-4.5em, yshift=-1em] {\texttt{executeStatement, nextRow}} (F)
+        (D) edge[stateEdge]   node[edgeLabel, xshift=3em] {} (E)
+        (E) edge[stateEdge, bend left=35]   node[edgeLabel, xshift=3em, yshift=-5em] {\texttt{finaliseValid}} (A)
+        (F) edge[stateEdge, bend right=35]   node[edgeLabel, xshift=-3em, yshift=-5em] {\texttt{finaliseInvalid}} (A)
+        (E) edge[stateEdge]   node[edgeLabel, yshift=-1em] {\texttt{nextRow}} (F)
+        (F) edge[stateEdge, bend left=35]   node[edgeLabel, yshift=-1em] {\texttt{reset}} (E)
+        (E) edge[stateEdge, loop below]   node[edgeLabel] {\texttt{nextRow}} (E);
+
 \end{tikzpicture}
@@
 \useverb{sqliteexecuting}
-
+\noindent
 Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.
@@
 \noindent
-In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct. The \texttt{nextRow} function 
+In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct.
+The \texttt{nextRow} function will either fetch a row into the buffer, or indicate that a row could not be loaded, as shown by its type:
+
+\begin{SaveVerbatim}{nextrow}
+nextRow : EffM IO [SQLITE (SQLiteExecuting ValidRow)] 
+                  [SQLITE (Either (SQLiteExecuting InvalidRow)
+                                  (SQLiteExecuting ValidRow))] 
+                  StepResult
+\end{SaveVerbatim}
+\useverb{nextrow}
+
+\noindent
+The \texttt{if\_valid} construct provides failure checking functionality, allowing different operations to be performed depending on whether or not a row was successfully fetched for processing.
 %If a failure happens at any point during the computation, the resource is
@@
 
-\noindent
-Each call to \texttt{nextRow}, which is a wrapper around the underlying
-\texttt{sqlite3\_step()} C library function, returns a result of type
-\texttt{StepResult}, which is then reflected in the resource. Calls to
-\texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult}
-is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore
-statically guarantee that only calls that will return a valid result are
-executed.
-
 By incorporating pointers to open connections and prepared statements into the
 resource associated with the effect, we introduce a further layer of
-abstraction, in order to hide implementation details from the developer and
-encourage cleaner, less error-prone code.
+abstraction, which hides implementation details from the developer and
+encourages less error-prone code.
 
@@
 String ->
-Eff IO [SQLITE ()] (Either String (List (String, String)))
+Eff IO [SQLITE ()] (Either QueryError (List (String, String)))
 
@@
 testSelect : String -> Eff IO [SQLITE ()] 
-             (Either String (List (String, String)))
+             (Either QueryError (List (String, String)))
 testSelect dept = do
-  open_db <- openDB "people.db"
-  if open_db then do
+  db_res <- openDB "people.db"
+  if_valid then do
     let sql = "SELECT name, address FROM `staff` 
@@
     sql_prep_res <- prepareStatement sql
-    if sql_prep_res then do 
-      startBind
+    if_valid then do 
       bindText 1 dept
-      finishBind
-      beginExecution
+      bind_res <- finishBind
+      if_valid then do
+        executeStatement
         results <- collectResults
-      finaliseStatement
+        finaliseInvalid
         closeDB
-      return $ Right results
-    else do err <- stmtFail
-            return $ Left err
-  else do err <- connFail
-          return $ Left err 
+        Effects.pure $ Right results
+      else do
+        cleanupBindFail
+        Effects.pure $ Left (getBindError bind_res)
+    else do
+      cleanupPSFail
+      Left $ getQueryError sql_prep_res
+  else do 
+    Effects.pure $ Left (getQueryError db_res)
 \end{SaveVerbatim}
@@
 
-\texttt{CollectResults} makes a call to
-\texttt{nextRow} in order to make the next row of the result set available for
-processing, and uses \texttt{getColumnText} and \texttt{getColumnInt}
+\texttt{CollectResults} operates on the row currently held in the buffer.
+Firstly, the function checks, through the use of the \texttt{if\_valid} construct, that there is a valid row in the buffer, and if so uses the \texttt{getColumnText} and \texttt{getColumnInt} functions
 to retrieve the data from the database. This function is then
-recursively called until there are no more rows to process.
+called recursively until there are no more rows to process.
 
 \begin{SaveVerbatim}{collectresty}
-
-collectResults : Eff IO 
-            [SQLITE (SQLiteRes PreparedStatementExec)] 
+collectResults :
+  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
+                          (SQLiteExecuting ValidRow))] 
+          [SQLITE (SQLiteExecuting InvalidRow)] 
           (List (String, Int))
@@
 \begin{SaveVerbatim}{collectres}
-collectResults = do
-  step_result <- nextRow
-  case step_result of
-      StepComplete => do name <- getColumnText 1
+collectResults = 
+  if_valid then do name <- getColumnText 1
                    age <- getColumnInt 2
+                   step_res <- nextRow
                    xs <- collectResults
                    return $ (name, age) :: xs
-      NoMoreRows => return []
-      StepFail => return []
+  else return []
 
@@
 \noindent
-Using these, we may implement a function, \texttt{ExecuteSelect}, which, given
+Using these, we may implement a function, \texttt{executeSelect}, which, given
 a query, a list of variables to bind and their associated indices within the
@@
   String -> String -> List (Int, DBVal) -> 
-  Eff m [SQLITE (SQLiteRes PreparedStatementExec)] 
-                 (List DBVal) -> 
-  Eff m [SQLITE ()] (Either String ResultSet)
+  (Eff IO [SQLITE (SQLiteExecuting ValidRow)] (List DBVal)) -> 
+  Eff IO [SQLITE ()] (Either QueryError ResultSet)
 

### 2013-11-17 paper/conclusion.tex
Proof reading/trimming pass over the paper

@@
 
-To date,
-dependently-typed languages have remained relatively unused for practical
-purposes. Advances in embedded domain-specific languages
-have allowed many of the advantages of such languages, including the additional
-guarantees given by the successful compilation of a program (assuming the use
-of specific enough types) to be adopted without additional proofs.
+Dependently-typed languages promise to support machine checkable program
+correctness proofs, but to date they have remained relatively unused for
+practical purposes. By using embedded domain-specific languages, we can
+abstract away some of the complexities of creating correctness proofs and
+provide expressive libraries, giving guarantees by the successful compilation
+of a program (assuming the use of specific enough types) without additional
+proofs.
 
-Our framework, through the use of dependent types, provides several static
-guarantees. Data submitted by users is inherently unsafe due to the possibility
-of attacks such as SQL injection, and this particular threat is ameliorated due
-to elements to be associated with specific types during form construction. This
+Our framework provides several static
+guarantees. Data submitted by users is inherently unsafe and
+means systems are vulnerable to attacks
+such as SQL injection. This particular threat is ameliorated due
+to elements being associated with specific types during form construction. This
 immediately eliminates the possibilities of SQL injection attacks on non-string
@@
 
-Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations.
-Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly.
-By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly.
-Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.
+Many external libraries also follow (unchecked or dynamically checked) 
+resource usage protocols.
+Incorrect usage is however still possible, for
+example by forgetting to release acquired resources or failing to
+initialise a library correctly. By creating high-level bindings to these
+libraries, however, we may statically enforce these resource-usage protocols,
+ensuring that the libraries are used correctly. Whilst previous work has
+demonstrated that this is possible through the use of embedded DSLs
+\cite{brady:edsl} and dependent algebraic effects \cite{brady:effects},
+this paper has provided more substantial examples of real-world applications.
 
-In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers.
-With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available.
-Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error or memory access violation.
+In particular, the framework guarantees that it is not possible for a CGI 
+application to
+produce an internal server error due to content being written to the remote
+host prior to headers. With regard to database access, we may statically
+guarantee that library calls are made in the correct order, and calls to
+retrieve rows of data are made only when more data is available. Additionally,
+by recording any failures in the resource associated with the database effect,
+we may ensure that any failures at certain stages in the process may be
+correctly dealt with, and not propagate to further library calls---if a
+prepared statement fails to compile, for example, the framework guarantees that
+it will not be executed, without causing a fatal runtime error or memory access
+violation.
+%
+Enforcing resource usage protocols also guards against common programmer
+errors, saving debugging time by identifying errors at compile time.
 
-Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
 \subsection{Further Work}
-We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.
 
-There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
-Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence in their security.
+We have shown that embedded domain-specific languages using dependent types and
+algebraic effects can be used to increase confidence in web applications by
+providing additional static guarantees about runtime behaviour, but much more
+can be done using the same approach.
 
-Whilst the use of CGI allows for experimenting with the use of dependent types in a real-world scenario such as web programming, there remain practical considerations about its scalability, as a separate process must be created for each individual request.
-We believe that the use of FastCGI may alleviate this, but ultimately, we would like to create a web server written in \idris{}, which would make more efficient usage of resources.
+There are many other applications which make use of specific resource usage
+protocols, for example popular libraries such as
+\textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
+Applying a similar approach would allow for sensitive programs requiring
+cryptographic routines to be written using a language with full dependent
+types, in turn adding an extra layer of confidence in their security.
 
-Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner.
-Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness.
-Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.
+Whilst the use of CGI allows for experimenting with the use of dependent types
+in a real-world scenario such as web programming, there remain practical
+considerations about its scalability, as a separate process must be created for
+each individual request. We believe that the use of FastCGI may alleviate this,
+but ultimately, we would like to create a web server written in \idris{}, which
+would make more efficient usage of resources.
 
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used during compilation.
-Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures.
-If data structures within the program do not conform to a given database schema, for example, then the program will not type-check and as such will produce a compile-time error.
-This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step.
-Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness.
-Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
+Since at this stage we have concentrated on the use of dependent types for
+enforcing resource usage protocols and type-safe form handling, we currently
+handle the generation of HTML in an unstructured manner. Future work will
+entail a DOM library to facilitate the generation and manipulation of HTML, in
+turn giving stronger guarantees about its correctness. Other planned features
+include a template system, allowing for web pages to be automatically generated
+from data, and an object-relational mapping system which will allow users to
+manipulate records which can then be automatically written to the database,
+instead of having to update tables manually via SQL queries.
 
-Dependently-typed languages provide great promise for the construction of secure and correct programs.
-Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
+Type providers, as originally implemented in F\# \cite{msr:tp}, are an
+interesting method by which external data sources may be used to import
+external information, such that it may be used during compilation. 
+In this way, it becomes possible to use the extra type
+information to statically ensure the validity of artefacts such as SQL
+queries and data structures. If data structures within the program do not
+conform to a given database schema, for example, then the program will not
+type-check.  This has been implemented for 
+\idris{} 
+\cite{christiansen:dtp}, exploiting the fact that types can be
+calculated by functions to avoid generating extra code in the type provider
+step. Dependent type providers additionally have stronger safety guarantees as
+they may not generate unchecked code, but at the same time this is matched by a
+decrease in expressiveness. Nonetheless, such techniques provide a promising
+mechanism to verify the semantic soundness of programs and we look to
+investigate their integration in further work.
+
+Dependently-typed languages provide great promise for the construction of
+secure and correct programs. Through the use of embedded domain-specific
+languages, we hope that more developers may benefit from the extra guarantees
+afforded by dependent types, resulting in more stable, secure applications.
+
 % Cryptography bindings would be really good.

### 2013-11-17 paper/effects.tex
Proof reading/trimming pass over the paper

@@
 the resource type. \texttt{Put} returns nothing, but has the effect of updating
-the resource. In order to make an effect usable, we implement a handler
+the resource. To make an effect usable, we implement a handler
 for a computation context by making an instance of the following class:
@@
 class Handler (e : Effect) (m : Type -> Type) where
-     handle : res -> 
-              (eff : e res res' t) -> 
+     handle : res -> (eff : e res res' t) -> 
               (k: res' -> t -> m a) -> m a
@@
 times a handler may invoke the continuation (raising an exception, for example,
-will not call the continuation). Reading and updating states in handled
-for all computation contexts \texttt{m} as follows:
+will not invoke it). Reading and updating states is handled
+for all computation contexts \texttt{m}:
 
@@
 \end{SaveVerbatim}
+\useverb{statehandler}
 
@@
 
+\noindent
 A concrete effect is simply an algebraic effect type paired with its current
@@
 
+\begin{SaveVerbatim}{fileeff}
+{-                                             { Input resource type }   { Output resource type } { Value } -}
+
+data FileIO : Effect where
+     Open      : String -> (m : Mode) -> FileIO     ()                    (Either () (OpenFile m)) ()
+     Close     :                         FileIO     (OpenFile m)          ()                       ()
+
+     ReadLine  :                         FileIO     (OpenFile Read)       (OpenFile Read)          String
+     WriteLine : String ->               FileIO     (OpenFile Write)      (OpenFile Write)         ()
+     EOF       :                         FileIO     (OpenFile Read)       (OpenFile Read)          Bool
+\end{SaveVerbatim}
+
+\begin{figure*}[t]
+\begin{center}
+\useverb{fileeff}
+\end{center}
+\caption{File Protocol Effect}
+\label{fig:fileeffect}
+\end{figure*}
 More generally, a program might modify the set of effects available.
@@
 example is a file access protocol, where a file must be opened before it
-is read or written, and a file must be closed on exit:
+is read or written, and a file must be closed on exit.
 
-\begin{SaveVerbatim}{fileeff}
-
-data FileIO : Effect where
-     Open  : String -> (m : Mode) -> 
-             FileIO () (OpenFile m) ()
-     Close : FileIO (OpenFile m) () ()
-
-     ReadLine  : FileIO (OpenFile Read)  
-                        (OpenFile Read) String
-     WriteLine : String -> 
-                 FileIO (OpenFile Write) 
-                        (OpenFile Write) ()
-     EOF       : FileIO (OpenFile Read)  
-                        (OpenFile Read) Bool
-
-\end{SaveVerbatim}
-\useverb{fileeff}
-
-\noindent
+Figure \ref{fig:fileeffect} shows how the protocol is encoded as an
+effect.
 Note that the types of the input and output resources describes how resource
-state changes in each operation: opening a file changes an empty resource to
-a resource containing an open file; reading a line is only possible if the
+state changes in each operation: opening a file may fail, so
+changes an empty resource to
+a resource containing either a unit or an open file; 
+reading a line is only possible if the
 resoource is a file open for reading, etc.
 The handler for this effect for an \texttt{IO} computation context will
-execute the required primitive I/O actions, as well as throwing an exception
-if any operation fails.
+execute the required primitive I/O actions.
 
@@
 testFile : Eff IO [FILE_IO (), STDIO] () 
-testFile = catch (do open "testFile" Read
-                     str <- readLine
+testFile = do open "testFile" Read
+              if_valid then do str <- readLine
                             close
-                     putStrLn str)
-            (\err => putStrLn ("Error: " ++ show err))
+                            putStrLn str
+                       else putStrLn "Error"
 
@@
 the resource associated with the File I/O will be in the same state on entry
-and exit.
-Therefore, attempting to write to the file, or failing to open or close the
+and exit. We use \texttt{if\_valid} to handle possible failure---this is
+a function provided by the \texttt{Effects} library which checks whether
+a resource of type \texttt{Either a b} indicates failure (\texttt{Left a})
+or success (\texttt{Right b}) and branches and updates the resource
+accordingly.
+Therefore, attempting to write to the file, failing to check for
+success, or failing to open or close the
 file, would cause a \emph{compile-time} error.

### 2013-11-17 paper/forms.tex
Proof reading/trimming pass over the paper

@@
 then be used in further effectful computations. Data submitted using a form is
-transmitted over the internet as a string as part of an HTTP request, which
+transmitted as a string as part of an HTTP request, which
 traditionally involves losing associated type information.
@@
 
-In this section, we present a mechanism by which we introduce a DSL
+In this section, we present a DSL
 for the creation of web forms which preserve type information, implemented
-as a dependent algebraic effect. Once the form has
+as a resource-dependent algebraic effect. Once the form has
 been submitted, retrieved information is passed directly to a
@@
 provided by the form handler. It has one argument of type \texttt{Maybe
-String}, which accounts for the possibility that the user may have specified
-incorrect data within the form.
+String}, which accounts for the possibility that the user may have 
+provided invalid data:
 
@@
 
-sayHello : Maybe String -> 
+echo : Maybe String -> 
      FormHandler [CGI (InitialisedCGI TaskRunning)]
-sayHello (Just name) = output ("Hello, " ++ name ++ "!")
-sayHello _ = output "Error!"
+echo (Just name) = output ("Hello, " ++ name ++ "!")
+echo _ = output "Error!"
 
@@
 
+\noindent
 We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:
@@
                     effects=[CgiEffect], 
-                    fn=sayHello, 
-                    name="sayHello"]
+                    fn=echo, 
+                    name="echo"]
 
@@
 
-We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.
+\noindent
+We also define a form to take in a name from the user, and specify that it
+should use the \texttt{echo} handler.
 
@@
   useEffects [CgiEffect]
-  addSubmit sayHello handlers
+  addSubmit echo handlers
 
@@
 
-In order to specify a form, we once again use \texttt{Effects}. By recording
+\noindent
+Again, we use \texttt{Effects} to build a form.
+By recording
 the type of each form element as it is added in the type of the form, we may
 statically ensure that the user-supplied handler function is of the correct
-type to handle the data supplied by the form: the specification of an
+type to handle the data supplied by the form: using an
 incompatible handler will result in a compile-time type error. The \texttt{Form}
@@
 \noindent
-These make use of the resource associated with the effect, \texttt{FormRes}, to
-construct the form. Adding a field such as a text box adds a new type, \texttt{fty}
-to the list of field types, carried in the resource. When the form is complete,
-the \texttt{Submit} operation adds a submit button and returns the
-HTML text for the form, flushing the list of field
-types, and using it to construct the type for an appropriate handler function.
+These use the resource associated with the effect, \texttt{FormRes}, to
+construct the form. Adding a field such as a text box adds a new type,
+\texttt{fty} to the list of field types, carried in the resource. When the form
+is complete, the \texttt{Submit} operation adds a submit button and returns the
+HTML text for the form, flushing the list of field types, and using it to
+construct the type for an appropriate handler function.
 %
@@
 \noindent
-All forms are required to include a submit button, as mandated by the
-requirement that
-
+%All forms are required to include a submit button, as mandated by the
+%requirement that 
+%
 The input and output resource contains an empty list of types, which means that
@@
 proof that the specified name is associated with the function. If this were not
-the case, it would be possible to specify a function which satisfies the type
-requirement, without guaranteeing that it the serialised data corresponded to
-that function, thus rendering the check pointless.
+required, it would be possible to use a function satisfying the type
+requirement, without guaranteeing that the serialised data corresponded to
+that function.
 
 Before associating a handler function with the form, we must specify the
-effects available to the handler. This is done through the use of the
+effects available to the handler. This is done with
 \texttt{useEffects}, which updates the list of effects in the type of the form
@@
 
-\begin{SaveVerbatim}{mkhandlerfn}
+\begin{SaveVerbatim}{mkhandlerfnp}
 
@@
                               mkHandlerFn' xs effs 
+\end{SaveVerbatim}
 
+\begin{SaveVerbatim}{mkhandlerfn}
 mkHandlerFn : MkHandlerFnTy -> Type 
@@
 
+\useverb{mkhandlerfnp}
+
 \useverb{mkhandlerfn}
@@
 web effects available to the handler function. When constructing the function
-type, we encase all arguments withing \texttt{Maybe} type, in order to handle
-failure should the supplied data fail to parse to the specified type.
+type, we wrap all arguments in a \texttt{Maybe}, in order to handle
+failure should the supplied data fail to parse as the required type.
 %
@@
 \noindent
-The \texttt{**} notation denotes a dependent pair, where the type of the second
+The \texttt{**} notation denotes a dependent pair, in which the type of the second
 element of the pair is parameterised over the value of the first element. It is
@@
 In order to abstract away from this implementation detail, we make use of
-\idris{} syntax rewriting rules. This allows us to define the following syntax
-rewrite rule:
+\idris{} syntax rewriting rules. This allows us to define the following:
 
@@
 with one String argument, making use of the CGI effect, associated with the
-\texttt{sayHello} handler function as follows:
+\texttt{echo} handler function as follows:
 
@@
         effects=[CgiEffect], 
-        fn=sayHello, 
-        name="sayHello"
+        fn=echo, 
+        name="echo"
 
@@
 
-To enforce the requirement that a supplied handler function must reside in the
+To enforce the requirement that a supplied handler function must be in the
 list of available handlers, and therefore allow us to retrieve the name with
@@
 %TODO: this should probably be rewritten
-We then use a simple automated proof search to
-automatically generate the proof at compile time, should one exist. The proof
-may then be used in subsequent computations: in our case, we use it to retrieve
+We then use an automatic proof search to
+generate the proof at compile time, should one exist. The proof
+may then be used in subsequent computations: we use it to retrieve
 the unique identifier for the function. If the automated proof search fails,
@@
 a security risk, we envisage that the use of symmetric encryption or a
-cryptographic nonce would alleviate this. Ultimately, we hope to implement a
+cryptographic nonce would mitigate this. Ultimately, we hope to implement a
 web server with persistent state, which would eliminate the need for
@@
 
-Running form construction is achieved as an operation of the CGI
+Running form construction is achieved as an operation of the \texttt{CGI}
 effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
@@
 
-decEq : DecEq t => (x : t) -> (y : t) -> Dec (x1 = x2)
+decEq : DecEq t => (x : t) -> (y : t) -> Dec (x = y)
 
@@
 We may then parse the arguments according to the types specified by the handler
-function, and then apply the arguments to the handler function through the use
-of partial application. Finally, we may run the handler function, ensuring that
+function, and then apply the arguments to the handler function.
+Finally, we may run the handler function, ensuring that
 all updates made to the CGI state are propagated.

### 2013-11-17 paper/introduction.tex
Proof reading/trimming pass over the paper

@@
 such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. A significant drawback, however, is that such languages
+\cite{w3techs:webpls}. However, such languages
 do not provide the same static guarantees about runtime behaviour afforded by
@@
 \noindent
-We then check to see if the connection was successful, and exit
-if not.  This check is optional, so
-it would be possible to omit it. However, this would cause
-problems with later steps.
+We should then check to see if the connection was successful, and exit
+if not:
+%This check is optional, so
+%it would be possible to omit it. However, this would cause
+%problems with later steps.
 
@@
 
-  if (mysqli_connect_errno()) {
-      exit();
-  }
+if (mysqli_connect_errno()) { exit(); }
 
@@
 \noindent
-Even in this small example, there exists a specific resource usage protocol.
+Even in this small example, there exists a precise resource usage protocol
+which must be followed for successful and robust operation.
 Firstly, a connection to the database must be opened. The object-oriented style
@@
 However, in conventional programming languages, there is no way to check
-automatically that a protocol is followed.
+automatically, at compile-time, that a protocol is followed.
 
@@
 %
-The difficulty is 
-that automatic verification by a compiler can be difficult or
+Unfortunately, automatic verification by a compiler can be difficult or
 often impossible, requiring additional proofs to be given by the programmer.
 
-This complexity can be ameliorated through the use of \textit{embedded
-domain-specific languages} (EDSLs), which aim to abstract away the
-complexity of the underlying type system. EDSLs allow domain experts to
-write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.
+This complexity can be addressed through the use of \textit{embedded
+domain-specific languages} (EDSLs) to abstract away the complexity of the
+underlying type system. Embedding a domain-specific language in
+a dependently typed host language
+allows domain experts to write 
+domain-specific code, with the EDSL itself used to provide the proof that the
+code is correct.
 
@@
 \begin{itemize}
-\item A form-handling mechanism, which preserves type information and
-manages user input, therefore
-increasing applications' resilience to attacks such as SQL injection and cross-site scripting.
+\item Representations of CGI, Databases and sessions as
+\textit{resource-dependent algebraic effects}, allowing programs to be accepted
+only when they follow clearly defined resource usage protocols. 
+(Section~\ref{rup})
 
-\item Representations of CGI, Databases and sessions as
-\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
-follow clearly defined resource usage protocols.
+\item Type-safe form handling, preserving type information and managing
+user input, therefore increasing applications' resilience to attacks such as
+SQL injection and cross-site scripting. (Section~\ref{form})
 
-\item A message board application, demonstrating the usage of the framework.
+\item An extended example: a message board application, demonstrating the usage
+of the framework in practice. (Section~\ref{effects})
 
@@
 
-We structure the remainder of this paper as follows. We provide a brief
-overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
-how this may be used to ensure adherence to resource usage protocols for CGI,
-SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
-type-safe form handling in Section ~\ref{form}, implemented using
-\texttt{Effects}; and discuss the larger example
-of a message board system making use of these components in Section
-~\ref{messageboard}.
+We achieve these without extending the host language. Every
+resource protocol we implement is pure \Idris{}, using a library for
+resource-dependent algebraic effects~\cite{brady:effects} and \Idris{}'
+features for supporting domain-specific language implementation such as
+syntax macros and overloading. In particular, this means the same techniques
+can be applied to other resources, and most importantly, combinations
+of resources and DSLs implemented in this way are \emph{composable}.
 
-The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/idris-hackers/IdrisWeb}.
+%We structure the remainder of this paper as follows. We provide a brief
+%overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
+%how this may be used to ensure adherence to resource usage protocols for CGI,
+%SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
+%type-safe form handling in Section ~\ref{form}, implemented using
+%\texttt{Effects}; and discuss the larger example
+%of a message board system making use of these components in Section
+%~\ref{messageboard}.
+
+The code used to implement the framework and all associated examples used in
+this paper is available online at
+\url{http://www.github.com/idris-hackers/IdrisWeb}.
+
 % =================================================

### 2013-11-17 paper/main.tex
Proof reading/trimming pass over the paper

@@
 Dependently-typed languages allow precise types to be used during development,
-facilitating easier reasoning about the operation of programs. However,
+facilitating reasoning about programs. However,
 stronger types bring a disadvantage that it becomes increasingly difficult to
@@
 
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in a high-level languages which uses
-dependent types to enforce certain invariants without additional proof
-obligations.
+Embedded domain-specific languages (EDSLs) can help address this problem by
+introducing a layer of abstraction over more precise underlying types,
+allowing domain-specific code to be written in a high-level language which uses
+dependent types to enforce invariants without imposing additional proof
+obligations on an application programmer.
 

### 2013-11-17 paper/messageboard.tex
Proof reading/trimming pass over the paper

@@
 
-Secondly, we use a GET variable, \texttt{action}, to indicate which page of the
+Secondly, we use a \texttt{GET} variable, \texttt{action}, to indicate which page of the
 message board should be displayed, and pattern-match on these to call the
@@
 which shows all of the posts in a thread, require a second argument,
-\texttt{thread\_id}.  \subsection{Handling requests}
-The entry point to any CGI application is the \texttt{main} function.
-From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations.
+\texttt{thread\_id}.
+
+\subsection{Handling requests}
+The entry point to any CGI application is the \texttt{main} function. From
+here, we run the remainder of the program through a call to \texttt{runCGI},
+which we initialise with empty initial environments for the \texttt{CGI}, 
+\texttt{Session} and
+\texttt{SQLite} effects, so they may be used in further computations.
 
@@
 \noindent
-We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled.
-If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}.
-If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
+We define a function, \texttt{handleRequest}, which firstly determines
+whether submitted form data must be handled, by checking whether a handler
+variable exists.
+If so, then the form handling routine is called, which executes the
+corresponding handler function as specified in Section ~\ref{formhandling}. If
+not, then the \texttt{handleNonFormRequest} function is called, which inspects
+the \texttt{GET} variables in order to display the correct page.
 
@@
 \subsection{Thread Creation}
-We specify four forms: one to handle registration, one to handle logging in,
+We create four forms: one to handle registration, one to handle logging in,
 one to handle the creation of new threads, and one to handle the creation of
@@
 \noindent
-Creating a new thread (show in Figure \ref{fig:handlethread})
+Creating a new thread (shown in Figure \ref{fig:handlethread})
 requires a user to be logged in, so that the thread
@@
 Implementations for the insertion and display of posts, alongside registration,
-are similar in form to those described above, and are as such omitted from this
-paper.
+follow the same structure.
 
 Although we have described a relatively simple application, we have shown that
-through the use of embedded domain-specific languages, we may write verified
+through the use of embedded domain-specific languages, 
+and particularly by encapsulating resource usage protocols in the types,
+we can write verified
 code that fails to compile should resources be incorrectly accessed.

### 2013-11-17 paper/protocols.tex
Proof reading/trimming pass over the paper

@@
 \label{rup}
-In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.
+In this section, we show how three effects 
+(CGI, database access and a simple session handler) may be implemented, and
+describe the benefits of developing programs using this technique over simply
+handling them in \texttt{IO} or as monad transformers.
 
@@
 
-CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client.
-Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error being shown.
+CGI is used to invoke an application on a web server, making use of environment
+variables to convey information gained from an HTTP request and using standard
+output to communicate with the remote client. Importantly, HTTP headers must be
+correctly written to the browser prior to any other output; failure to do so
+will result in an internal server error being shown.
 
-A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
+%A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
 Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
@@
 
-By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly.
-In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}.
-This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request.
-We represent an uninitialised CGI process as the unit type, ().
+By modelling CGI as a resource-dependent algebraic effect, we may enforce a
+resource usage protocol which
+%even though the program may be running in an IO execution context, 
+prevents arbitrary IO from being performed and therefore
+ensures that the headers are written correctly. We
+define an effect, \texttt{Cgi}, and an associated resource,
+\texttt{InitialisedCGI}, parameterised over the current state,
+\texttt{CGIStep}, and containing a
+\texttt{CGIInfo} record which contains information from the request. We
+represent an uninitialised CGI process as the unit type, \texttt{()}.
 
@@
 that can be performed in this state is initialisation: by calling
-\texttt{initialise}, a CGIInfo record is populated, and the state transitions
+\texttt{initialise}, a \texttt{CGIInfo} record is populated, and the state transitions
 to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
@@
 
-Additional operations, including those to query POST and GET variables, are
-omitted in the interest of brevity.
+Additional operations, including those to query \texttt{POST} and \texttt{GET}
+variables, are omitted in the interest of brevity.
 
 User code executes in the \texttt{TaskRunning} state. Several operations, such
-as querying the POST and GET variables, are available in this state, alongside
+as querying the \texttt{POST} and \texttt{GET} variables, are available in this state, alongside
 functions to output data to the web page and append data to the response
@@
 page, with the \texttt{output} and \texttt{addHeader} functions instead
-modifying the CGIInfo record. This data may then be printed at the end of the
+modifying the \texttt{CGIInfo} record. This data may then be printed at the end of the
 program's execution, in accordance with the resource usage protocol.
@@
 In \idris{}, types are first-class, meaning that they may be treated like other
-terms in computations. We may therefore define the following type synonym, used within the CGI section of the framework to denote an effectual CGI program:
+terms in computations. We may therefore define the following type synonym, used
+within the CGI section of the framework to denote an effectful CGI program:
 
@@
 determine attack vectors for SQL injection attacks.
-
-Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
-access, which we have implemented for the SQLite library. Although some additional states are used to capture failing computations, these are omitted from the diagram. The protocol is encapsulated
-by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
-bind and retrieve data types other than \texttt{String}, in the interest of
-brevity).
 
@@
 
-\noindent
+Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
+access, which we have implemented for the SQLite library. Although some
+additional states are used to capture failing computations, these are omitted
+from the diagram.
+The effect implementation is given in Figure \ref{fig:dbeffect}.
+%
 There are three main phases involved in the usage of the SQLite protocol:
@@
 
-We first define the \texttt{SQLiteConnected} resource, which signifies that a successful connection has been made to the database.
-This resource contains a pointer to the database structure which is used in further computations.
+We first define the \texttt{SQLiteConnected} resource, which signifies that a
+successful connection has been made to the database. This resource contains a
+pointer to the database structure which is used in further computations.
 
@@
 \noindent
-We secondly define the \texttt{SQLitePSSuccess}, \texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail} resources to represent the process of creating and binding variables to a prepared statement.
+We secondly define resource types to capture success and failure states
+of binding a prepared statement:
+\texttt{SQLitePSSuccess}, 
+\texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail}. The types are
+declared as follows (we leave the definitions abstract):
 
@@
 data BindStep = Binding | Bound
-
 data SQLitePSSuccess : BindStep -> Type where
-  SQLitePS       : ConnectionPtr -> 
-                   StmtPtr -> SQLitePSSuccess a
-  SQLiteBindFail : ConnectionPtr -> 
-                   StmtPtr -> 
-                   BindError -> SQLitePSSuccess a
-
 data SQLitePSFail : Type where
-  PSFail : ConnectionPtr -> SQLitePSFail
-
 data SQLiteFinishBindFail : Type where
-  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
-                 SQLiteFinishBindFail
 
 \end{SaveVerbatim}
+
+%  SQLitePS       : ConnectionPtr -> 
+%                   StmtPtr -> SQLitePSSuccess a
+%  SQLiteBindFail : ConnectionPtr -> 
+%                   StmtPtr -> 
+%                   BindError -> SQLitePSSuccess a
+%data SQLitePSFail : Type where
+%  PSFail : ConnectionPtr -> SQLitePSFail
+%
+%data SQLiteFinishBindFail : Type where
+%  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
+%                 SQLiteFinishBindFail
+
 \useverb{sqliteconnected}
@@
 
-To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor.
+%To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor.
 Should any binding operations be called whilst in this state, the operations will not be passed to the underlying library.
@@
 \noindent
-Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.
+Column access functions may only be called when there is a valid row in the
+buffer, as signified by the input resources:
 
@@
 \noindent
-In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct.
-The \texttt{nextRow} function will either fetch a row into the buffer, or indicate that a row could not be loaded, as shown by its type:
+In order to provide the necessary static guarantee that there is a valid row to
+process in the buffer, we make use of the \texttt{if\_valid} construct. The
+\texttt{nextRow} function will either fetch a row into the buffer, or indicate
+that a row could not be loaded, as shown by its type:
 
+\noindent
 \begin{SaveVerbatim}{nextrow}
+
 nextRow : EffM IO [SQLITE (SQLiteExecuting ValidRow)] 
          [SQLITE (Either (SQLiteExecuting InvalidRow)
-                                  (SQLiteExecuting ValidRow))] 
-                  StepResult
+                 (SQLiteExecuting ValidRow))] StepResult
+
 \end{SaveVerbatim}
@@
 
-Programs making use of the DSL should look familiar to developers even without
-a background in functional programming. To demonstrate the functionality of the
+%Programs making use of the DSL should look familiar to developers even without
+%a background in functional programming. 
+To demonstrate the 
 library, we return to the previous example of selecting the names and addresses
-of all staff working in a given department. Due to the fact that the
-\texttt{Effects} library overloads the bind operator, we may make use of
-\texttt{do}-notation, facilitating the usage of an imperative style.
+of all staff working in a given department. 
+%Since the
+%\texttt{Effects} library overloads the bind operator, we may make use of
+%\texttt{do}-notation, facilitating the usage of an imperative style.
 We define a function \texttt{textSel} of type:
@@
 
-String ->
-Eff IO [SQLITE ()] (Either QueryError (List (String, String)))
+String -> Eff IO [SQLITE ()] 
+           (Either QueryError (List (String, String)))
 
@@
 \noindent
-This means that the program will be run in \texttt{IO}, and must
-start and end with no active resources. The return type indicates that either a
-list of \texttt{(String, String)} pairs, representing names and addresses in the
-database, or an error will be returned.
+The program will be run in \texttt{IO}, and starts and finishes with no active
+resources.  It reutnrs either list of \texttt{(String, String)} pairs,
+representing names and addresses in the database, or an error.
 %This example is shown in Figure
@@
 the program transitions to the \texttt{ConnectionOpened} state. The
-\texttt{openDB} function returns a Boolean value indicating whether or not the
-operation is successful. If not, then the \texttt{connFail} function is called
-to generate an appropriate error and dispose of the resources, as shown in
-~\ref{fig:sqlitestates}.
+\texttt{openDB} function returns a \texttt{Bool} indicating whether the
+operation is successful. If not, the \texttt{connFail} function
+generates an appropriate error and disposes of the resources, as shown in
+Figure~\ref{fig:sqlitestates}.
 
@@
 from the database.
-
+%
 \texttt{CollectResults} operates on the row currently held in the buffer.
-Firstly, the function checks, through the use of the \texttt{if\_valid} construct, that there is a valid row in the buffer, and if so uses the \texttt{getColumnText} and \texttt{getColumnInt} functions
-to retrieve the data from the database. This function is then
-called recursively until there are no more rows to process.
+Firstly, the function checks
+that there is a valid row in the buffer, and if so uses the
+\texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data
+from the database. This function is then called recursively until there are no
+more rows to process.
 
 \begin{SaveVerbatim}{collectresty}
+
 collectResults :
@@
 
-In order to decrease unnecessary boilerplate code in user applications, we
-provide functions which abstract out unnecessary parts of this process. In
-order to do this, we define the algebraic data type \texttt{DBVal}, which is a
-tagged union over simple primitive types:
+\noindent
+Using this, we can build a function to execute a full query:
+
+\begin{SaveVerbatim}{execsel}
+
+executeSelect : 
+  String -> String -> List (Int, DBVal) -> 
+  (Eff IO [SQLITE (SQLiteExecuting ValidRow)] 
+               (List DBVal)) -> 
+  Eff IO [SQLITE ()] (Either QueryError ResultSet)
+
+\end{SaveVerbatim}
+\useverb{execsel}
+
+\noindent
+This returns either an error, or a set of results, defined
+as follows:
 
 \begin{SaveVerbatim}{dbvals}
+
+ResultSet : Type
+ResultSet = List (List DBVal)
 
@@
 
-\noindent
-We also make use of the \texttt{ResultSet} type, which is a list of returned
-database rows.
-
-\begin{SaveVerbatim}{resultset}
-
-ResultSet : Type
-ResultSet = List (List DBVal)
-
-\end{SaveVerbatim}
-\useverb{resultset}
-
-\noindent
-Using these, we may implement a function, \texttt{executeSelect}, which, given
-a query, a list of variables to bind and their associated indices within the
-query, and a function which is used to extract information out of each returned
-database row, returns either a \texttt{ResultSet} or an error.
-
-\begin{SaveVerbatim}{execsel}
-
-executeSelect : 
-  String -> String -> List (Int, DBVal) -> 
-  (Eff IO [SQLITE (SQLiteExecuting ValidRow)] (List DBVal)) -> 
-  Eff IO [SQLITE ()] (Either QueryError ResultSet)
-
-\end{SaveVerbatim}
-\useverb{execsel}
-
 % =================================================
 \subsection{A Simple Session Handler}
-More complex web applications require some persistent state across separate
-requests. This is often done through an abstraction of a \textit{session},
-wherein a cookie is set on the remote host containing a unique session ID,
-which is in turn used to retrieve data. In this section, we describe the
-implementation of a simple session handler, and the resource usage protocol
+Larger web applications require persistent state across separate
+requests. This can be achieved using a \textit{session},
+in which a cookie is set on the remote host containing a unique session ID,
+which is used to retrieve data. In this section, we describe
+a simple session handler, and the resource protocol
 involved.
 
-A major strength of the \texttt{Effects} library is that it allows for simple
-composition of individual, fine-grained effects. By combining the individual
-CGI and SQLite components, we may construct a simple session handler to provide
-a notion of state across separate web requests.
+The \texttt{Effects} library allows for composition of individual, fine-grained
+effects. By combining the \texttt{CGI} and \texttt{SQLite} components, we can
+construct a simple session handler to provide a notion of state across separate
+web requests. 
+We implement this with a SQLite database containing tables for storing
+session keys and expiry dates, along with the data associated with
+the session.
 
-We implement this by having a SQLite database, containing two tables:
-\texttt{session}, which stores session keys and their associated expiry dates,
-and \texttt{sessiondata}, which contains the data associated with each session.
-A datum associated with the session is described as a tagged union containing
-one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int},
-which is serialised alongside a type tag for storage in the database.
+%We implement this with a SQLite database, containing two tables:
+%\texttt{session}, which stores session keys and their associated expiry dates,
+%and \texttt{sessiondata}, which contains the data associated with each session.
+%A datum associated with the session is described as a tagged union containing
+%one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int},
+%which is serialised alongside a type tag for storage in the database.
 %TODO: Fix up the labels
@@
 with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
-algebraic effect. In this application, there exist two states:
+algebraic effect. In this application, there are two states:
 In \texttt{SessionClosed}, the user may load or create a
@@
 representation of the session in memory, serialise the session and write it to
-the database, or delete the session and invalidate the user's session key. The
-introduction of these two states ensures that changes are explicitly either
+the database, or delete the session and invalidate the user's session key. 
+These two states ensure that changes are explicitly either
 written or discarded, eliminating the possibility of a developer updating the
@@
 
-The \texttt{SessionRes} data type is parameterised over the current state,
-which determines which operations may be performed, and has two constructors:
-\texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as
-creating a new session fails, no further side-effecting calls will be made, in
-turn preserving integrity.
+%The \texttt{SessionRes} data type is parameterised over the current state,
+%which determines which operations may be performed, and has two constructors:
+%\texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as
+%creating a new session fails, no further side-effecting calls will be made, in
+%turn preserving integrity.
 

### 2013-11-18 paper/conclusion.tex
(Hopefully) final revision of paper before IFL2013 submission.

@@
 guarantee that library calls are made in the correct order, and calls to
-retrieve rows of data are made only when more data is available. Additionally,
-by recording any failures in the resource associated with the database effect,
-we may ensure that any failures at certain stages in the process may be
-correctly dealt with, and not propagate to further library calls---if a
-prepared statement fails to compile, for example, the framework guarantees that
-it will not be executed, without causing a fatal runtime error or memory access
-violation.
-%
+retrieve rows of data are made only when more data is available.
+Additionally, by encoding desired invariants within operation types, we may gain static guarantees about adherence to resource usage protocols and failure handling.
 Enforcing resource usage protocols also guards against common programmer

### 2013-11-18 paper/protocols.tex
(Hopefully) final revision of paper before IFL2013 submission.

@@
 functions to output data to the web page and append data to the response
-headers. It is important to note, that at this stage nothing is written to the
+headers. It is important to note that at this stage nothing is written to the
 page, with the \texttt{output} and \texttt{addHeader} functions instead
@@
 The program will be run in \texttt{IO}, and starts and finishes with no active
-resources.  It reutnrs either list of \texttt{(String, String)} pairs,
+resources.  It returns either a list of \texttt{(String, String)} pairs,
 representing names and addresses in the database, or an error.
@@
 the program transitions to the \texttt{ConnectionOpened} state. The
-\texttt{openDB} function returns a \texttt{Bool} indicating whether the
-operation is successful. If not, the \texttt{connFail} function
-generates an appropriate error and disposes of the resources, as shown in
-Figure~\ref{fig:sqlitestates}.
+\texttt{openDB} function returns either an error code, if the connection fails, or a unit type should the connection succeed, as shown in Figure \ref{fig:sqlitestates}.
 

### 2014-01-29 paper/effects.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

@@
 list of side effects \texttt{es} that the program is permitted to use, and the
-programs return type \texttt{a}. The name \texttt{m} for the computation context is
+program's return type \texttt{a}. The name \texttt{m} for the computation context is
 suggestive of a monad, but there is no requirement for it to be so.
@@
 
-Effects such as state and exception are described as algebraic data types,
+Effects such as `State' and `Exception' are described as algebraic data types,
 and run by giving \emph{handlers} for specific computation contexts. 
@@
 reading a line is only possible if the
-resoource is a file open for reading, etc.
+resource is a file open for reading, etc.
 The handler for this effect for an \texttt{IO} computation context will
@@
 a resource of type \texttt{Either a b} indicates failure (\texttt{Left a})
-or success (\texttt{Right b}) and branches and updates the resource
-accordingly.
-Therefore, attempting to write to the file, failing to check for
+or success (\texttt{Right b}). In this case, the initial resource in the valid
+branch is \texttt{OpenFile Read}, indicating that the file has been successfully 
+opened for reading, whereas in the failing branch, the resource remains as \texttt{()}.
+
+Attempting to write to the file, failing to check for
 success, or failing to open or close the
@@
 throughout this paper: describe a resource usage protocol in terms of
-state transitions; implement an effect which captures that protocol; implement
+state transitions; implement an effect which captures that protocol; and implement
 programs which, by using this effect, implicitly carry a proof that the resource

### 2014-01-29 paper/introduction.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

@@
 such as PHP, Ruby or Python, which facilitate rapid development
-\cite{w3techs:webpls}. However, such languages
-do not provide the same static guarantees about runtime behaviour afforded by
+\cite{w3techs:webpls}. Such languages do not, however, provide the same static 
+guarantees about runtime behaviour afforded by
 programs with more expressive, static type systems, instead relying on
@@
 \emph{algebraic effects}~\cite{brady:effects}, we
-present a dependently-typed web framework, which allows the construction of
+present a dependently-typed web framework allowing the construction of
 programs with additional guarantees about correctness and security, whilst

### 2014-01-29 paper/main.tex
(Hopefully) final revision of paper before IFL2013 submission.
Fix error introduced in last commit

--- 2013-11-11 paper/main.tex
@@
 Dependently-typed languages allow precise types to be used during development,
-facilitating easier reasoning about the operation of programs. However,
+facilitating static reasoning about program behaviour. However,
 stronger types bring a disadvantage that it becomes increasingly difficult to
-write programs that are accepted by a type checker and additional proofs may
-have to be specified by a programmer.
+write programs that are accepted by a type checker, meaning additional proofs may
+have to be specified by a developer.
 
-Embedded domain-specific languages (EDSLs) address this problem by introducing
-a layer of abstraction over more specific underlying types, allowing
-domain-specific code to be written in a high-level languages which uses
-dependent types to enforce certain invariants without additional proof
-obligations.
+Embedded domain-specific languages (EDSLs) can help address this problem by
+introducing a layer of abstraction over more precise underlying types,
+allowing domain-specific code to be written in a high-level language which uses
+dependent types to enforce invariants without imposing additional proof
+obligations on an application developer.
 
@@
 
-%\keywords
+\keywords
+Dependent types, web applications, security, verification
 %keyword1, keyword2
@@
 This work has been supported by the Scottish Informatics and Computer Science Alliance (SICSA) and the EPSRC.
-The authors would like to thank contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
+We would like to thank the contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
+We are very grateful to Peter Thiemann and the anonymous reviewers for their insightful and constructive comments.
 % SICSA / EPSRC (grant number?

### 2014-01-29 paper/protocols.tex
Fix error introduced in last commit

@@
 that there is a valid row in the buffer, and if so uses the
-\texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data
+\texttt{getColumnText} function to retrieve the data
 from the database. This function is then called recursively until there are no
@@
           [SQLITE (SQLiteExecuting InvalidRow)] 
-          (List (String, Int))
+          (List (String, String))
 \end{SaveVerbatim}
@@
   if_valid then do name <- getColumnText 1
-                   age <- getColumnInt 2
+                   addr <- getColumnText 2
                    step_res <- nextRow
                    xs <- collectResults
-                   return $ (name, age) :: xs
+                   return $ (name, addr) :: xs
   else return []

### 2014-01-30 paper/conclusion.tex
Paper revisions

@@
 callbacks may be associated with submit buttons, with nameless representations
-for form inputs. Due to the unavailability of full dependent types in Haskell,
-however, this implementation does not statically verify the suitability of the
-callback function for the form inputs. Both implementations of the CGI library,
+for form inputs. Both implementations of the CGI library,
 being built upon monads, mean that the use of additional effects such as
@@
 edit-driven. This means that upon a component being edited, a computation
-occurs, given the state of the current form. This is saved for
-future computations. Should a user enter invalid data, for example by entering
+occurs, given the state of the current form. Should a user enter invalid data, for example by entering
 text in a field designated for integers, the change will be reverted. This is
@@
 high-level combinators which allow the implementation of concepts such as
-recursion, sequence and choice in a scalable fashion. These high-level
-abstractions are elegant, but the style and syntax differ substantially from a
-traditional web application. Our approach takes the concept of type-safe input
-handling and uses it in a more traditional fashion, whilst retaining the
-type-retention guarantees afforded by \idata{} elements.
+recursion, sequence and choice in a scalable fashion.
 
@@
 
-%Ur\/Web \cite{urweb} is a library for the Ur language which makes use of
-%dependent types to enable the generation of provably-correct and secure queries
-%and DOM code, without requiring overly complex types or proofs. Through the use
-%of the Ur\/Web DSL, developers may generate well-typed code from records. By
-%statically verifying the \textit{code generator}, it is therefore possible to
-%assert that the generated code is secure. Such ideas are extremely interesting
-%for the implementation of object-relational mapping, which we hope to implement
-%in the near future.
+Formlets \cite{cooper:formlets} are a well-known functional abstraction over web forms, making use of McBride and Paterson's applicative functors \cite{mcbride:applicative} to provide an extensible and powerful method of building and handling web forms.
+Our approach differs in that we check that handler functions conform to the types of the elements in the forms by parameterising the resource associated with the form construction effect, as opposed to using a preprocessor to rewrite the form elements in applicative notation.
+Our framework does not yet support composition of sections of forms, as is the case with formlets, but we foresee no problems with extending our DSL to add this functionality.
 
+Java is often used as a language to write enterprise-level web applications.
+Frameworks such as the Java Persistence API (JPA) \cite{jpa} use Java annotations to translate data models into the appropriate database schemata, but such code may soon become unwieldy due to the large amount of redundant boilerplate code such as accessor and mutator functions.
 WebDSL \cite{webdsl} is a domain-specific language written primarily to
 introduce new abstractions which aim to reduce the amount of boilerplate code
-that must be written and maintained by developeres. WebDSL is built on top of
-Java, which often includes a large amount of redundant code such as accessor
-and mutator functions within entity classes. The Java Persistence API (JPA)
-\cite{jpa} provides an object-relational mapping through the use of Java 5
-annotations, which may then be used to construct database tables. These
-annotations soon become complex, however, and coupled with redundant
-boilerplate code, data model declarations may soon become unwieldy. Through the
-use of WebDSL, these data model declarations can be much more compactly
-declared, and elaborated into Java code by parsing the data-modelling DSL into
-an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL
-also applies similar concepts to implement a \textit{template system} for such
-objects, which allows the data to be used in code generation. We look to
+that must be written and maintained by developers. 
+The DSL is parsed into an abstract syntax tree, modified using rewrite rules, and elaborated back into Java code.
+WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities.
+We look to
 implement many of these ideas, but as effects within the IdrisWeb
@@
 There are many other applications which make use of specific resource usage
-protocols, for example popular libraries such as
-\textit{libgcrypt}\footnote{\texttt{http://directory.fsf.org/wiki/Libgcrypt}}.
+protocols, for example popular cryptography libraries such as
+\textit{Sodium}\footnote{\texttt{https://github.com/jedisct1/libsodium}}.
 Applying a similar approach would allow for sensitive programs requiring
@@
 \cite{christiansen:dtp}, exploiting the fact that types can be
-calculated by functions to avoid generating extra code in the type provider
-step. Dependent type providers additionally have stronger safety guarantees as
-they may not generate unchecked code, but at the same time this is matched by a
-decrease in expressiveness. Nonetheless, such techniques provide a promising
-mechanism to verify the semantic soundness of programs and we look to
-investigate their integration in further work.
+calculated by functions to avoid unsafely generating extra code in the type provider
+step. We believe that this technique would be an interesting avenue of exploration to provide further static guarantees about the form of database queries.
 

### 2014-01-30 paper/protocols.tex
Paper revisions

@@
 
-%A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell.
-Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program.
-If this were to happen, then the program would fail due to the fact that headers had not been written to the client.
-
 By modelling CGI as a resource-dependent algebraic effect, we may enforce a
 resource usage protocol which
-%even though the program may be running in an IO execution context, 
 prevents arbitrary IO from being performed and therefore
@@
 % =================================================
+
+Here, \texttt{output} is a function which appends some output to the CGI output buffer, which is later written to the page.
 

### 2014-02-03 paper/conclusion.tex
More paper revisions...

@@
 
-Type providers, as originally implemented in F\# \cite{msr:tp}, are an
-interesting method by which external data sources may be used to import
-external information, such that it may be used during compilation. 
+Type providers, as originally implemented in F\# \cite{msr:tp}, allow external data sources to be used to import
+external information to be used during compilation. 
 In this way, it becomes possible to use the extra type
@@
 conform to a given database schema, for example, then the program will not
-type-check.  This has been implemented for 
+type-check.  Type providers have been implemented for 
 \idris{} 

### 2014-02-03 paper/forms.tex
(Hopefully) final revision of paper before IFL2013 submission.
Paper revisions
More paper revisions...

@@
-\section{Type-aware form handling}
+\section{Dependently-typed Form Handling}
 
@@
   else do
-    addForm "nameform" "helloform" showHelloForm
+    addForm showHelloForm
     return ()
@@
 \ref{fig:formeffect}.
-The \texttt{using} notation here indicates that within
-the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
-the given type.
+
+The \texttt{using} notation indicates that where \texttt{G} and \texttt{E} occur within the block, they correspond to \emph{implicit arguments} within the type of the form construction resource. 
+\texttt{G} is of type \texttt{List FormTy}, indicating the types of the elements within the form, and \texttt{E} is of type \texttt{List WebEffect}.
+This allows us to record the types of form elements, which we may later use to ensure that the type of the handler function corresponds to the types of the elements within the form.
+%The \texttt{using} notation here indicates that within
+%the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
+%the given resource type.
 
 The general process of form construction is illustrated by the \texttt{AddTextBox}
-and \texttt{Submit} operations of the \texttt{Form} effect:
-
-\begin{SaveVerbatim}{formeffsmall}
-
-data Form : Effect where
-  AddTextBox : (label : String) -> (fty : FormTy) -> 
-               Maybe (interpFormTy fty) -> 
-               Form (FormRes G E) 
-                    (FormRes (fty :: G) E) () 
-     ...
-  Submit : (mkHandlerFn ((reverse G), E)) -> String -> 
-           Form (FormRes G E) (FormRes [] []) String
-
-\end{SaveVerbatim}
-\useverb{formeffsmall}
-
-\noindent
-These use the resource associated with the effect, \texttt{FormRes}, to
-construct the form. Adding a field such as a text box adds a new type,
+and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure ~\ref{fig:formeffect}.
+These use the resource associated with the effect, \texttt{FormRes}, to construct the form.
+Adding a field such as a text box adds a new type,
 \texttt{fty} to the list of field types, carried in the resource. When the form
@@
                         (FormRes [])] String
-
 \end{SaveVerbatim}
@@
 default.
-
-Handlers may only be associated with a form if they have argument types
-corresponding to the types associated with the form elements. Additionally, we
-wish to name the function in order for it to be serialised, whilst requiring a
-proof that the specified name is associated with the function. If this were not
-required, it would be possible to use a function satisfying the type
-requirement, without guaranteeing that the serialised data corresponded to
-that function.
 
@@
 \begin{SaveVerbatim}{interpwebeff}
-
 webEffect : WebEffect -> EFFECT
-
 \end{SaveVerbatim}
@@
 
-Whilst it is not possible to serialise arbitrary effects due to the associated
+While it is not possible to serialise arbitrary effects due to the associated
 difficulties with serialising initial resource environments, we allow for three
@@
 permits introduction of additional effects within an effectful computation.
-%
+
+Handlers may only be associated with a form if they have argument types
+corresponding to the types associated with the form elements. Additionally, we
+wish to name the function in order for it to be serialised, whilst requiring a
+proof that the specified name is associated with the function. If this were not
+required, it would be possible to use a function satisfying the type
+requirement, without guaranteeing that the serialised data corresponded to
+that function.
+
 We may specify a handler function of type \texttt{FormHandler}:
@@
 \noindent
-In order to associate a handler with a form, we may call the \texttt{addSubmit}
+In order to associate a handler with a form, we call the \texttt{addSubmit}
 function:
@@
 This function takes a handler function and a list of available handlers, along with
-an automatically constructed proof (using the a \texttt{default} argument) that the
+an automatically constructed proof (using the \texttt{default} argument) that the
 handler is available.
@@
 to the list of \texttt{FormTy}s as opposed to appending them, we must reverse
-the list.
+the list of form elements \texttt{G}.
 
@@
 element of the pair is parameterised over the value of the first element. It is
-an existential binding;
+an existential binding:
 the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
@@
 \noindent
-\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the
+\texttt{FnElem} is parameterised over \texttt{G} and \texttt{E}, the types of the
 form elements and the effects used by the handler function. \texttt{FnHere}
@@
 the list. 
-%TODO: this should probably be rewritten
 We then use an automatic proof search to

### 2014-02-03 paper/introduction.tex
More paper revisions...

@@
 and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
-scripting \cite{owasp:xss}. Security breaches using such exploits are
+scripting \cite{owasp:xss}. Security breaches are
 far-reaching, and high profile cases involve large corporations such as Sony,
@@
 %%------- EXAMPLE
+\noindent
 Let us consider a simple database access routine, written in
@@
 A developer may, for example, accidentally close a statement whilst still
-retrieving rows, which would cause a runtime error. Similarly, a programmer may
+retrieving rows, which would cause a runtime error. Similarly, a developer may
 omit closing the statement or connection, which can lead to
@@
 Unfortunately, automatic verification by a compiler can be difficult or
-often impossible, requiring additional proofs to be given by the programmer.
+often impossible, requiring additional proofs to be given by the developer.
 

### 2014-02-03 paper/messageboard.tex
(Hopefully) final revision of paper before IFL2013 submission.
More paper revisions...

@@
 \begin{SaveVerbatim}{handlereq}
-
-handleRequest : CGIProg 
-     [SESSION (SessionRes SessionUninitialised), 
+ handleRequest : 
+   CGIProg [SESSION (SessionRes SessionUninitialised), 
             SQLITE ()] ()
@@
    if handler_set then do
-    lift' (handleForm handlers)
-    Effects.return ()
+     handleForm handlers
+     return ()
    else do
-    action <- lift' (queryGetVar "action")
-    thread_id <- lift' (queryGetVar "thread_id")
+     action <- queryGetVar "action"
+     thread_id <- queryGetVar "thread_id"
      handleNonFormRequest action (map strToInt thread_id)
@@
 
-\noindent
 Once we have loaded the session data from the database, we then check whether
@@
 performed by \texttt{threadInsert}, shown in Figure \ref{fig:threadins}.
+
 This uses a library function \texttt{executeInsert}, which abstracts over the
 low-level resource usage protocol, enabling for provably-correct database
-access without the excess boilerplate code. In addition, \texttt{executeInsert}
+access without the additional DSL code. In addition, \texttt{executeInsert}
 returns the unique row ID of the last item which was inserted, which may be
@@
 \useverb{traversethreads}
-\caption{Thread Insertion}
+\caption{Listing Threads}
 \label{fig:traverse}
@@
 handleLoginForm (Just name) (Just pwd) = do
-  auth_res <- lift' (authUser name pwd)
+  auth_res <- (authUser name pwd)
   case auth_res of
@@
       if set_sess_res then do
-        lift' (output $ "Welcome, " ++ name)
+        output $ "Welcome, " ++ name
         return ()
       else do
-        lift' (output "Could not set session")
+         output "Could not set session"
          return ()
     Right Nothing => do
-      lift' (output "Invalid username or password")
+      output "Invalid username or password"
       return ()
     Left err => do
-      lift' (output $ "Error: " ++ err)
+      output $ "Error: " ++ err
       return ()
@@
 \useverb{handlelogin}
-\caption{Thread Insertion}
+\caption{Authentication}
 \label{fig:handlelogin}
@@
 can be used to increase confidence in an (albeit simplified) real-world
-application, without requiring developers to supply proofs or indeed work
-explicitly with dependent types.
+application, without requiring developers to supply proofs.
 

### 2014-02-03 paper/main.tex
Paper revisions

@@
 
-\documentclass[preprint]{sigplanconf}
+\documentclass[]{sigplanconf}
 
@@
 
+\authorversion
+
 \titlebanner{}        % These are ignored unless
@@
 facilitating static reasoning about program behaviour. However,
-stronger types bring a disadvantage that it becomes increasingly difficult to
-write programs that are accepted by a type checker, meaning additional proofs may
-have to be specified by a developer.
+with the use of more specific types comes the disadvantage that it becomes increasingly difficult to
+write programs that are accepted by a type checker, meaning additional proofs may have to be specified manually.
 
@@
 introducing a layer of abstraction over more precise underlying types,
-allowing domain-specific code to be written in a high-level language which uses
-dependent types to enforce invariants without imposing additional proof
+allowing domain-specific code to be written in a verified high-level language without imposing additional proof
 obligations on an application developer.
@@
 In this paper, we apply this technique to web programming.  Using the
-dependently typed programming language \Idris{}, we introduce an EDSL to
-facilitate the creation and handling of \emph{statically checked} web forms,
-reducing the scope for programmer error and attacks such as SQL injection. We
-also show how to enforce resource usage protocols associated with common web
-operations such as CGI, database access and session handling.
+dependently typed programming language \Idris{}, we show how to use EDSLs to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling.
+We also introduce an EDSL which uses dependent types to facilitate the creation and handling of web forms, reducing the scope for programmer error and possible security implications.
 
@@
 
-%\category{CR-number}{subcategory}{third-level}
+\category{D.3.2}{Programming Languages}{Language Classifications---Applicative (functional) Languages}
 
@@
 \keywords
-Dependent types, web applications, security, verification
+Dependent Types, Web Applications, Verification
 %keyword1, keyword2
@@
 We would like to thank the contributors to the \idris{} language, especially the authors of the original \texttt{Network.Cgi} and \texttt{SQLite} libraries.
-We are very grateful to Peter Thiemann and the anonymous reviewers for their insightful and constructive comments.
+We are very grateful to Peter Thiemann and the anonymous reviewers for their insightful and constructive comments and suggestions.
 % SICSA / EPSRC (grant number?

### 2014-02-03 paper/protocols.tex

@@
 
-CGI is used to invoke an application on a web server, making use of environment
+The Common Gateway Interface (CGI) is used to invoke an application on a web server, making use of environment
 variables to convey information gained from an HTTP request and using standard
@@
 correctly written to the browser prior to any other output; failure to do so
-will result in an internal server error being shown.
+will result in an HTTP Error 500.
 
@@
 This is then passed, along with initial values for other effects that the user
-may wish to use, to the runAction function, which invokes the RunAction
+may wish to use, to the \texttt{runAction} function, which invokes the \texttt{RunAction}
 operation and executes the user-specified action.
@@
 In order to provide the necessary static guarantee that there is a valid row to
-process in the buffer, we make use of the \texttt{if\_valid} construct. The
+process in the buffer, we make use of the \texttt{if\_valid} construct, as described in Section ~\ref{rp-effects}. The
 \texttt{nextRow} function will either fetch a row into the buffer, or indicate

### 2014-02-03 paper/effects.tex
More paper revisions...
Tinkering and proof reading paper

@@
 in a monadic style, with \texttt{do}-notation, with their type stating which
-specific effects are available.
-Effectful
-programs are described using the following data type,
-in the simplest case:
+specific effects are available:
+%Effectful
+%programs are described using the following data type,
+%in the simplest case:
 
@@
 \subsection{Resource Protocols as Effects}
-
+\label{rp-effects}
 \begin{SaveVerbatim}{fileeff}

### 2014-02-03 paper/forms.tex
Tinkering and proof reading paper

@@
 
-In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
-of the form-handling system: namely, the effect which allows the creation of
+In Sections~\ref{formcons} and~\ref{formhandling}, we examine implementation
+of the form-handling system: the effect which allows the creation of
 forms, and the handling code which deserialises the data and passes it to the
@@
 \label{formcons}
-Each form element is specified to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function.
-In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
+Each form element holds a value of a specific type, which, assuming that the
+correct type of data is given by the user, is passed directly to the
+handler function. In order to encapsulate this, we firstly define the allowed
+data types as part of an algebraic data type, \texttt{FormTy}.
 
@@
 \noindent
-Since types in \idris{} are first-class, we may use this to convert
-between abstract and concrete representations of allowed form types:
+%Since types in \idris{} are first-class, 
+We convert
+from an abstract to a concrete representation of an allowed form type as
+follows:
 
@@
 The general process of form construction is illustrated by the \texttt{AddTextBox}
-and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure ~\ref{fig:formeffect}.
+and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure~\ref{fig:formeffect}.
 These use the resource associated with the effect, \texttt{FormRes}, to construct the form.
@@
                     [FORM (FormRes G effs)] ()
-useEffects effs = (UseEffects effs)
+useEffects effs = UseEffects effs
 
@@
 \noindent
-We may then define handlers in a more intuitive fashion, without being
+We may then define handlers in a more readable fashion, without being
 concerned with the implementation details. This allows us to write a handler
@@
 
+\noindent
 To enforce the requirement that a supplied handler function must be in the

### 2014-02-03 paper/introduction.tex
Tinkering and proof reading paper

@@
 far-reaching, and high profile cases involve large corporations such as Sony,
-who suffered a well-publicised and extremely costly SQL Injection breach in
+who suffered a well-publicised and extremely costly SQL injection breach in
 2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
@@
 %%------- EXAMPLE
-\noindent
 Let us consider a simple database access routine, written in

### 2014-02-03 paper/protocols.tex
Tinkering and proof reading paper

@@
 resource usage protocol which
-prevents arbitrary IO from being performed and therefore
+prevents arbitrary I/O from being performed and therefore
 ensures that the headers are written correctly. We
@@
 After the user code has finished execution, control returns to the library
-code. At this point, the state transitions to TaskCompleted, and the headers
+code. At this point, the state transitions to \texttt{TaskCompleted}, 
+and the headers
 are written.  Finally, the headers and content are written which completes the
@@
 
+\noindent
 Here, \texttt{output} is a function which appends some output to the CGI output buffer, which is later written to the page.
@@
 {-                           { Input resource type }    { Output resource type }             { Value }           -}
+
 data Sqlite : Effect where
@@
 
-Through the use of dependent types, we may encode these invariants statically, using the \texttt{SQLiteExecuting} resource.
-This is parameterised by the \texttt{ExecutionResult} data type, which indicates whether there is currently a valid row in the buffer.
+We encode these invariants using the \texttt{SQLiteExecuting}
+resource. This is parameterised by \texttt{ExecutionResult},
+which indicates whether there is currently a valid row in the buffer:
 
@@
 Column access functions may only be called when there is a valid row in the
-buffer, as signified by the input resources:
+buffer, as specified by the input resources:
 
@@
 In order to provide the necessary static guarantee that there is a valid row to
-process in the buffer, we make use of the \texttt{if\_valid} construct, as described in Section ~\ref{rp-effects}. The
+process in the buffer, we make use of \texttt{if\_valid}, as described in Section ~\ref{rp-effects}. The
 \texttt{nextRow} function will either fetch a row into the buffer, or indicate
@@
 \noindent
-The \texttt{if\_valid} construct provides failure checking functionality, allowing different operations to be performed depending on whether or not a row was successfully fetched for processing.
+The \texttt{if\_valid} construct checks for failure,
+allowing different operations to be executed depending on whether or not a row
+was successfully fetched for processing.
 %If a failure happens at any point during the computation, the resource is
@@
 
+\begin{figure}[htpb!]
+\centering
+\scalebox{0.8}{
+\begin{tikzpicture}[>=latex]
+  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
+
+  \tikzstyle{stateEdgePortion} = [black,thick];
+  \tikzstyle{stateEdge} = [stateEdgePortion,->];
+  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
+
+  \node[initial,state] (A)              {SessionClosed};
+  \node[state]         (B) [below of=A] {SessionOpen};
+
+  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
+        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
+            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
+\end{tikzpicture}
+}
+\caption{Session Handler Resource Usage Protocol}
+\label{fig:sessionstates}
+\end{figure}
+
 \begin{SaveVerbatim}{sessioneff}
@@
 \end{figure*}
-
-\begin{figure}[htpb!]
-\centering
-\scalebox{0.8}{
-\begin{tikzpicture}[>=latex]
-  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]
-
-  \tikzstyle{stateEdgePortion} = [black,thick];
-  \tikzstyle{stateEdge} = [stateEdgePortion,->];
-  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];
-
-  \node[initial,state] (A)              {SessionClosed};
-  \node[state]         (B) [below of=A] {SessionOpen};
-
-  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
-        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
-            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
-\end{tikzpicture}
-}
-\caption{Session Handler Resource Usage Protocol}
-\label{fig:sessionstates}
-\end{figure}
 

### 2014-02-03 paper/conclusion.tex
Few more revisions

@@
 
+Ensuring conformance to resource usage protocols has been attempted using the notion of \emph{typestate} in object-oriented languages \cite{deline:typestates}.
+Our approach differs in that we make no changes to the type system in order to implement this functionality: we check resource usage protocol conformance purely within \idris{}.
+Additionally, our handler-based approach facilitates greater control over side effects.
+
 Java is often used as a language to write enterprise-level web applications.
@@
 The DSL is parsed into an abstract syntax tree, modified using rewrite rules, and elaborated back into Java code. 
-WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities.
-We look to
-implement many of these ideas, but as effects within the IdrisWeb
-framework, as with the form construction effect.
+%WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities. We look to
+%implement many of these ideas, but as effects within the IdrisWeb
+%framework, as with the form construction effect.
 
@@
 immediately eliminates the possibilities of SQL injection attacks on non-string
-types. Since failures are handled transparently, no runtime errors are output
+types. Since failures are handled transparently, no runtime errors are written
 to the browser, meaning that attackers may not use such information to aid
@@
 We have shown that embedded domain-specific languages using dependent types and
-algebraic effects can be used to increase confidence in web applications by
-providing additional static guarantees about runtime behaviour, but much more
+algebraic effects can be used to increase confidence in web applications, but much more
 can be done using the same approach.
@@
 include a template system, allowing for web pages to be automatically generated
-from data, and an object-relational mapping system which will allow users to
-manipulate records which can then be automatically written to the database,
-instead of having to update tables manually via SQL queries.
+from data, and an object-relational mapping system. 
+It would also be interesting to 
+explore the use of continuation-passing approaches \cite{queinnec:ioc} to enhance the usability of the framework.
 

### 2014-02-03 paper/protocols.tex
Few more revisions

@@
 %\texttt{do}-notation, facilitating the usage of an imperative style.
-We define a function \texttt{textSel} of type:
+We define a function \texttt{textSelect} of type:
 
@@
         closeDB
-        Effects.pure $ Right results
+        return $ Right results
       else do
         cleanupBindFail
-        Effects.pure $ Left (getBindError bind_res)
+        return $ Left (getBindError bind_res)
     else do
       cleanupPSFail
-      Left $ getQueryError sql_prep_res
+      return $ Left (getQueryError sql_prep_res)
   else do 
-    Effects.pure $ Left (getQueryError db_res)
+    return $ Left (getQueryError db_res)
 \end{SaveVerbatim}

### 2014-02-03 paper/conclusion.tex
Tiny changes

@@
 
-Ensuring conformance to resource usage protocols has been attempted using the notion of \emph{typestate} in object-oriented languages \cite{deline:typestates}.
-Our approach differs in that we make no changes to the type system in order to implement this functionality: we check resource usage protocol conformance purely within \idris{}.
-Additionally, our handler-based approach facilitates greater control over side effects.
+Ensuring conformance to resource usage protocols has been attempted using the
+notion of \emph{typestate} in object-oriented languages
+\cite{deline:typestates}. Our approach differs in that we make no changes to
+the type system in order to implement this functionality; we check resource
+usage protocol conformance directly within \idris{}.  Additionally, our
+handler-based approach gives greater control over side effects.
 
